---
title: "Data Frame Operations"
subtitle: "Lecture 10"
format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
lang: en
geometry:
  - margin = 1in
  - paperwidth = 11in
  - paperheight = 6.875in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{sectsty}
  - \allsectionsfont{\centering}
  - \pagenumbering{gobble}
execute: 
  error: true
---

# Guitars Again
```{r}
  mfr <- rep(c("Fender", "Gibson", "Martin"), times = c(2,4,1))
  model <- c("Stratocaster", "Telecaster", "Firebird V", "Les Paul Standard", "ES175",
             "J-100 Xtra", "M-21")
  serial <- c("EJ15230", "60671", "616619", "00235628", "03531746", "90783019", "1452869")
  type <- rep(c("electric", "acoustic"), times = c(5, 2))
  body <- rep(c("solid", "hollow", "flat top"), times = c(4, 1, 2))
  w_case_kg <- c(8.96, 8.60, 9.88, 8.40, 8.10, 6.78, 5.62)
  case_kg <- c(5.70, 4.98, 5.90, 3.46, 4.92, 4.76, 3.72)
  
  guitars <- data.frame(mfr, model, serial, type, body, w_case_kg, case_kg)
```

***

* Deleting column of serial numbers to declutter slides.
```{r}
#| echo: true
guitars$serial <- NULL

guitars
```
```{r}
#| echo: true

M <- matrix(runif(100), 10, 10)
M
```


# OPERATING ON COLUMNS
# Transform Variables (Columns)
* Add columns for weights of the guitars in kg and lb
```{r}
guitars$kg <- guitars$w_case_kg - guitars$case_kg
guitars$lb <- 2.2 * guitars$kg
guitars
```
* This is often how it’s done, but that’s a lot of `guitars$`'s to type.

***
* Base R provides `transform()` to make this a little easier.
* `transform()` lets us work within the data frame, treating columns like vectors. 
* `transform()` returns a data frame including any new or modified columns.
```{r}
guitars <- transform(guitars, kg = w_case_kg - case_kg)
guitars <- transform(guitars, lb = 2.2 * kg)
guitars
```
* Note that naming inside `transform()` uses `=` and not `<-`.

***

* Note: when using `transform()`
  - You **can** do several transformations with a single call to transform, ...
  - But **not** if one transformation depends another, i.e., transformations can only use pre-existing columns.
  
So this would have worked:
```{r}
guitars <- transform(guitars,
                     kg = w_case_kg - case_kg,
                     lb = 2.2 * (w_case_kg - case_kg))
```

But this would **not** have worked, because `lb` is defined using `kg`.
```{r}
guitars <- transform(guitars,
                     kg = w_case_kg - case_kg,
                     lb = 2.2 * kg)
```

# Rename Columns
Rename `lb` column to `weight`
```{r}
names(guitars)[8] <- "weight"    # names(guitars)[names(guitars) == "lb"] <- "weight" 
guitars
```

* When might this process be mistake-prone or otherwise problematic?
  - Hint: What if we had 40 columns and wanted to rename several of them?

# Select Columns
* Select all the columns except for `w_case_kg`, `case_kg`, and `kg`
```{r}
guitars <- guitars[, -c(5:7)]
## Or guitars <- guitars[, c(1:4, 8)]
## Or guitars <- guitars[, c("mfr", "model", "type", "body", "weight")]
## Or guitars <- guitars[, !(colnames(guitars) %in c("w_case_kg", "case_kg", "kg"))]
guitars
```
* When might this process be difficult and/or mistake-prone?

# Relocate Columns
Suppose we want to make `type` the first column.
```{r}
guitars
```

***

```{r}
guitars <- guitars[, c(3, 1:2, 4:5)]
## Or guitars <- guitars[, c("type", "mfr", "model", "body", "weight")]

guitars
```
Now suppose we want to move `type` back to the 3rd column.

***


```{r}
guitars <- guitars[, c(2:3, 1, 4:5)]
## Or guitars <- guitars[, c("mfr", "model", "type", "body", "weight")]

guitars
```
* When might this process be difficult and mistake-prone?
  - What if we had 40 columns and wanted to relocate several of them?
  
# OPERATING ON ROWS
# Subsetting (Filtering) the Observations  
* View only guitars under 8 lb
```{r}
guitars[guitars$weight < 8, ]
```    
    
***

* View only electric guitars under 8 lb
```{r}
guitars[(guitars$type == "electric") & (guitars$weight < 8), ]
```  
* Seems like overkill to type `guitars` 3 times.

***

* Base R provides the `subset()` function to make this easier.
```{r}
subset(guitars, weight < 8)
subset(guitars, (type == "electric") & (weight < 8))
```

***

* Note that `subset()` can also select columns with its `select` argument.
  - Do **not** have to quote column names
  - Can use "-" with column numbers and **with names!**
  

```{r}
subset(guitars, select = c(model, weight))
```

***

```{r}
subset(guitars, (type == "electric") & (weight < 8), select = -c(type, weight))
```


# Ordering Observations
Arrange the guitars by weight (lightest to heaviest):

```{r}
guitars[order(guitars$weight), ]    # How does this work?
```

* `order(guitars$weight)` returns an index vector that could be used to put the column of weights in increasing order.
* We use that index vector to order the rows of the data frame.

***

Look over the following:
```{r}
guitars$weight
order(guitars$weight)    # Returns an index vector ...
guitars$weight[order(guitars$weight)]    # ... that puts the weights in increasing order.
sort(guitars$weight)    # Note: Do this if you just want the sorted weights.
```

***
**Decreasing Order**  
Arrange the guitars from heaviest to lightest:
```{r}
guitars[order(guitars$weight, decreasing = TRUE), ]
```

***

**Using Multiple Sort Keys**  
Arrange the guitars alphabetically by mfr and by weight within mfr:
```{r}
guitars[order(guitars$mfr, guitars$weight), ]
```
* We could have used `with()` here, like this:
```{r}
guitars[with(guitars, order(mfr, weight)), ]
```

***

**A Complication**  
What if we wish to arrange the guitars alphabetically by type and from heaviest to lightest within type?  

* Because the `type` column is character and the `weight` column is numeric, we have to add `method = "radix"` in the call to `order()`.
```{r}
guitars[order(guitars$type, guitars$weight, decreasing = c(FALSE, TRUE),
              method = "radix"), ]
```

# Getting the First or Last Few Rows
**First Few Rows**
```{r}
head(mtcars)    # Same as  head(mtcars, n = 6)
```
* `head(mtcars, n = 4)` is the same as `mtcars[1:4, ]`

***

**Last Few Rows**
```{r}
tail(mtcars)   # Same as tail(mtcars, n = 6)
```

* Much easier than `mtcars[(nrow(mtcars) - 5):nrow(mtcars), ]`

# GROUPED CALCULATIONS
# Using `tapply()`
**Calculate the Mean Weight by Type of Guitar**

* Grouping variable (`type` in this example) is usually a factor (qualitative).
* `tapply()` returns a named numeric vector in this example.
```{r}
tapply(guitars$weight, guitars$type, mean)
with(guitars, tapply(weight, type, mean))    # Same thing using with().
```

***

**Calculate the Mean Weight by Mfr and Type of Guitar**  

* Again, grouping variables are usually factors (qualitative).
```{r}
with(guitars, tapply(weight, list(mfr, type), mean))
```

* `tapply()` returns a numeric matrix in this example.
```{r}
is.matrix(with(guitars, tapply(weight, list(mfr, type), mean)))
```
* A matrix is fine for displaying these results in table form, but...

***

**What if we wanted a data frame containing the group means?**
```{r}
avg_wts <- with(guitars, tapply(weight, list(mfr, type), mean))
avg_wts <- data.frame(avg_wts)
avg_wts     # Printed looks the same as the matrix.
```
* Is this what we want as a data frame?
* No!
  - Column headers are values (of the variable `type`), not variable names.
  - Mfr values used as rownames. Should be a column.
  - Should be a data frame with 4 (or 6?) rows and 3 cols (mfr, type, avg_wt).
  
# Using `aggregate()`
* `aggregate()` will return a data frame.
* `aggregate()` has an S3 method for data frames, but it can be difficult to use.
```{r}
aggregate(guitars["weight"], by = guitars["type"], FUN = "mean")
```
***
* I prefer the formula method for `aggregate()`
```{r}
aggregate(weight ~ type, data = guitars, FUN = "mean")
aggregate(weight ~  mfr + type, data = guitars, FUN = "mean")
```
***
* What if we wanted the counts of the types as well? Could try something like:
```{r}
aggregate(weight ~ mfr + type, data = guitars,
          FUN = function(x) c(n = length(x), avg = mean(x)))
```
  - Note that the counts have been coerced to dbl and the column names need repair, so this isn’t a great result.

# Refresh **Guitars** Again
```{r}
mfr <- rep(c("Fender", "Gibson", "Martin"), times = c(2,4,1))
model <- c("Stratocaster", "Telecaster", "Firebird V", "Les Paul Standard", "ES175",
           "J-100 Xtra", "M-21")
serial <- c("EJ15230", "60671", "616619", "00235628", "03531746", "90783019", "1452869")
type <- rep(c("electric", "acoustic"), times = c(5, 2))
body <- rep(c("solid", "hollow", "flat top"), times = c(4, 1, 2))
w_case_kg <- c(8.96, 8.60, 9.88, 8.40, 8.10, 6.78, 5.62)
case_kg <- c(5.70, 4.98, 5.90, 3.46, 4.92, 4.76, 3.72)

guitars <- data.frame(mfr, model, serial, type, body, w_case_kg, case_kg)
guitars <- transform(guitars, weight = w_case_kg - case_kg)
guitars <- subset(guitars, select = -c(serial, w_case_kg, case_kg))
```
***

# Using `tapply()` and `expand.grid()`
```{r}
avg_wts <- with(guitars, tapply(weight, list(mfr, type), mean))  # Matrix 
guitar_smry <-
    with(guitars, expand.grid(mfr = unique(mfr), type = unique(type)))  # 2-col data frame
guitar_smry <- transform(guitar_smry, mfr = as.character(mfr), type = as.character(type))

# To reverse the reordering:
# temp_order <- with(guitar_smry, order(type, mfr))

guitar_smry <- guitar_smry[temp_order, ]
guitar_smry$avg_wt <- as.vector(avg_wts)  # Add column of avg_wts 
guitar_smry

# To reverse the reordering:
# temp_order <- cbind(1:length(temp_order), temp_order)
# temp_order <- temp_order[order(temp_order[, 2]), ]
# guitar_smry <- guitar_smry[temp_order[, 1],]

```
***

Let’s add a column giving the number of guitars in each category (tricky?).
```{r}
#| error: true

n <- with(guitars, tapply(weight, list(mfr, type), length))   # Matrix of counts
guitar_smry$n <- as.vector(n)    # Add column of counts 
guitar_smry
```
***

* The NA values occur because there were no Fender acoustic guitars and no Martin electric guitars in the sample. I would probably remove these rows.
* I would probably move the `n` column before the `avg_wt` column too.

```{r}
guitar_smry <- guitar_smry[!is.na(guitar_smry$avg_wt), c("mfr", "type", "n", "avg_wt")] 

## Equivalent
## guitar_smry <- subset(guitar_smry, !is.na(avg_wt), select = c(mfr, type, n, avg_wt))

guitar_smry
```
    
# Using `tapply()` and `do.call("rbind", ...)` 
Here’s another way to use `tapply()` with an anonymous function to get a list of 1-row data frames, which is then passed `to do.call("rbind", ...)` to bind all the rows together into a single data frame.


```{r}
#| error: true

do.call("rbind",
        tapply(guitars$weight, guitars$type,
               function(x) data.frame(n = length(x), avg_wt = mean(x))))
```

* Unfortunately the types are used as row labels, not a separate column.


***

How does this approach work if we are grouping by multiple variables? Not well — we lose the grouping information altogether.

```{r}
#| error: true
with(guitars,
     do.call("rbind",
             tapply(weight, list(mfr, type),
                    function(x) data.frame(n = length(x), avg_wt = mean(x)))))
```

More:

```{r}
#| error: true
temp_guitars <- split(guitars, list(mfr, type))
do.call(rbind, lapply(temp_guitars, function(x){
  data.frame(avg_wgt = mean(x$weight), n_mfr_type = nrow(x))
}))
```