---
title: "Quarto Demonstration"
description: This file demonstrates many of the most relevant options for a Quarto document for use with assignments. Please use this as a jumping-off point for your own work.

format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
lang: en

execute:
  error: false
  warning: false
  fig-width: 6
  fig-height: 6
---

# Section

## Sub-section

### Sub-sub-section

Hello, this is new text

```{r}
#| label: block1
#| eval: true

  # Block 1
  # "eval: true" means block code will be run
  
  print("This sentence is output from Block 1.")
  x <- 1:5
```

```{r}
print(x)
cat(x)
```

`{r} print(x)`

`{r} x`

`{r} cat(x)`

`{r} print(x)`

`{r} x^2`

```{r}
x^2
```

```{r}
(1:1000)^2

sum(1:100)
mean(1:100)
sd(1:100)
```

```{r}
#| fig-width: 6
#| fig-height: 4 
#| fig-alt: "Plot of vector containing values 1:100"
plot(1:100)
```



```{r}
#| label: block2
#| eval: false

  # Block 2
  # "eval: false" means block code will not be run
  
  print("This sentence is output from Block 2.")
  
  quit()
```

No ending spaces/tabs:

New Paragraph

Ending 1 tab/2 spaces:\
New Line

| No Indent
|  Indent 1 Space
|   Indent 2 Spaces

*Italics* **Bold** ***Bold+Italics***

*Italics* **Bold** ***Bold+Italics***

Script ^Superscript^ ~Subscript~

Latex-style math formulas: $\sum_{i=1}^{4} i^3 = 1+8+27+64 = 100$

$$\sin(\frac{\pi}{2})=1$$

```{r}
#| label: block3
#| echo: true
  
  # Block 3
  # "echo: true" means code will be shown
  
  print("This sentence is output from Block 3.")
```

```{r}
#| label: block4
#| echo: false
 
  # Block 4
  # "echo: false" means block code will not be shown
  
  print("This sentence is output from Block 4.")
```

In the following, note that *tabs are important*

1.  Numbered
2.  List
    a.  Lettered (4 spaces/2 tabs)
    b.  Sublist
        a)  Different (8 spaces/4 tabs)
        b)  Format

-   Unnumbered
-   List

```{r}
#| label: block5
#| include: true

  # Block 5
  # "include: true" means block output will be shown
  
  print("This sentence is output from Block 5.")
```

```{r}
#| label: block6
#| include: false

  # Block 6
  # "include: false" means block output will not be shown

  print("This sentence is output from Block 6.")
```

```{r}
  # Blocks do not need names or options
  # Sets "x" to be 23.42892 and prints
  
  x <- 23.42892
  print(x)
```

This is inline code. The variable *x* is `{r} round(x, 2)` and *x*^2^ is `{r} round(x^2, 3)`.

```{r}
  data1 <- rnorm(100)
```

1)  The mean of **data1** is `{r} mean(data1)`. The variance is `{r} var(data1)`. The range is `{r} max(data1)-min(data1)`.

```{r}
  y <- runif(20)
  print(y)
```

```{r}
#| results: "asis"
  
  cat("The formatted output is ", y)
```