---
title: "Arrays"
subtitle: "Lecture 7"
format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
lang: en
geometry:
  - margin = 1in
  - paperwidth = 11in
  - paperheight = 6.875in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{sectsty}
  - \allsectionsfont{\centering}
  - \pagenumbering{gobble}
execute: 
  error: true
---

# Arrays and `apply()`

* Arrays are like matrices but the dimension may be more than 2 
  - i.e., matrices are 2-dimensional arrays.
* The `apply()` function _applies_ a function to one or more dimensions of an array.
* Matrices and arrays are often used to store multi-way _contingency tables_, although this is not their only use.

# Constructing an Array
```{r}
  array(1:24, dim = c(3, 4, 2))
```

## `?iris`  
`Description:`   

\setlength{\leftskip}{0.6cm}

`This famous (Fisher's or Anderson's) iris data set gives the`     
`measurements in centimeters of the variables sepal length and width and`   
`petal length and width, respectively, for 50 flowers from each of 3`    
`species of iris.  The species are _Iris setosa_, _versicolor_, and`  
`_virginica_.`

\setlength{\leftskip}{0cm}

`Format:`  

\setlength{\leftskip}{0.6cm}

`‘iris3’ gives the same data arranged as a 3-dimensional array of size`  
`50 by 4 by 3, as represented by S-PLUS.  The first dimension gives the`  
`case number within the species subsample, the second the measurements`  
`with names ‘Sepal L.’, ‘Sepal W.’, ‘Petal L.’, and ‘Petal W.’, and the`  
`third the species.`   

\setlength{\leftskip}{0cm}

\newpage

```{r}
  dim(iris3)
  dimnames(iris3)
```
    
\newpage

```{r}
  iris3[1,,"Setosa"]     # Measurements from the first iris setosa
  iris3[2,,"Setosa"]     # Measurements from the second iris setosa
  iris3[1,,"Versicolor"] # Measurements from the first iris versicolor
```

\newpage

```{r}
  iris3[,"Petal L.","Virginica"]     # Petal lengths of all 50 iris virginica observed
  iris3[1:2,,"Versicolor"]           # All measurements from 1st two iris versicolor observed
```

# Aside
Compare to a data frame containing the same data (more on this later):
```{r}
  iris[1:3,]
  dim(iris)
  c( prod(dim(iris3)), prod(dim(iris)) )
```

\newpage

# `apply()` for Arrays
```{r}
  ## Means for each comb. of measurement (dim 2) and species (dim 3)
  apply(iris3, c(2,3), mean)
```
```{r}
  ## Means for each measurement type (dim 2) (ignoring species)
  apply(iris3, 2, mean)
```

\newpage

```{r}
  # Std devs for each combination of measurement and species
  apply(iris3, c(2,3), sd)  
```

# `sweep()` 
`Description:`  

\setlength{\leftskip}{0.6cm} 

`Return an array obtained from an input array by sweeping out a summary`      
`statistic.`   

\setlength{\leftskip}{0cm}


`Usage :` 

\setlength{\leftskip}{0.6cm}

`sweep(x, MARGIN, STATS, FUN = "-", check.margin = TRUE, ...)`    

\setlength{\leftskip}{0cm}

\newpage

* In this example we center the species means by subtracting the overall mean for each measurement type (refer to earlier slide).  
```{r, echo = TRUE}
## Means for each comb. of measurement (dim 2) and species (dim 3)
species_means <- apply(iris3, c(2,3), mean)

## Means for each measurement type (dim 2) (ignoring species)
overall_means <- apply(iris3, 2, mean)

sweep(species_means, 1, overall_means, "-")
```

\newpage

# Death and Survival on the Titanic
```{r}
  dim(Titanic)
  dimnames(Titanic)
```

\newpage

```{r}
  Titanic["3rd", "Female", "Adult", "Yes"]
  Titanic["3rd", "Female", "Adult", ]
  Titanic[, "Female", "Adult", ]
```

\newpage

```{r}
  Titanic[, , "Child", ]
  Titanic[, , "Adult", ]
```

\newpage

# Flattening a Multiway Contingency Table
```{r}
  is.table(Titanic)
  ftable(Titanic, row.vars = "Class", col.vars = c("Age", "Sex", "Survived"))
```

\newpage

# Using `apply()`
```{r}
  # Total number of passengers
  sum(Titanic)
  ## Marginal table for survival: overall died and survived. Could use `marginSums()`.
  apply(Titanic, "Survived", sum)
  ## Marginal percentages died and survived (ignoring class, sex, age)
  100 * apply(Titanic, "Survived", sum) / sum(Titanic)
```

\newpage

```{r}
  apply(Titanic, c("Survived", "Class"), sum)    # Number died and survived by class
```
The following calculation shows that of the 2201 passengers, 5.5% were first class passengers who died, 9.2% were first class passengers who survived, etc.
```{r}
  100 * apply(Titanic, c(4, 1), sum) / sum(Titanic)
```

\newpage
What would we really like to know?

* We suspect that passenger class may be predictive of survival.
* Thus, we would like to know the proportion of passengers who survived _within each class_.
* These proportions can be computed using `apply()` and `sweep()` but it isn’t much fun to do it this way.
  - The process is illustrated in the next few slides, but we will not spend much class time on them because fortunately...
* R provides some convenience functions to make this sort of calculation with tables much easier.

\newpage

```{r}
  ## Marginal table for survival and class (ignoring sex and age).
  surv_class_tot <- apply(Titanic, c("Survived", "Class"), sum)
  surv_class_tot
```

```{r}
  ## Marginal table for class.
  class_tot <- apply(Titanic, "Class", sum)
  class_tot
```

Now we want to divide (`FUN = "/"`) the columns (`MARGIN = 2`) of the first table by values in the second (`STAT = class_tot`). We can use `sweep()` to do this.

\newpage

```{r}
  sweep(surv_class_tot, 2, class_tot, "/")
```
Can replace `2` by the name of the column dimension, which is “Class” in this case:
```{r}
  round( 100 * sweep(surv_class_tot, "Class", class_tot, "/"), 1 )
```
Thus 62.5% of 1st class passengers survived, while only 25.2% of 3rd class passengers survived.

\newpage

```{r}
  surv_class_age_tot <- apply(Titanic, c("Survived", "Class", "Age"), sum)
  
  class_age_tot <- apply(Titanic, c("Class", "Age"), sum)
  
  round(100 * sweep(surv_class_age_tot, c("Class", "Age"), class_age_tot, "/"), 1)
```

\newpage

# Three Convenience Functions for Tables

* The `apply()` and `sweep()` functions are for general matrices and arrays.
* The Titanic dataset is a (contingency) “table”.
```{r}
  class(Titanic)
```
* There are several convenience functions which apply a combination of `apply()` and `sweep()` to multi-way contingency tables to carry out some common tasks with tables:  
  - `marginSums()`
  - `proportions()`
  - `addmargins()`
  
  \newpage
  
```{r}
  ## We did this before with apply().  Here we use `marginSums()`.
  surv_class_tot <- marginSums(Titanic, c("Survived", "Class"))
  
  surv_class_tot 
```

```{r}
  addmargins(surv_class_tot)    # Glues on row and column totals
```

\newpage

* Only the first of the following two sets of proportions is really useful or interesting, so make sure you know how to calculate the right one.
```{r}
  ## Proportion dying and surviving within each class:
  proportions(surv_class_tot, "Class")
```

```{r}
  ## Class proportions among those who died and among those who survived:
  proportions(surv_class_tot, "Survived")
```

\newpage

`proportions(Titanic, c("Class", "Sex", "Age"))`

* Gives proportions dying and surviving for each comb. of class, sex, and age.
* We convert to percentages and round to one place past the decimal point.
* We then use `ftable()` to print a flattened version of the resulting table of percentages.
```{r}
  ftable(round(100 * proportions(Titanic, c("Class", "Sex", "Age")), 1),
    row.vars = c("Sex", "Survived"), col.vars = c("Age", "Class"))
```

\newpage

# Chi-Square Test of Independence
We can use the usual chi-square test to test the null hypothesis of (marginal) independence of passenger class and survival (ignoring sex and age).
```{r}
  surv_class_tot
  chisq.test(surv_class_tot)
```
