---
title: "Timing and Writing Functions"
subtitle: "Lecture 36"
format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
lang: en
geometry:
  - margin = 1in
  - paperwidth = 11in
  - paperheight = 6.875in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{sectsty}
  - \allsectionsfont{\centering}
  - \pagenumbering{gobble}
execute: 
  error: true
--- 

# Timing Calculations and Writing Simple Functions
# Timing

* It can be useful to time calculations in order to [\textcolor{blue}{compare different computational approaches}](https://stats.oarc.ucla.edu/r/faq/how-can-i-time-my-code/).
* There are [\textcolor{blue}{several ways}](https://www.r-bloggers.com/2017/05/5-ways-to-measure-running-time-of-r-code/) to do this, but we will just look at `system.time()`.
* For more complex problems, sophisticated [\textcolor{blue}{packages for profiling R code}](https://adv-r.hadley.nz/perf-measure.html) may be useful.
* `system.time(expr)` evaluates `expr` and returns the execution time in seconds. Actually, it returns three times:
  - `user`: the “user CPU time” taken by the current process (R session)
  - `system`: the “system CPU time” spent by the operating system kernel on
behalf of the current process
  - `elapsed`: the wall clock time taken

```{r}
nsim <- 10000 
ntosses <- 100
```

* Pre-allocating the `maxrl` vector.

```{r}
system.time({
  maxrl <- vector(length = nsim) 
  for (i in 1:nsim) {
    x <- sample(c("H","T"), ntosses, replace = TRUE) 
    xr <- rle(x)
    maxrl[i] <- max(xr$lengths)
  }})
```

* Building the `maxrl` vector as we go.

```{r}
system.time({ 
  maxrl <- NULL
  for (i in 1:nsim) {
    x <- sample(c("H","T"), ntosses, replace = TRUE) 
    xr <- rle(x)
    maxrl <- c(maxrl, max(xr$lengths))
  }})
```

* Generating all $10,000 \times 100 = 1,000,000$ tosses at once.
* Arranging the tosses into a matrix.
* Using `apply()` to get the 10,000 maximum run lengths.

```{r}
system.time({
  sims <- matrix(sample(c("H","T"), ntosses * nsim, replace = TRUE), 
                 ncol = nsim)
  maxrl <- apply(sims, 2, function(x) max(rle(x)$lengths))
  })
```

# Defining an R Function

Recall

* A simple R function can be defined on one line.

```{r}
incr <- function(x) x + 1 
incr(1:5)
```

* Function calls can be nested (or piped).

```{r}
incr(incr(1:5))
```

* Arguments may have default values which may be overridden when the function is called.

```{r}
incr <- function(x, a = 1)  x + a 
incr(1:5)
incr(1:5, a = 3)
```

* For more complicated functions, the body is enclosed in curly braces `{...}`.
* Usually the last line or the last function call in the body determines the value returned by the function.

***

**Example: Head or Tails or Either**

Let’s write an R function to calculate the longest run in a sequence and optionally to calculate the longest run of a given value (such as “H” or “T”).

* We use a common R trick: an optional argument with default value `NULL`.
  - Inside the function we check the value of this argument with `is.null()`
to decide whether to perform a default operation or something else.

```{r}
fmaxrl <- function(x, outcome = NULL) { 
  xr <- rle(x)
  if (is.null(outcome)) { 
    res <- max(xr$lengths)
    } else {
      res <- max(xr$lengths[xr$values == outcome])
  }
  res 
}
```

* Because there was no further calculation to do after `if() {...} else {...}`, the previous function definition could be abbreviated slightly ...

```{r}
fmaxrl <- function(x, outcome = NULL) { 
  xr <- rle(x)
  if (is.null(outcome)) { 
    max(xr$lengths)
  } else {
    max(xr$lengths[xr$values == outcome])
  } 
}
```

* ... but this is harder to understand, so I wouldn’t do it this way.

```{r}
x <- sample(c("H","T"), 100, replace = TRUE) 
fmaxrl(x)
fmaxrl(x, outcome = "H")
fmaxrl(x, outcome = "T")
fmaxrl(x, outcome = 5)       # May wish to handle this more gracefully.
```

***

**Nesting if-else Statements**

```{r}
fmaxrl <- function(x, outcome = NULL) { 
  xr <- rle(x)
  if (is.null(outcome)) { 
    res <- max(xr$lengths)
  } else {
    eq_to_outcome <- (xr$values == outcome) 
    if (any(eq_to_outcome)) {
      res <- max(xr$lengths[eq_to_outcome]) 
    } else {
      res <- 0 
    }
  }
  res 
}
```

```{r}
fmaxrl(x)
fmaxrl(x, outcome = "H")
fmaxrl(x, outcome = "T")
fmaxrl(x, outcome = 5)
mean(apply(sims, 2, fmaxrl) >= 6)
mean(apply(sims, 2, fmaxrl, outcome = "H") >= 6)
mean(apply(sims, 2, fmaxrl, outcome = "T") >= 6)   # Q: Why not equal to answer for "H"?
```

Q: What is the estimated probability that in 100 tosses we observe at least one run of 6 or more heads **and** at least one run of 6 or more tails?


***



Here is better version of the `fmaxrl` function.

```{r}
fmaxrl <- function(x, outcome = NULL) { 
  xr <- rle(x)
  if (!is.null(outcome)) {
    is_outcome <- (xr$values == outcome)
    xr$lengths <- xr$lengths[is_outcome] 
  }
  if (length(xr$lengths) == 0L) 0 else max(xr$lengths)  # Return 0 or max(xr$lengths)
}
```


* Could add `xr$values <- xr$values[is_outcome]` in the first `if` statement, but we don’t use `xr$values` for anything after this.
* `is_outcome` is only used once, but was defined anyway to make the function easier to understand. The two lines in the first `if` statement could be replaced by the single line  
`xr$lengths <- xr$lengths[xr$values == outcome]`