---
title: "`dplyr`: Two-Tibble Operations"
subtitle: "STA3100 Lecture 18"
format:
  html:
    embed-resources: true
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    geometry:
      - margin = 1in
      - paperwidth = 11in
      - paperheight = 6.875in
    header-includes:
      - \usepackage{booktabs}
      - \usepackage{sectsty}
      - \allsectionsfont{\centering}
      - \pagenumbering{gobble}
lang: en

execute:
  error: true
---

# JOINING/MERGING DATA FRAMES
# `dplyr` Two-Tibble “Verbs” and Base-R Equivalents
Functions used to merge two data frames by matching on common variables.  
Besides the assigned readings, see:  

* the Two-tibble verbs section of [`dplyr` <-> Base-R](https://cran.r-project.org/web/packages/dplyr/vignettes/base.html) for a comparison of `dplyr` and base-R functions. 
* the [Two-Tibble Verbs](https://cran.r-project.org/web/packages/dplyr/vignettes/two-table.html) vignette for more on `dplyr` two-tibble verbs.
* [`dplyr` 1.1.0: Joins](https://www.tidyverse.org/blog/2023/01/dplyr-1-1-0-joins/)

# Mutating Joins
| Rows included                  | `dplyr` function(s)            |Base-R function(s)                |
|:-------------------------------|:-----------------------------|:---------------------------------|  
|matched in both `df1` & `df2`   |`inner_join(df1, df2)`        |`merge(df1, df2)`                 |
|all from both `df1` & `df2`     |`full_join(df1, df2)`         |`merge(df1, df2, all = TRUE)`     |
|all in `df1`                    |`left_join(df1, df2)`         |`merge(df1, df2, all.x = TRUE)`   |
|all in `df2`                    |`right_join(df1, df2)`        |`merge(df1, df2, all.y = TRUE)`   |
    
# Filtering Joins

* Mostly for diagnosing join mismatches.
* Returns matched (`semi_join()`) or unmatched (`anti_join()`) rows of `df1`.
* Does not actually perform the merge: just acts as a row-filter on `df1`.

| Rows included                 | `dplyr` function(s)   |Base-R function(s)                      |
|:------------------------------|:--------------------|:---------------------------------------|  
|in `df1` & matched in `df2`    |`semi_join(df1, df2)`|`df1[df1$x %in% df2$x, , drop = FALSE]` |
|in `df1` & not matched in `df2`|`anti_join(df1, df2)`|`df1[!df1$x %in% df2$x, , drop = FALSE]`|


# Guitars Again
```{r}

mfr <- rep(c("Fender", "Gibson", "Martin"), times = c(2,4,1))
model <- c("Stratocaster", "Telecaster", "Firebird V", "Les Paul Standard", "ES175",
           "J-100 Xtra", "M-21")
serial <- c("EJ15230", "60671", "616619", "00235628", "03531746", "90783019", "1452869")
type <- rep(c("electric", "acoustic"), times = c(5, 2))
body <- rep(c("solid", "hollow", "flat top"), times = c(4, 1, 2))
w_case_kg <- c(8.96, 8.60, 9.88, 8.40, 8.10, 6.78, 5.62)
case_kg <- c(5.70, 4.98, 5.90, 3.46, 4.92, 4.76, 3.72)

guitars_df <- data.frame(mfr, model, serial, type, w_case_kg, case_kg)

library("tidyverse")
guitars_tbl <- as_tibble(guitars_df)
```

* Removing `w_case_kg` and `case_kg` to save room in slides.
```{r}
guitars_df <- subset(guitars_df, select = -c(w_case_kg, case_kg))
guitars_tbl <- select(guitars_tbl, -w_case_kg, -case_kg)
```

# Suppose Color is Recorded in a Separate Data Frame
```{r}
colors_tbl <-
    tribble(~model, ~color, ~serial,
            "Firebird V", "sunburst", "616619",
            "Les Paul", "wine", "00235628",
            "ES175", "sunburst", "03531746",
            "Stratocaster", "black", "EJ15230",
            "Telecaster", "blonde", "60671",
            "J-100", "natural", "90783019",
            "D-28", "natural", "2457919")
colors_df <- as.data.frame(colors_tbl)
```

***

```{r}
guitars_tbl
colors_tbl
```

* Notice that:
  - a guitar is missing (M-21) from colors data.
  - a new guitar has been added (D-28) to colors data.
  - "Les Paul Standard" appears as "Les Paul" and "J-100 Xtra" as "J-100" in colors data.

# Inner Join: All Key(s) Must Match

* Keep rows for which the default keys `model` and `serial` match.
  - Default keys for matching are any columns with the same names in both data frames.

**Base-R**
```{r}
merge(guitars_df, colors_df)
```

***

**`dplyr::inner_join()`**

```{r}
inner_join(guitars_tbl, colors_tbl)
```

* Presumably we are surprised that our guitar collection has shrunk.
* Note that:
  - `merge()` moved the join keys to the front
  - `inner_join()` kept column order from left to right, dropping any duplicated columns.

***

**Specify Matching on `serial` only.**  
**Base-R**
```{r}
merge(guitars_df, colors_df, by = "serial")
```

* Get 2 model cols because the entries do not match.
* We are still missing two guitars (M-21 and D-28).
* Again, `merge()` has made the matching key the first column.

***

**`dplyr::inner_join()`**
```{r}
inner_join(guitars_tbl, colors_tbl, by = "serial")
```

* Get 2 model cols because the entries do not match.
* We are still missing two guitars (M-21 and D-28).
* Columns still in original left-to-right order.

***

* Since `dplyr` 1.1.0, there is an optional `unmatched` argument.
  - The default is `unmatched = "drop"`.
  - Can change this to `unmatched = "error"`.
  - Useful when each item should have at least one match. (With larger data frames we might not notice otherwise.)

```{r}
inner_join(guitars_tbl, colors_tbl, by = "serial", unmatched = "error")
```

# Diagnose Mismatches

**Base-R**
```{r}
guitars_df[!(guitars_df$model %in% colors_df$model),]    # In guitars, not in colors
colors_df[!(colors_df$model %in% guitars_df$model),]     # In colors, not in guitars
```

***

**`dplyr::anti_join()`**
```{r}
anti_join(guitars_tbl, colors_tbl)         # In guitars, not in colors
anti_join(colors_tbl, guitars_tbl)         # In colors, not in guitars
```

***

**Let’s repair the mismatched model names.**

* This is the most basic of base-R ways to do it.

```{r}
colors_df$model[colors_df$model == "Les Paul"] <- "Les Paul Standard"
colors_df$model[colors_df$model == "J-100"] <- "J-100 Xtra"
colors_df
```

***

* We could have fixed the mismatches with the base-R `replace()` function.

```{r}
colors_df <- colors_df |>
  transform(model = replace(model, model == "Les Paul", "Les Paul Standard")) |>
  transform(model = replace(model, model == "J-100", "J-100 Xtra"))
```

* Or with the base-R `ifelse()` function.
```{r}
colors_df <- colors_df |>
  transform(model = ifelse(model == "Les Paul", "Les Paul Standard", model)) |>
  transform(model = ifelse(model == "J-100", "J-100 Xtra", model))
```
  - `ifelse()` is useful for other kinds of transformations as well, e.g.,
```{r}
x <- 1:10
y <- ifelse((x %% 2) == 0, -x, 2*x)
y
```

***

* The `dplyr::case_match()` can fix both mismatches at once.
  - See also `dplyr::case_when()` (can do other kinds of tranformations).
```{r}
colors_tbl <- colors_tbl |>
  mutate(model = case_match(model,
                              "Les Paul" ~ "Les Paul Standard",
                              "J-100" ~ "J-100 Xtra",
                              .default = model))
colors_tbl
```

***

**Which mismatches remain?**

```{r}
anti_join(guitars_tbl, colors_tbl)        # In guitars, not in colors
anti_join(colors_tbl, guitars_tbl)        # In colors, not in guitars
```


# Full Outer Join: All Rows from Both Data Frames

**Base-R**

* Default matching on all common column names.

```{r}
merge(guitars_df, colors_df, all = TRUE)
```

***

**`dplyr::full_join()`**

* Default matching on all common column names.

```{r}
full_join(guitars_tbl, colors_tbl)
```

***

**Contrast With ...**  
**Base-R**

* Matching on specified column(s).

```{r}
merge(guitars_df, colors_df, by = "serial", all = TRUE)
```

***

**`dplyr::full_join()`**  

* Matching on specified column(s).

```{r}
full_join(guitars_tbl, colors_tbl, by = "serial")
```

# Left Outer Join: All Rows from First Data Frame

**Base-R**  

* Default matching on all common column names.

```{r}
merge(guitars_df, colors_df, all.x = TRUE)
```

***

**`dplyr::left_join()`**   

* Matching on specified column(s).

```{r}
left_join(guitars_tbl, colors_tbl, by = "serial")
```

# Right Outer Join: All Rows from Second Data Frame

**Base-R**  

* Matching on specified column(s).

```{r}
merge(guitars_df, colors_df, all.y = TRUE, by = "serial")
```

***

**`dplyr::right_join()`**  

* Default matching on all common column names.
```{r}
right_join(guitars_tbl, colors_tbl)
```
