---
title: "Lists"
subtitle: "Lecture 8"
format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
lang: en
geometry:
  - margin = 1in
  - paperwidth = 11in
  - paperheight = 6.875in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{sectsty}
  - \allsectionsfont{\centering}
  - \pagenumbering{gobble}
execute: 
  error: true
---

\newpage  

# A List Can Contain Any Combination of R Objects
  
## Creating a Named List
```{r}
  big_sleep <- list(title = "The Big Sleep",
                    release_date = as.Date("1946-08-23"),
                    boxoffice = 4.9e6,
                    rottentomatoes = c(tomatometer = 97, audience = 91),
                    cast = c("Humphrey Bogart", "Lauren Bacall"))
  str(big_sleep)
```

\newpage

# A Great Film Noir
```{r}
  big_sleep
```

\newpage

# A List of Lists
```{r}
  big_lebowski <-
    list(title = "The Big Lebowski",
      release_date = as.Date("1998-01-18"),
      boxoffice = 46.7e6,
      rottentomatoes = c(tomatometer = 82, audience = 94),
      cast = c("Jeff Bridges", "John Goodman", "Julianne Moore",
               "Steve Buscemi", "David Huddleston", "John Turturro"))
  movies <- list("big_sleep" = big_sleep, "big_lebowski" = big_lebowski)
```

```{r}
  str(movies)
```

\newpage

# Combining Lists  

* `list()` makes a list with the arguments as its elements
* `c()` concatenates (as usual)

```{r}
  l1 <- list(a = "Hello", b = 1:3)    # A list of length 2
  l2 <- list("World", 4:6)            # A list of length 2
  v <- 7:9                            # A vector of length 3
```

\newpage

### Combining Lists With `list()`
```{r}
  l.l1l2 <- list(l1, l2)
  str(l.l1l2)
```
```{r}
  l.l1l2
```

\newpage

### Combining Lists With `c()`
```{r}
  c.l1l2 <- c(l1, l2)
  str(c.l1l2)
```
```{r}
  c.l1l2
```

\newpage

### `list()` with 2 lists and a vector
```{r}
  l.l1l2v <- list(l1, l2, v)
  str(l.l1l2v)
```
* List of length 3
```{r}
  l.l1l2v
```

### `c()` with 2 lists and a vector
```{r}
  c.l1l2v <- c(l1, l2, v)
  str(c.l1l2v)
```

* List of length 7

### `c()` with 3 lists
```{r}
  c.l1l2lv <- c(l1, l2, list(v))
  str(c.l1l2lv)
```

* List of length 5

\newpage

# One Way to Add a Director Element
```{r}
  big_sleep <- list(title = "The Big Sleep",
                    release_date = as.Date("1946-08-23"),
                    boxoffice = 4.9e6,
                    rottentomatoes = c(tomatometer = 97, audience = 91),
                    cast = c("Humphrey Bogart", "Lauren Bacall"))
  big_sleep <- c(big_sleep, list(director = "Howard Hawks"))
  str(big_sleep)
```

\newpage

# Indexing Lists: `[`, `[[`, and `$`  
Lists are sometimes indexed using the familiar `[` and other times using `[[` or `$`.  

* `[` applied to a list returns a sublist of the original list.
* `[[` is used to return the contents of a single element of the list.  
  - The index can be a single number or character string (if a named list).
* `$string` is an alias for `[["string"]]`.  

It may be helpful to think of a list as a train made up of train cars, with each car having its own contents. In this analogy:

* `[` is used to select cars from the train and return a new (shorter) train;
* `[[` applied to a list returns the contents of a car (as does `$`).  

Personally, I think of a list as a vector of pointers to other R objects.

```{r}
  names(big_sleep)
  big_sleep[4]       # Or big_sleep["rottentomatoes"].   Returns a list of length 1.
  big_sleep[[4]]     # Or big_sleep[["rottentomatoes"]]. Returns a vector of length 2.
```

```{r, error = TRUE}
  big_sleep$rottentomatoes
  # Shorthand for big_sleep[["rottentomatoes"]].
  big_sleep[c(1, 4)]
  # Or big_sleep[c("title", "rottentomatoes")]. List of length 2.
  big_sleep[[c(1, 4)]]
  # What is this trying to do? <!-- Return 4th elt of 1st elt of the list. -->
```

\newpage

# The Usual Way to Add a Director Element
```{r}
  big_lebowski$director <- c("Ethan Coen", "Joel Coen")
```
or
```{r}
  big_lebowski[["director"]] <- c("Ethan Coen", "Joel Coen")
```
```{r}
  big_lebowski$director
```

\newpage

# R Functions Often Return Lists as Output
```{r}
  mpg.lm <- lm(mpg ~ wt, data = mtcars) # Fit a linear (regression) model
  mpg.lm                                # Printed representation [ print(mpg.lm) ]
```
```{r}
#| fig-align: center
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "Regression of Miles/(US) Gallon on Weight (1000 lbs)"

  opar <- par(mar = c(5.1, 4.1, 1.1, 1.1))
  
  plot(mpg ~ wt, data = mtcars)
  abline(mpg.lm)
  
  par(opar)
```

```{r}
  str(mpg.lm, max.level = 1)
```

```{r}
  class(mpg.lm)
  typeof(mpg.lm)
  length(mpg.lm)
  names(mpg.lm)
```

```{r}
  mpg.lm["coefficients"]
  # Or mpg.lm[1].  List of length 1.
  mpg.lm[["coefficients"]]
  # Or mpg.lm[[1]] or mpg.lm$coefficients. Vector of length 2.
  mpg.lm$coefficients
```

```{r}
  mpg.lm[c("call", "coefficients")]
  # Or mpg.lm[c(10, 1)]
  mpg.lm[[c("call", "coefficients")]]
  # Nope. Cannot index two list elts at once with `[[`.
```

\newpage

# Brief Primer on S3 Classes and Generic Functions
* `print()` is a generic function
* An S3 generic function uses `UseMethod()` to dispatch to a specific method.
```{r}
  print
  # Printing print().
```
* `print()` methods are defined for many R object classes.
```{r}
  length(methods(print))
  # This number will depend on what packages are loaded.
```

```{r}
  methods(print)[1:17]
  # The first few (again, will depend on what packages are loaded).
```

There is a `print()` method for `lm` objects:
```{r}
  methods(class = "lm")
  # Results may depend on what packages are loaded.
```

```{r}
  getS3method("print", "lm")
```

* If `getS3method()` doesn’t work, try `getAnywhere(print.lm)`.
