---
title: "Matrices"
subtitle: "Lecture 6"
format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
lang: en
geometry:
  - margin = 1in
  - paperwidth = 11in
  - paperheight = 6.875in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{sectsty}
  - \allsectionsfont{\centering}
  - \pagenumbering{gobble}
execute: 
  error: true
---

# Defining Matrices Using `matrix()` 

By default, `matrix()` converts vectors to matrices in _column-major_ order...
```{r}
  matrix(1:12, nrow = 4)
```
... but this can be overridden
```{r}
  matrix(1:12, nrow = 4, byrow = TRUE)
```

\newpage

# Column-Bind and Row-Bind
```{r}
  X <- matrix(1:8, nrow = 4)
  X
  X <- cbind(rep(1, 4), X) 
  X
  X <- rbind(X, c(1, 9, 9)) 
  X
```

\newpage

# Q: What is a Matrix in R Anyway?
Matrices have a `dim` attribute which can be accessed with the `dim()` function.
```{r}
  X <- matrix(1:6, nrow = 2) 
  X 
  is.matrix(X)
  attributes(X)
  dim(X)
```


# And if we change the `dim` attribute?
```{r}
  dim(X) <- c(3,2)
  dim(X)
  X
```
Aside: `The attr()` function provides a general way to assign attributes, e.g., 
```{r}
  attr(X, "dim") <- c(3,2)
```

\newpage

# What if we remove the `dim` attribute?
```{r}
  dim(X) <- NULL
  X
  attributes(X)
  is.matrix(X)
```

\newpage

* So, a matrix in R is just a vector with a `dim` attribute
* Dimension is usually preserved when transforming elements of a matrix.
```{r}
  X <- matrix(1:6, nrow = 2) 
  X
  X^2
  X >= 4
```
* Matrices can be logical matrices (as above) and character.

# Matrix Arithmetic with Scalars
```{r}
  X <- matrix(1:6, nrow = 2) 
  X
  2 * X
  X + 1
```

# Matrix Multiplication
```{r}
  X
  Y <- matrix(sample(0:3, 9, replace = TRUE), ncol = 3) 
  Y
  X %*% Y
  Y %*% X
```

\newpage

### Matrix-Vector Multiplication
```{r}
  X
  a <- 3:1
  a
  X %*% a      # Here the vector a is treated like a column matrix.
```

\newpage

```{r}
  X
  b <- 3:4
  b
  b %*% X      # Here vector b is treated like a row matrix.
```

# Element-Wise Multiplication
```{r}
  X
  y <- 1:2
  Y <- matrix(rep(y, times = 3), nrow = 2) 
  Y
  Y * X
```

\newpage

Be very careful if you decide to do something like this.
```{r}
  X
  y
  
  y * X            ## What is this doing? (Hint: "recycling" and "column-major")
  X * y
```

\newpage

# `apply()`
Can _apply_ a function to the rows (dimension 1) or columns (dimension 2) of a matrix.
```{r}
  X
  apply(X, 1, sum)
  apply(X, 2, prod)
```
See also: `rowSums()`, `colSums()`, `rowMeans()`, `colMeans()`, and `sweep()`.

\newpage

# Named Rows and Columns
```{r}
  X <- matrix(1:6, nrow = 2) 
  dimnames(X)
  
  rownames(X) <- c("a", "b")
  colnames(X) <- paste0("x", 1:3)       # same as:  paste("x", 1:3, sep = "")
  
  X
  
  rownames(X)
  colnames(X)
```

\newpage

# More Naming
```{r}
  ages <- c("bob" = 39, "carol" = 31, "ted" = 31, "alice" = 32) 
  X <- cbind(1, ages)
  X
  
  colnames(X) <- c("intercept", "age")
  X
```

\newpage

# Adding a Named Column
```{r}
  # X <- cbind(X, sex = c("M","F","M","F"))     # This doesn't work (why?)
  X <- cbind(X, sex = c(0,1,0,1))
  X
  
  dimnames(X)
```

# Naming the Names
`dimnames(X)` is a _list_ (more on this later) and lists have names too.
```{r}
  names(dimnames(X)) <- c("Subject", "Variable")
  dimnames(X)
  X
```

\newpage

# Still More Naming
```{r}
  dimnames(X) <- NULL
  X
  
  dimnames(X) <- list(Subject = c("bob", "carol", "ted", "alice"),
                      Variable = c("intercept", "age", "sex"))
  X
```

\newpage

# Indexing Matrices
```{r}
  X[-c(2,4), c(1,3)]
  X[c("ted","alice"), c(1,3)]
  X[c("ted","alice"), c("age","sex")]
```

\newpage

**An empty index means "include all"**
```{r}
  X[c(1,3), ]
  X[ , c(1,3)]
```
```{r}
  X[c(2,4), 2]              # Expected a 2 x 1 column matrix but get a vector instead.
  
  X[c(2,4), 2, drop = FALSE] # Setting drop = FALSE preserves the dimension.
  
  X[3,2]                    # Here we get a scalar.
  
  X[3,2, drop = FALSE]      # This gives a 1 x 1 matrix
```

# Indexing with Logic
```{r}
  males <- X[, "sex"] == 0
  males <- ( X[, "sex"] == 0 )     # Same thing, but maybe this is easier to read?
  males
  X[males, ]
```

\newpage

# Transpose
```{r}
  X
  t(X)
```

# Diagonal
```{r}
  X
  diag(X)     # Extracts the main diagonal of X (even if X is not a square matrix).
  diag(1:3)   # Creates a diagonal (square) matrix with diagonal 1,2,3 and 0's elsewhere.
```

# Inverse
```{r}
  A <- matrix(rnorm(9), nrow = 3)
  A
  solve(A)
```

\newpage

```{r}
  A %*% solve(A)       # Identity matrix + floating point error
  round(solve(A) %*% A, 10)
```

# Least Squares Regression (the Hard Way)
How many feature films did the actors make after _Bob and Carol and Ted and Alice_ (1969) through 1993?  

|      Actor          | Number of Films | 
|:-------------------:|:---------------:|  
|Robert Culp (bob)    |         12      |
|Natalie Wood (carol) |          7      |
|Elliot Gould (ted)   |         30      |
|Dyan Cannon (alice)  |         16      |


\newpage

```{r}
  y <- c(bob = 12, carol = 7, ted = 30, alice = 16) 
  y
  X
```

The formula for the coefficient estimates in least squares (linear) regression is
$$ \hat{\beta} = (X'X)^{-1}X'y $$
where $X′$ represents the transpose of the design matrix $X$.
```{r}
  b <- solve(t(X) %*% X) %*% t(X) %*% y
  b
```
The fitted regression equation is
$$
\hat{y} = 93.7 − 2.1 ∗ \text{age} − 16.8 ∗ \text{sex}
$$
This is the only time we’ll do it this way.