---
title: "`forcats`: Factors"
subtitle: "STA3100 Lecture 32"
format:
  html:
    embed-resources: true
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    geometry:
      - margin = 1in
      - paperwidth = 11in
      - paperheight = 6.875in
    header-includes:
      - \usepackage{booktabs}
      - \usepackage{sectsty}
      - \allsectionsfont{\centering}
      - \pagenumbering{gobble}
lang: en
execute:
  error: true
---

# Factors in Base-R

Why factors? 

* Factors are used to represent qualitative/categorical variables in R.
* Factors are important for fitting statistical models with qualitative predictors.
* Qualitative data may be
  - _nominal_: e.g., party affiliation, religious affiliation, gender.
  - _ordinal_: e.g., socioeconomic status, highest degree attained, strength of agreement with a statement.
* Nominal variables are represented by factors.
  - Class:`"factor"`
* Ordinal variables can additionally be represented by ordered factors.
  - Class: `"ordered"` inherits from `"factor"`.

***
  
**Creating a Factor**
```{r}
#| message: false

x <- sample(rep(c("M","F"), 3), 6)            # x is a character vector 
x   
f <- factor(x)                                # Creating a factor f 
f
levels(f)                                     # Used to get or modify levels of a factor
```

* Note that level `F` comes before level `M`. By default, `factor()` orders the levels “alphabetically” (according to the current locale).

```{r}
attributes(f)
is.character(f)
is.factor(f)
as.character(f)       # Coercion to character gives factor levels of obs. as character.
```

So far everything looks pretty simple, but consider this:

```{r}
typeof(f)             # The internal type/storage mode
is.integer(f)         # See R help for is.integer
is.numeric(f)         # See R help for is.numeric
```

The storage type is `integer`, but `is.numeric(f)` returns FALSE because:

* arithmetic with factors (_qualitative_ variables) does not make sense, and
* for unordered factors, the levels cannot be meaningfully compared (`<`, `<=`, `==`, `>=`, `>`) using the underlying integers.

```{r}
f
unclass(f)      # Stripping off the class ("factor")
as.integer(f)   # Coerced to integer (there is rarely any reason to do this)
```

* An observation with the 1st factor level (`F`) is stored as integer 1 (1L).
* An observation with the 2nd factor level (`M`) is stored as integer 2 (2L).
* If the factor had 3 levels, an obs. at the 3rd factor level would be stored as 3L.
* Etc.

***

**Changing and Rearranging the Levels of a Factor**

* Modeling functions typically treat the first level of a factor as the _control_ level. 
  - We may wish to _reorder_ the levels of a factor to make some other level
first/control.
  - E.g., make M the control level instead of F.
* We may wish to _change_ the labels for the levels of a factor for readability: 
  - E.g., change levels of age from
      `014  1524  2534  3544  4554  5564 65`
      to
      `0-14  15-24  25-34  35-44  45-54  55-64  65+`
      
**How NOT to Reorder Levels**      
```{r}
f2 <- f 
f2
levels(f2) <- c("M", "F")       # Trying to make M first. 
levels(f2)
f2
```

* **NOT** what we intended. Changed all the M’s to F’s and the F’s to M’s!

*** 

What Happened?

```{r}
as.integer(f)
levels(f)
as.integer(f2)
levels(f2)
```

* The underlying integer representation didn’t change.
* We just changed the labels associated with those integer values.
* We can correctly use `levels()<-` to change, e.g., “F” to “Female” and “M” to “Male”, but _not_ to change the order of the levels:

```{r}
f
f2 <- f
levels(f2) <- c("Female", "Male")
f2
```

Correct Ways to Make “M” First with Base-R

* When constructing the factor, you can explicitly set `levels`.
```{r}
x
f2 <- factor(x, levels = c("M", "F")) 
f2
```

* Also useful if there are levels that are not observed in the current data.

```{r}
factor(x, levels = c("M", "F", "NB"))
```

* With an existing factor, can use base-R’s `relevel()` function.

```{r}
f
f2 <- relevel(f, ref = "M") 
f2
```

* Notice that the integer associated with “M” is now 1; before “F” was 1.

```{r}
as.integer(f2)
```

# Ordered Factors
```{r}
x <- rep(c("Agree", "Neutral", "Disagree"), times = c(3, 1, 2)) 
x
of <- ordered(x) 
of
```

* This isn’t right (because levels defaulted to alphabetical order).
* Specifying the levels explicitly fixes the ordering.

```{r}
of <- ordered(x, levels = c("Disagree", "Neutral", "Agree")) 
of
```

* Again, can include levels not observed in the data (esp. useful if merging/joining datasets).

```{r}
of2 <- ordered(x,
              levels = c("Strongly Disagree", "Disagree",
                         "Neutral", "Agree", "Strongly Agree"))
of2
of
attributes(of)
is.ordered(of)
is.factor(of)   # `of` is already a factor so `as.factor(of)` won't change it.
```

*** 

* Unordered version can be obtained with `factor()` (rarely any reason for this).

```{r}
f <- factor(of, ordered = FALSE) 
is.ordered(f)
f
```

**The Wrong Way to Reorder Levels (again)**

* Suppose we want to order by level of disagreement rather than agreement.

```{r}
of
levels(of)
of2 <- of
levels(of2) <- rev(levels(of2))     # WRONG!!!
of2                                 # "Agree"s and "Disagree"s are flipped!
```

**A Correct Way to Reverse the Ordering of Levels**
```{r}
of
of2 <- ordered(of, levels = rev(levels(of))) 
of2
```

***

# Creating Factors from Numeric Variables

A simple Weibull model for household incomes.

```{r}
inc_shape <- 1.2 
inc_scale <- 100000
round(qweibull(c(0.10, 0.25, 0.50, 0.75, 0.90), shape = inc_shape, scale = inc_scale))
```

What is the shape of this distribution’s density?

**Base-R Plot**
```{r}
#| fig-cap: "Weibull density with shape parameter 1.2 and scale parameter 100,000."

opar <- par(mar = c(5.1, 4.1, 0.1, 0.1)) 
curve(dweibull(x, shape = inc_shape, scale = inc_scale),
      xlim = c(0, qweibull(0.99, shape = inc_shape, scale = inc_scale)), 
      xlab = "Household Income",
      ylab = "Density")
par(opar)
```

**ggplot**
```{r}
#| echo: false

library(ggplot2)
```

```{r}
#| fig-cap: "Weibull density with shape parameter 1.2 and scale parameter 100,000."

ggplot() +
    geom_function(fun = dweibull, args = list(shape = inc_shape, scale = inc_scale)) +
    xlim(c(0, qweibull(0.99, shape = inc_shape, scale = inc_scale))) +
    labs(x = "Household Income", y = "Density")
```

The second one is better:
```{r}
#| fig-cap: "Weibull density with shape parameter 1.2 and scale parameter 100,000."

ggplot() +
    geom_function(fun = dweibull, args = list(shape = inc_shape, scale = inc_scale)) +
    labs(x = "Household Income", y = "Density") +
    scale_x_continuous(
        limits = c(0, qweibull(0.99, shape = inc_shape, scale = inc_scale)),
        labels = scales::label_comma() 
    )
```

* Drawing a small sample.

```{r}
income <- round(rweibull(15, shape = inc_shape, scale = inc_scale)) 
income
```

* Creating groups by dividing `range(income)` into 5 intervals of equal length.

```{r}
income_group <- cut(income, breaks = 5) 
income_group
```

* Providing explicit breaks and measuring in thousands of dollars

```{r}
income
income_group <- cut(income/1000, breaks = c(0, 15, 30, 50, 75, 100, 150, 200, Inf))
income_group
table(income_group)
```

* Using equally spaced breaks and measuring in tens-of-thousands of dollars.

```{r}
breaks<-seq(from = 0, to = 20, by = 5) 
breaks <- c(breaks, Inf)
income_group <- cut(income/10000, breaks) 
income_group
table(income_group)
```

* Equally spaced, providing our own labels, and measuring in thousand of dollars.


```{r}
breaks <- seq(from = 0, to = 200, by = 50)

labels <- paste(breaks[-length(breaks)], breaks[-1], sep = "-") 
labels <- c(labels, max(breaks))
labels <- paste0("$", labels, "K")
labels[length(labels)] <- paste0(labels[length(labels)], "+")

breaks <- c(breaks, Inf)

income_group <- cut(income/1000, breaks, labels = labels) 

table(income_group)
```

* `cut()` returns an un-ordered factor by default.
* Can change this with `ordered_result` argument.

```{r}
income_group <- cut(income/1000, breaks, labels = labels, ordered_result = TRUE)
income_group
```

# Factors in the Tidyverse
# Forcats

The `tidyverse` package `forcats` provides functions that solve common problems related to factors.

* `as_factor()` : generic, similar to base-R `as.factor()` but levels default to the order in which they appear in the data (unless ‘x’ is numeric).

Changing Levels

* `fct_recode()` : change levels 
  - compare to `levels() <-` 
  - can also use to combine levels
* `fct_collapse()` : similar to `fct_recode()` but aimed at combining levels
* `fct_lump_min()`, `fct_lump_prop()`, `fct_lump_n()`, `fct_lump_lowfreq()` : combine infrequent levels to “Other”.

Reordering Levels

* `fct_relevel()` : move specified factor level(s) to front (similar to `relevel()` but more general).
  - has `after` agrument.
* `fct_rev()` : reverse levels of factor. 
  - If only plotting, can use `coord_flip()` in `ggplot2`.
* `fct_shuffle() `: function that randomly reorders factor levels (not sure why
we would need this).
* `fct_infreq()` : order levels by frequency (automatically puts NA at small end, regardless of its freq).
* `fct_reorder()` : reorder levels by values of another variable (useful for tables and 1d plots) 
* `fct_reorder2()` : reorder factor levels by last values of another variable (useful for 2d plots)

***

**HCV Vertical Transmission (Mother to Infant)**

* Original data were collected by Paul Wasuwanich and collaborators at Shands.
* The data presented here have been altered:
  - Only 4 columns retained.
  - Some columns randomly shuffled.
  - The responses (`infant_hcv`) are randomly generated.

```{r}
#| message: false

library(tidyverse)
hcv <- read_csv("hcv.csv")
hcv
```

* Convert `hcv_genotype` and `bloodtype` to factors.

```{r}
hcv <- hcv |> 
  mutate(hcv_genotype = as_factor(hcv_genotype), 
         bloodtype = as_factor(bloodtype))
hcv
```

# Combining Levels

* Many of the genotype frequencies are very low.
```{r}
hcv |> count(hcv_genotype)   # Same as hcv |> summarize(n = n(), .by = hcv_genotype)
```
Could have also used `fct_count(hcv$hcv_genotype)` here.

* Genotype is the numeric part; letter indicates subtype. Recoding to just the primary genotype.

```{r}
hcv <- hcv |> 
  mutate(hcv_genotype = fct_collapse(hcv_genotype, 
                                     "1" = c("1", "1A", "1B"), 
                                     "2" = c("2", "2B"),
                                     "3" = "3",
                                     "4" = "4",
                                     "6" = "6"))
hcv |> count(hcv_genotype)
```

# Lumping

* May wish to combine low frequency categories under “Other”.
* `fct_lump_min()` lumps levels that appear fewer than `min` times.

```{r}
hcv |> 
  mutate(hcv_genotype = fct_lump_min(hcv_genotype, 10)) |> 
  count(hcv_genotype)
hcv |> 
  mutate(hcv_genotype = fct_lump_min(hcv_genotype, 20)) |> 
  count(hcv_genotype)
```

* `fct_lump_prop()` lumps levels that appear in fewer than `prop` proportion of the observations.

```{r}
hcv |> 
  mutate(hcv_genotype = fct_lump_prop(hcv_genotype, 0.10)) |> 
  count(hcv_genotype)
```

* `fct_lump_n()` lumps all levels except for the `n` most frequent.

```{r}
hcv |> 
  mutate(hcv_genotype = fct_lump_n(hcv_genotype, 3)) |> 
  count(hcv_genotype)
hcv |> 
  mutate(hcv_genotype = fct_lump_n(hcv_genotype, 2)) |> 
  count(hcv_genotype)
```

* “`fct_lump_lowfreq()` lumps together the least frequent levels, ensuring that “other” is still the smallest level.” (This description seems imprecise to
me.)
* Here are the bloodtypes in order of frequency of occurence in the dataset.
  - Also previews use of `fct_infreq()` which orders levels by frequency.

```{r}
hcv |> mutate(bloodtype = fct_infreq(bloodtype)) |> 
  count(bloodtype)
```
  
* And here is what `fct_lump_lowfreq()` does with those frequencies.

```{r}
hcv |> 
  mutate(hcv_genotype = fct_lump_lowfreq(bloodtype)) |> 
  count(hcv_genotype)
```

# Reordering Levels

* Most common blood type in U.S. is O+.
* Suppose we want `O+` to be the control/reference level for bloodtype.
  - Have seen base-R function `relevel()` (moves one category to the front).
  - `forcats` has `fct_relevel()` which is more general.
* Let’s make `O+` first while keeping `O+` and `O-` together.

```{r}
hcv <- hcv |> 
  mutate(bloodtype = fct_relevel(bloodtype, c("O+", "O-")))
hcv |> count(bloodtype)
```

* `fct_rev()` may also be useful, esp. for ordered factors.
```{r}
of
fct_rev(of)
```

## Reordering Levels by Frequency

* Example: A barplot may be easier to interpret if categories are ordered by their frequencies.
* Here are barplots of a factor with and without reordering levels by frequency.

```{r}
#| fig-cap: "Blood type of mother."

ggplot(hcv, aes(x = bloodtype)) + geom_bar()
```
```{r}
#| fig-cap: "Blood type of mother."

ggplot(hcv, aes(x = fct_infreq(bloodtype))) + geom_bar()
```

* Here is the same for HCV genotype, this time with nicer axis labels.
* Notice that `NA` is last regardless of its frequency.

```{r}
#| fig-cap: "HCV genotype carried by mother."

ggplot(hcv, aes(x = hcv_genotype)) + 
  labs(x = "Genotype", y = "Frequency") + 
  geom_bar()
```

```{r}
#| fig-cap: "HCV genotype carried by mother."

ggplot(hcv, aes(x = fct_infreq(hcv_genotype))) + 
  labs(x = "Genotype", y = "Frequency") + 
  geom_bar()
```

## Reordering Levels by Values of Another Variable

```{r}
transmission_by_bloodtype <- hcv |> 
  group_by(bloodtype, infant_hcv) |> 
  summarise(n = n()) |> # up to 8 * 2 = 16 rows, 3 cols (bloodtype, infant_hcv, n)
  pivot_wider(names_from = infant_hcv, values_from = n, values_fill = 0L) |>
  mutate(n = N + Y, rate = Y / n)
transmission_by_bloodtype
```

* `fct_reorder()` reorders factor levels by values of another variable
* Compare these two plots.

```{r}
#| fig-cap: "Vertical tranmission rate by blood type."

transmission_by_bloodtype |> 
  ggplot(aes(x = rate, y = bloodtype)) + 
  geom_point()
```

```{r}
#| fig-cap: "Vertical tranmission rate by blood type."

transmission_by_bloodtype |> 
  ggplot(aes(x = rate, y = fct_reorder(bloodtype, rate))) + 
  geom_point()
```

* Another version of the last plot with:
  - size of dots indicating number of observations
  - better labels

```{r}
#| fig-cap: "Vertical tranmission rate by blood type."

transmission_by_bloodtype |> 
  ggplot(aes(x = rate, y = fct_reorder(bloodtype, rate), size = n)) + 
  geom_point() + 
  labs(x = "Transmission Rate", y = "Bloodtype", size = "Sample Size")
```

```{r}
hcv <- hcv |> 
  mutate(abo = fct_collapse(bloodtype, 
                            A = c("A+", "A-"),
                            B = c("B+", "B-"), 
                            AB = c("AB+", "AB-"), 
                            O = c("O+", "O-")),
         rh = fct_collapse(bloodtype,
                           "pos" = c("A+", "B+", "AB+", "O+"),
                           "neg" = c("A-", "B-", "AB-", "O-"))
)
hcv |> count(abo) 
hcv |> count(rh)
```


```{r}
transmission_by_abo <- hcv |> 
  group_by(abo, infant_hcv) |>
  summarise(n = n()) |> # 4 * 2 = 8 rows, 3 columns (abo, infant_hcv, n)
  pivot_wider(names_from = infant_hcv, values_from = n, values_fill = 0L) |>
  mutate(n = N+Y, rate = Y/n)
transmission_by_abo
```

```{r}
#| fig-cap: "Vertical tranmission rate by blood group."

transmission_by_abo |> 
  ggplot(aes(x = rate, y = fct_reorder(abo, rate), size = n)) + 
  geom_point() + 
  labs(x = "Transmission Rate", y = "Bloodtype", size = "Sample Size")
```
