---
title: "`lubridate`: Dates and Times"
subtitle: "STA3100 Lecture 19"
format:
  html:
    embed-resources: true
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    geometry:
      - margin = 1in
      - paperwidth = 11in
      - paperheight = 6.875in
    header-includes:
      - \usepackage{booktabs}
      - \usepackage{sectsty}
      - \allsectionsfont{\centering}
      - \pagenumbering{gobble}
lang: en
execute:
  error: true
---

# THE `LUBRIDATE` PACKAGE
# Dates with `lubridate`
`lubridate` 

* tries to make it easier to work with dates and date-times. 
* does not introduce new classes for dates or date-times.
* has 3 new difftime-like classes ("durations”, "periods”, and "intervals”).

**Note**

* `lubridate` is _not_ part of the core `tidyverse` and the `readr` package does _not_ use `lubridate` when reading dates and date-times.
  - dates are handled with `parse_date()`
  - date-times are handled with `parse_date_time()`
  - Some of `lubridate`’s functionality is based on these functions.

  More on `readr` later.

***

```{r}
#| message: false

	library(tidyverse)
	# lubridate is not loaded by tidyverse package.
	library(lubridate)
	# lubridate version of Sys.Date()
	today <- today()
	class(today)

	# lubridate version of Sys.time()
	now <- now()        
	class(now)
```


# Creating Dates
| Purpose                               | `lubridate` function(s)       |Base-R function(s)       |
|:--------------------------------------|:------------------------------|:------------------------|  
|Current date                           |`today()`                      |`Sys.Date()`             |
|General conversion to date             |`as_date()`                    |`as.Date()`              |
|Convert character string to date       |`ymd()`, `mdy()`, `dmy()`, etc.|`as.Date()`              |
|Make date from numeric year, month, day|`make_date()`                  |`ISOdate()` + `as.Date()`|
    
# Dates from Character Strings

```{r}
	as_date("1980-12-08")         # John Lennon killed by a revolver.
```

* To facilitate conversion of strings to dates `lubridate` provides `ymd()`, `ydm()`, `myd()`, `mdy()`, `dym()`, and `dmy()`.
  - Choose function according to order of year, month, and day in string.
  - The function will try to figure out the rest of the format, e.g.,
    + Is month represented by “1” or “01” or “Jan” or “January”?
    + Are fields separated by “-” or “/” or “,” or “”, or whatever.

***

**These are pretty easy with `lubridate`.**
```{r}
	mdy("Dec 8, 1980")
	mdy("12/8/80")
	mdy("December 8 (1980)")
```

***

**Base-R requires more from the user.**
```{r}
	as.Date("Dec 8, 1980", format = "%b %d, %Y")
	as.Date("12/8/80", format = "%m/%d/%y")
	as.Date("December 8 (1980)", format = "%B %d (%Y)")
```

***

`?strptime`

`Details:`

`...`

    '%b' Abbreviated month name in the current locale on this platform.
          (Also matches full name on input: in some locales there are
          no abbreviations of names.)

    '%B' Full month name in the current locale.  (Also matches abbreviated
          name on input.)

***

**But it is possible to come up with a format that `dmy()` can’t parse.**
```{r}
	dmy("08121980")
	dmy("081280")
	dmy("8121980")
	dmy("8 12 1980")       # Separators help a lot.
```

# Dates from Pieces of Dates
`make_date()` : useful if numeric year, month, and day stored in separate cols.

```{r}
	make_date(1980, 12, 8)
	make_date(1980, 12)
	make_date(1980)
```

# Creating Date-Times

| Purpose                                                              | `lubridate` function(s)     |Base-R function(s)          |
|:---------------------------------------------------------------------|:----------------------------|:---------------------------|  
|Current date-time                                                     |`now()`                      |`Sys.time()`                |
|General conversion to date-time                                       |`as_datetime()`              |`as.POSIXct()`              |
|Convert character string to date-time                                 |`ymd_hms()`, `mdy_hm()`, etc.|`strptime()`, `as.POSIXct()`|
|Make date-time from numeric year, month, day, hour, minute, and second|`make_datetime()`            |`ISOdatetime()`             |

# Example: TV Times of NCAA Tournament Games

```{r}
	cat(readLines("tvRndOf32-2019.csv", n = 4), sep = "\n")
	tv <- read_csv("tvRndOf32-2019.csv")
```

***
  
```{r}
	tv |>
		slice_head(n = 8)
```

* Date and time are split across two columns.
* Format(s) may be problematic (read as character).
* Some times are specified as `After conc. I`.

***

```{r}
	tv |>
		pull(`Time (ET)`)      # Same as tv$`Time (ET)` or tv[["Time (ET)"]]
```

* Let’s try to construct date-times anyway.
```{r}
	tv |>
		mutate(dt = ymd_hm(str_c(Day, `Time (ET)`, sep = " "))) |>
			pull(dt)
```
       - Note: `stringr::str_c()` is very similar to `paste0()`.
  
***

* Notice that the time zone is set to UTC (Universal Coordinated Time).
* Run `OlsonNames()` for list of time zones.

```{r}
	tv |>
		mutate(dt = ymd_hm(str_c(Day, `Time (ET)`, sep = " "), tz = "America/New_York")) |>
			pull(dt)
```

* “`After conc. I`” rows leave us with `NA`, which seems reasonable.
* Any other problems?

***

* “p.m.” is not being used.
```{r}
	ymd_hm("2019-03-23 5:15 p.m.", tz = "America/New_York")
```

  
* How would the correct time be formatted with AM/PM in our current locale?
  - Note use of `%I` for the hour with `%p` for the AM/PM indicator.
```{r}
	strftime(as.POSIXct("2019-03-23 17:15:00"), "%Y-%m-%d %I:%M:%S %p")
```

* Checking that “PM” works where “p.m.” does not.

```{r}
	ymd_hm("2019-03-23 5:15 PM", tz = "America/New_York")
```

***

`?strptime`

`Details:`

    '%H' Hours as decimal number (00-23).  As a special exception strings
          such as '24:00:00' are accepted for input, since ISO 8601
          allows these.

    '%I' Hours as decimal number (01-12).
    
    '%p' AM/PM indicator in the locale.  Used in conjunction with '%I' and
          *not* with '%H'.  An empty string in some locales (for
          example on some OSes, non-English European locales including
          Russia). The behaviour is undefined if used for input in such
          a locale.
          
          Some platforms accept '%P' for output, which uses a
          lower-case version ('%p' may also use lower case): others
          will output 'P'.

***

* Use `stringr::str_replace()` to replace “p.m.” with “PM”.
  - Could use base-R `sub()` or `gsub()` instead.

```{r}
	tv <- tv |>
		mutate(`Time (ET)` = str_replace(`Time (ET)`, "p.m.", "PM"))
	tv |>
		pull(`Time (ET)`)
```

* Note: the following code replaces all “a.m.” with “AM” and all “p.m.” with “PM”. Not required here but would have also worked.

```{r}
	tv <- tv |>
		mutate(`Time (ET)` = str_replace_all(`Time (ET)`, c(a.m. = "AM", p.m. = "PM")))
```

***

```{r}
	tv <- tv |>
		mutate(dt = ymd_hm(str_c(Day, `Time (ET)`, sep = " "), tz = "America/New_York"))
	tv$dt
```

* More on this example in next lecture.

# Extracting Date and Date-Time Components

| Purpose        | `lubridate` function(s)                        |Base-R function(s)          |
|:---------------|:-----------------------------------------------|:---------------------------|  
|Extract year    |`year()`                                        |`strftime()` , `format()`   |
|Extract month   |`month()`                                       |`strftime()` , `format()`   |
|Extract day     |`day()`, `yday()`, `qday()`, `mday()`, `wday()` |`strftime()` , `format()`   |
|Extract hour    |`hour()`                                        |`strftime()` , `format()`   |
|Extract minute  |`minute()`                                      |`strftime()` , `format()`   |
|Extract second  |`second()`                                      |`strftime()` , `format()`   |

***

* `strftime()` always returns type `character`. Use `as.integer()` (or `as.double()` for seconds) to convert to numeric.
* Any of the `lubridate` functions in the previous table can also be used to **set** a date component to a new value.
* `update()` also can be used to change a date or date-time:
  - `lubridate` defines `update()` methods for classes `Date` and `POSIXt`.
  - `POSIXct` and `POSIXlt` both inherit from `POSIXt`.

***

```{r}
	year(as_date("1967-12-10"))        # Otis Redding dies in plane crash.
	month(as_date("1967-12-10"))
	day(as_date("1967-12-10"))
```

***
  
```{r}
	wday(as_date("1967-12-10"))
	mday(as_date("1967-12-10"))
	qday(as_date("1967-12-10"))
	yday(as_date("1967-12-10"))
```

***  

```{r}
	wday(as_date("1967-12-10"), label = TRUE)
	is.factor(wday(as_date("1967-12-10"), label = TRUE))       # More on factors ...
	is.ordered(wday(as_date("1967-12-10"), label = TRUE))      # ... and ordered factors later.
	as.numeric(wday(as_date("1967-12-10"), label = TRUE))
```

***

```{r}
	month(as_date("1967-12-10"), label = TRUE)
	as.numeric(month(as_date("1967-12-10"), label = TRUE))
```

# Time Spans

`lubridate` defines three new classes that are similar to difftime:

* `duration` : most like difftime but always coded in seconds.
  - 1 min = 60 sec, 1 hr = 60 min, 1 day = 24 hr, 1 wk = 7 days
  - 1 month = 4.35 wks, 1 yr = 365.25 days.
* `period` : conforms to “normal” expectations when added to dates and times.
* `interval` : removes ambiguity by having a starting point.
* Please read the [`lubridate` vignette](https://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html).

***

|Class      | Construction Functions                                                                                                                    |
|:----------|:------------------------------------------------------------------------------------------------------------------------------------------|  
|`duration` |`duration()`, `dseconds()`, `dminutes()`, `dhours()`, `ddays()`, `dweeks()`, `dmonths()`, `dyears()`, `dmilliseconds()`, `dmicroseconds()`,|
|           |`dnanoseconds()`, `dpicoseconds()`                                                                                                         |
|`period`   |`period()`, `seconds()`, `minutes()`, `hours()`, `days()`, `weeks()`, `months()`, `years()`, `milliseconds()`, `microseconds()`,           |
|           | `nanoseconds()`, `picoseconds()`                                                                                                          |
|`interval` |`interval()`, `start %--% end`                                                                                                             |

# Fun With Dates and Times

```{r}
	dyears(1)           # A duration
	years(1)            # A period
	as_date("2019-07-01") + dyears(1)    # Adding a duration
	as_date("2019-07-01") + years(1)     # Adding a period
```

***

```{r}
	dhours(24)          # A duration
	hours(24)           # A period
	ymd_hms("2022-11-05 12:00:00", tz = "America/New_York") + dhours(24)    # Surprised?
	ymd_hms("2022-11-05 12:00:00", tz = "America/New_York") + hours(24)     # What you expected?
```

# Intervals

**A trip to Brevard**
```{r}
	arrive <- ymd_hm("2022-07-25 18:25", tz = "America/New_York") 
	depart <- ymd_hm("2022-08-15 11:22", tz = "America/New_York") 
	brevard <- interval(arrive, depart)                # Or arrive %--% depart 
	brevard
```

```{r}
	str(brevard)
```
***

**Sister Decides to Go to Brevard Too**

```{r}
	sis_arr <- ymd_h("2022-07-28 23", tz = "America/New_York")
	sis_dep <- ymd_h("2022-08-01 13", tz = "America/New_York")
	sis <- sis_arr %--% sis_dep
	sis
```

***

* Sister decides to move her trip up by a week.

```{r}
	sis <- int_shift(sis, -weeks(1))
	sis
```

* And adds 2 days to her stay.

```{r}
	int_end(sis) <- int_end(sis) + days(2)
	sis
```

***

* Do our stays overlap?

```{r}
	brevard
	sis
	int_overlaps(brevard, sis)
	intersect(brevard, sis)
```

# THE `LUBRIDATE` PACKAGE (EXTRA NOTES)
# TV Times Example Continued
Let’s read the data again and get back into it.
```{r}
#| message: false

	library(tidyverse)
	library(lubridate)
	tv <- read_csv("tvRndOf32-2019.csv")
```

***


* We constructed date-times for the first (I) game at each site as follows.

```{r}
	tv <- tv |>
		mutate(`Time (ET)` = str_replace_all(`Time (ET)`, c("a.m." = "AM", "p.m." = "PM"))) |>
			mutate(dt = ymd_hm(str_c(Day, `Time (ET)`, sep = " "), tz = "America/New_York"))
	tv |>
		select(Day, `Time (ET)`, Site, dt) |>
			slice_head(n = 8)
```

***

* Let’s separate the `Site` column into the site and game number (I or II).
* The `tidyr` function `separate_wider_regex()` is perfect for this.
* `.` will match any character
* `*` lets a pattern be optional or repeat (i.e. it matches any number of times, including 0)
* `?` makes a pattern optional (i.e. it matches 0 or 1 times)  
For more details, please see [https://r4ds.hadley.nz/regexps](https://r4ds.hadley.nz/regexps)
```{r}
	tv <- tv |>
		separate_wider_regex(Site, c(Site = ".*", " ", Number = "II?"))
	tv |>
		select(Day, `Time (ET)`, Site, Number, dt) |>
			slice_head(n = 7)
```


***

# Add Approximate Date-Times for the II Games

* According to [this source](https://www.scottfujita.com/how-long-is-a-college-basketball-game/), average game lasts 2 hours and 10 minutes.
* How much time between games at a tournament site? Guess: 30 minutes.
* So to get approximate times for a II-game, we find the I-game at the same site and add 2 hours and 40 minutes to its time.
* An easy way to do this is:
  1. create a data frame with the sites, number, and times for just the I-games.
  2. add 2 hours and 40 minutes to the times of these games.
  3. change the number to “II”, thus creating a data frame of sites, numbers, and approximate date-times for the II-games.
  4. `left_join()` the original data with these now II-games back into the original data frame by matching on site and number.

***

```{r}
	tv2 <- tv |>
		filter(Number == "I") |>
			select(Site, Number, dt) |>
				mutate(Number = "II",
		dt = dt + dhours(2) + dminutes(40))
	tv2
```

***

* `left_join()` creates columns `dt.x` and `dt.y`
* Combine them with `coalesce()`.

```{r}
	tv |>
		left_join(tv2, by = c("Site", "Number")) |>
			mutate(dt = coalesce(dt.x, dt.y)) |>
				select(Site, Number, dt.x, dt.y, dt) |>
					head(n = 8)
```

***
  
* Alternatively, the function `dplyr::rows_update()` avoids creating the `dt.x` and `dt.y` columns and does exactly what we need here.

```{r}
	tv |>
		rows_update(tv2, by = c("Site", "Number")) |>
			select(Site, Number, dt) |>
				head(n = 8)
```

***

* So let’s do this and then
  - delete the old columns that we don’t need
  - rename the dt column

```{r}
	tv <- tv |>
		rows_update(tv2, by = c("Site", "Number"))  |>
			select(-Day, -`Time (ET)`) |>
				rename("Date_Time" = dt)
```

* Have a look at the result on the next slide.
* The next thing one might want to do is to separate the "Game" column into 4 columns containing the two teams and their numeric seeds.

***

```{r}
	tv |>
		slice_head(n = 14)
```
