---
title: "Type Conversion"
subtitle: "STA3100 Lecture 3"
format:
  html:
    embed-resources: true
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
lang: en
geometry:
  - margin = 1in
  - paperwidth = 11in
  - paperheight = 6.875in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{sectsty}
  - \allsectionsfont{\centering}
  - \pagenumbering{gobble}
execute:
  error: true
---

\newpage

# Numeric Data Types

As we have seen, R usually stores numeric data as double precision by default.
```{r}
x <- c(1,2,3)
x
typeof(x)   # Returns the R internal type or storage mode of x.
```
But you can enter integers explicitly with the L suffix:
```{r}
x <- c(1L, 2L, 3L) 
x
typeof(x)
```

\newpage

And some functions return integer values:
```{r}
x <- 1:3
x
typeof(x)
```

\newpage

# Casting (Explicit Type Conversion)
You can _cast_ between types (explicitly convert from one type to another):
```{r}
x
typeof(x)

y <- as.double(x) 
y
typeof(y)

y <- as.character(y) 
y
typeof(y)
```

\newpage

* numerical to logical: 0 $\mapsto$ `FALSE` ; nonzero $\mapsto$ `TRUE`
* logical to numeric: `FALSE` $\mapsto$ 0 ; `TRUE`$\mapsto$  1

```{r}
x <- 0:3
x

y <- as.logical(x)
y

as.integer(y) 
as.double(y) 
as.character(y)
```

\newpage

# Implicit Type Conversion

* R often _implicitly_ coerces values to an (hopefully) appropriate type.
* Here are two **very useful** tricks that take advantage of this.
```{r}
y

sum(y)   # Number true : equal to sum(as.integer(y)) 

mean(y)  # Proportion true : equal to sum(as.integer(y))
```

```{r}
iq <- round(rnorm(18, mean = 100, sd = 15))
iq
sum(iq > 105)
```
    
\newpage

* If using integers and doubles together in arithmetic, R implicitly coerces the integers to double in order to do the calculation.
```{r}
typeof(1:3)

4L + (1:3)            # All the numbers involved are integers
typeof(4L + (1:3)) 

4.0 + (1:3)           # 4.0 is double ; 1:3 is integer
typeof(4.0 + (1:3))
```  

\newpage

* Similar implicit rules are applied in many other situations.
* E.g., all the elements of a vector have to be of the same type, so if forming a vector from elements of different types, R coerces them to the same type when it can do so without losing (much) information or precision.

```{r}
c(1:3, 4L)                # 1:3 = c(1L, 2L, 3L) and 4L all integers
typeof(c(1:3, 4L)) 

c(1:3, 4.0)               # 1:3 = c(1L, 2L, 3L) integer ; 4.0 double
typeof(c(1:3, 4.0)) 

c(1:3, "4")               # 1:3 = c(1L, 2L, 3L) integer ; "4" character
typeof(c(1:3, "4"))
```

\newpage

* Implicit type conversion happens in many other situations.

```{r}
x <- as.character(1:3) 
x

paste("q", x, sep = "_")      # Pasting together character vectors ("q" is recycled)

paste("q", 1:3, sep = "_")    # 1:3 is integer ; converted to character inside paste().

paste("That's ", c(TRUE, FALSE), "!", sep = "")   # logical converted to character.
```
