---
title: "TDA demo"
author: "Benjamin Roycraft"
format: 
  html:
    page-layout: custom
server: shiny
---

## Topological Data Analysis and Persistent Homolgy

This document demonstrates a little bit the behavior of topological statistics from persistent homology.

The first plot shows an example 2-D data set, both with some underlying shape and with additive Gaussian noise.

The second plot shows the persistence diagram for each, considering only 1-D homology (loops) for the Vietoris-Rips complex. 0-D homology gives the connected components which merge over time, and isn't shown.

2-D homology can be useful, but is more computationally expensive and only really provides interesting signals for 3-D data or higher.

```{r}
#| panel: sidebar

sliderInput("shape_x", "x Shape", 1, min = 1, max = 9)
sliderInput("shape_y", "y Shape", 1, min = 1, max = 9)
sliderInput("n", "Sample Size", 10^2, min = 1, max = 10^3)
sliderInput("sd", "Noise Level", 0, min = 0, max = .3, step = 10^-3)
actionButton("regen", "Regenerate")
sliderInput("thresh", "Diameter Threshold (Lower = Faster)", 2, min = 0, max = 2, step = 10^-3)
```

```{r}
#| panel: fill

plotOutput('plot1')
```

```{r}
#| panel: fill

plotOutput('plot2')
```

```{r}
#| context: server

X <- reactive({
  t <- seq(0, 2*pi, length.out=input$n)
  cbind(
    sin(input$shape_x*t),
    cos(input$shape_y*t)
  )
})

Z <- reactive({
  input$regen
  out <- rnorm(2*input$n)
  dim(out) <- c(input$n, 2)
  return(out)
})

Y <- reactive( X()+input$sd*Z() )

diag_X <- reactive({
  ripserr::vietoris_rips(X(), max_dim=2, threshold=input$thresh)
})

# diag0_X <- reactive({
#   as.matrix(subset(diag_X(), dimension == 0, select = c(birth, death)))
# })

diag1_X <- reactive({
  as.matrix(subset(diag_X(), dimension == 1, select = c(birth, death)))
})

diag_Y <- reactive({
  ripserr::vietoris_rips(Y(), dim=2, threshold=input$thresh)
})

# diag0_Y <- reactive({
#   as.matrix(subset(diag_Y(), dimension == 0, select = c(birth, death)))
# })

diag1_Y <- reactive({
  as.matrix(subset(diag_Y(), dimension == 1, select = c(birth, death)))
})

output$plot1 <- renderPlot({
    plot(Y(),
         xlim=c(-1.5,1.5),
         ylim=c(-1.5,1.5),
         asp=1,
         pch=19,
         col="blue")
    points(X(),
           pch=19,
           col="red")
})

output$plot2 <- renderPlot({
    
    d_max <- max(diag_X()$death, diag_Y()$death)
  
    plot(c(),
         xlim=c(0, d_max),
         ylim=c(0, d_max),
         asp=1,
         xlab="Birth",
         ylab="Death",
         main="1-D Persistent Homology"
    )
    segments(0,0,0,d_max)
    segments(0,0,d_max,d_max)
    # points(diag0_X(), pch=1, col="red")
    points(diag1_X(), pch=2, col="red")
    # points(diag0_Y(), pch=1, col="blue")
    points(diag1_Y(), pch=2, col="blue")
})
```