---
title: "Importing and Exporting Data"
subtitle: "STA3100 Lecture 26"
format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
lang: en
geometry:
  - margin = 1in
  - paperwidth = 11in
  - paperheight = 6.875in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{sectsty}
  - \allsectionsfont{\centering}
  - \pagenumbering{gobble}
execute: 
  error: true
---

# Initial Reading of the Data

The file `schedule_b-2023-10-27T01_16_01.csv` was downloaded from the
Federal Election Commission’s [\textcolor{blue}{website}](https://www.fec.gov/data/disbursements/?data_type=processed&two_year_transaction_period=2024&min_date=01%2F01%2F2023&max_date=12%2F31%2F2024). This file contains the disbursements were submitted by the "Devolder-Santos for Congress" campaign committee (committee id C00721365) and processed by the FEC for the 2021-2022 period.

* A warning occurs when the file is read using `readr::read_csv()`.

```{r}
#| message: false

library(tidyverse)
santos_file <- "schedule_b-2023-10-27T01_16_01.csv" 
santos <- read_csv(santos_file)
```

* The warning is suggesting that we run `problems(santos)`. BTW, `vroom()`?

```{r}
problems(santos)
```

* On row 1148 it appears that a logical (or binary) value was expected in column 50 but the character string "J." was actually found.
* Similarly on row 1193, logical (or binary) values were expected in columns 11, 12, 17 and 18, but character strings were found.
* In all cases, the actual values observed suggest that the columns should have been character, not logical.

```{r}
santos |> select(17, 18, 50)    # Looking at the columns where problems occurred.
```

* Again, character seems like the most likely type for columns named disbursement_type, disbursement_type_description, and payee_middle_name.

***

So what happened?

* By default, readr examines 1000 equally spaced rows (all if less than 1000) to determine the column types: logical?, numeric?, date or date-time?, character (default).
* What if all the values examined (up to 1000) in a column are missing? In this case the column type defaults to logical!
  - Why? In general, logical takes the least memory, character the most.
* We can read all the columns as character and perhaps figure out what happened.

```{r}
santos <- read_csv(santos_file, col_types = cols(.default = col_character()))
table(is.na(santos[[17]]))               # Base R seems easier here.
santos |> summarize(n = n(), .by = 17)   # dplyr is easy enough here.
```

* disbursement_type should be character (or maybe factor), but readr guessed incorrectly because it saw 1000 missing values.

*** 

```{r}
santos |> group_by(is.na(pick(18))) |> summarize(n = n())    # A dplyr version.
santos |> summarize(n = n(), .by = 18)
```

* disbursement_type_description should be character, but readr guessed incorrectly because it saw 1000 missing values.

***

```{r}
table(is.na(santos[[50]]))
table(santos[[50]], useNA = "ifany")                         # A base R version.
```

* payee_middle_name should be character, but readr guessed incorrectly because it saw 1000 missing values.

* `useNA` controls if an `NA` is a level in the resulting table.
  - If "`no`", ignores `NA` values in the table inputs.
  - If "`ifany`", includes `NA` values in the table.
  - If "`always`", adds an extra `NA` value added to each factor level if the `NA` does not already exist.

***

* In this case we could fix these problems by simply letting read_csv() use all 1558 rows to guess the column types.

```{r}
santos <- read_csv(santos_file, guess_max = 2000)
```

This eliminates the warning, but may not solve all problems.

* Some columns still have all missing values (NA). They default to logical type.

```{r}
summary(spec(santos))
```

# Taking Control of the Column Types

According to the R help for `readr::read_csv()`, there are three ways that we can specify the column types when reading a delimited file with readr.

`col_types: One of 'NULL', a 'cols()' specification, or a string. See `
            `'vignette("readr")' for more details.`
            
            If 'NULL', all column types will be inferred from 'guess_max' 
            rows of the input, interspersed throughout the file. This is 
            convenient (and fast), but not robust. If the guessed types 
            are wrong, you'll need to increase 'guess_max' or supply the 
            correct types yourself.
            
            Column specifications created by 'list()' or 'cols()' must 
            contain one column specification for each column. If you only 
            want to read a subset of the columns, use 'cols_only()'.
            
            Alternatively, you can use a compact string representation 
            where each character represents one column:
            
* I think this documentation is out of date, because there is actually a fourth way that the column types can be specified.
  - Here I am counting the default "let readr guess" as a method.
* `readr` now uses the package `vroom` to read delimited (e.g., csv) files. `vroom` is
much faster than the old `readr` code.
* Looking at `?vroom::cols` (instead of `?readr::cols`) shows an additional way
to specify the column types.
* The fourth way is similar to the way that column types are specified in base R using the `colClasses` argument of `read.csv()`.   

# From `?readr::cols`

`Description:`

      'cols()' includes all columns in the input data, guessing the column 
      types as the default. 'cols_only()' includes only the columns you 
      explicitly specify, skipping the rest. In general you can substitute 
      'list()' for 'cols()' without changing the behavior.
      
`Usage:`

      cols(..., .default = col_guess()) 
      
      cols_only(...)
      
# From `?vroom::cols`

| function                    | long name            |short name |
|:----------------------------|:---------------------|:----------|  
|`col_logical()`              |"logical"             |"l"        |
|`col_integer()`              |"integer"             |"i"        |
|` col_big_integer()`         |"big_integer"         |"l"        |
|`col_double()`               |"double", "numeric"   |"d"        |
|`col_character()`            |"character"           |"c"        |
|`col_factor(levels, ordered)`|"factor"              |"f"        |
|`col_date(format = "")`      |"date"                |"D"        |
|`col_time(format = "")`      |"time"                |"t"        |
|`col_datetime(format = "")`  |"datetime", "POSIXct" |"T"        |
|`col_number()`               |"number"              |"n"        |
|`col_skip()`                 |"skip", "NULL"        |"_", "-"   |
|`col_guess()`                |"guess", "NA"         |"?"        |

# Reading Only a Few Columns

What if we just want to read only the columns

* "disbursement_date", "recipient_name", "recipient_zip", and "disbursement_amount";
* and rename them to "date", "recipient", "zip", and "amount", respectively. 

There are several ways to do this, but note first that:

* zip codes are not really numeric (and they may begin with 0);
* "disbursement_date" is entered in the spreadsheet as date-time, although all the times are `00:00:00` (see next slide).
* Aside: Using lubridate functions and `summarize()` to verify that all times in `disbursement_date` are `00:00:00`.

```{r}
santos |>
  mutate(hr = hour(disbursement_date),
         min = minute(disbursement_date),
         sec = second(disbursement_date)) |>
  summarize(n = n(), .by = c(hr, min, sec))
```

Now we demonstrate the 3 ways of specifying the column types.

* With `col_*()` functions. (Here we have piped to `rename()` for renaming.)
* With long type names. (Here we have used the `col_select` arg for renaming.)
* With abbreviated types. (Again we have used `col_select` arg for renaming.)

```{r}
santos <- 
  read_csv(santos_file,
           col_types = cols_only(disbursement_date = col_datetime(),
                                 recipient_name = col_character(),
                                 recipient_zip = col_character(),
                                 disbursement_amount = col_double())
           ) |>
  rename(date = disbursement_date,
         recipient = recipient_name, 
         zip = recipient_zip,
         amount = disbursement_amount)
santos |> slice_head(n = 4)
```

***

```{r}
santos <- 
  read_csv(santos_file,
           col_types = cols_only(disbursement_date = "datetime", 
                                 recipient_name = "character",
                                 recipient_zip = "character",
                                 disbursement_amount = "double"), 
           col_select = c(date = disbursement_date, 
                          recipient = recipient_name, 
                          zip = recipient_zip,
                          amount = disbursement_amount))
santos |> slice_head(n = 5)
```

***

```{r}
santos <- 
  read_csv(santos_file,
           col_select = c(date = disbursement_date, 
                          recipient = recipient_name,
                          zip = recipient_zip,
                          amount = disbursement_amount), 
           col_types = "Tccd")
santos |> slice_head(n = 7)
```

# Coping With All 78 Columns

Three readr Functions

**`spec()`**
`‘spec()’` extracts the full column specification from a tibble created by `readr`.

**`spec_csv()`**
`spec_csv()` generates a column specification. It used internally by `read_csv()` and other readr functions.

* You can run `read_csv()` and then use `spec()` to extract the column spec; 
* Or you can just run `spec_csv()`. Faster if you have a really big file.

**`cols_condense()`**
`'cols_condense()'` takes a spec object and condenses its definition by setting the default column type to the most frequent type and only listing columns with a different type.


***

Working with readr’s Guesses

```{r}
## santos_cols <- spec_csv(santos_file, guess_max = 2000)

santos <- read_csv(santos_file, guess_max = 2000) 
santos_cols <- spec(santos)
summary(santos_cols)
```
***

* The column spec is a named list with class "col_spec".

```{r}
typeof(santos_cols)
attributes(santos_cols)
```

* `santos_cols$delim` seems simple enough.

```{r}
santos_cols$delim
```

* `santos_cols$default` is a `<collector>`.

```{r}
santos_cols$default
```

* `santos_cols$cols` is a named list.
* Its names are column names.

```{r}
typeof(santos_cols$cols)
santos_cols$cols[c(17,18,50)]
santos_cols
```

***

* This can be copied and pasted into a file for editing.
* Instead of copying and pasting, you could do this, but be careful:

```{r}
#| results: hide

sink("santos_cols.txt")
print(santos_cols)
sink()
```

* This would be safer (don’t overwrite an existing file):

```{r}
#| results: hide

file_name <- "santos_cols.txt" 
if (!file.exists(file_name)) { 
  sink(file_name) 
  print(santos_cols)
  sink() 
  }
```

* Or this (include a time stamp in the file name):

```{r}
#| results: hide

file_name <- paste0("santos_cols-", now(), ".txt") 
sink(file_name)
print(santos_cols)
sink()
```

***

* Using `cols_condense()` shortens the list from 78 to 31.
  - The most common column type is identified and all column specifications
of that type are removed.
  - Uses `$default` to represent the type of columns whose type is not explicitly specified, i.e., those removed.
  - In this example, the most common type guessed is character, which is pretty safe as a default type (no information will be lost).
  
```{r}
santos_cols <- cols_condense(santos_cols) 
santos_cols$default
summary(santos_cols)
santos_cols
```

* Note: `cols_condense()` might not do what you want, especially if there are more empty columns (set to logical) than any other type.

<!-- Dr. Presnell have [\textcolor{blue}{proposed}](https://github.com/tidyverse/readr/issues/1478) to the `readr` developers a new version of `cols_condense` that allows the user to specify the `.default` column type which should be "condensed out". By default, the most common column type is still used. -->

<!-- ```{r} -->
<!-- my_cols_condense <- function (x, .default = NULL) { -->
<!--   types <- vapply(x$cols, function(xx) class(xx)[[1]], character(1))  -->
<!--   if (is.null(.default)) { -->
<!--     counts <- table(types) -->
<!--     most_common <- names(counts)[counts == max(counts)][[1]]  -->
<!--     x$default <- x$cols[types == most_common][[1]] -->
<!--     x$cols <- x$cols[types != most_common] -->
<!--     }else{ -->
<!--       x$default <- .default  -->
<!--       x$cols <- x$cols[types != class(.default)[[1]]] -->
<!--       } -->
<!--   x  -->
<!-- } -->
<!-- ``` -->

# Final Result

After examining the column names and doing some other checking, I decided on the following column specification:

```{r}
santos <- read_csv(santos_file,
                   col_types = cols(.default = col_character(),
                                    report_year = col_integer(),
                                    line_number = col_integer(),
                                    file_number = col_integer(), 
                                    disbursement_date = col_datetime(format = ""),
                                    disbursement_amount = col_double(), 
                                    fec_election_year = col_integer(),
                                    load_date = col_datetime(format = ""),
                                    semi_annual_bundled_refund = col_double(),
                                    two_year_transaction_period = col_integer()))
summary(spec(santos))
```

* This is not necessarily "the best" choice, e.g.:
  - We could just change all the integer types to double.
  - We might want to make some of the character columns into factors.
* Of course we can always change the column types after the data have been read into R. This is fine as long as no important information has been lost.