---
title: "Assignment 7"
subtitle: "STA3100"
format:
  html:
    embed-resources: true
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    geometry:
      - margin = 1in
lang: en
execute:
  eval: false
  echo: true
---

## Reading Assignment:

* [r4ds](https://r4ds.had.co.nz/): 11
* [r4ds2e](https://r4ds.hadley.nz/): 17-19

## Problems:

1. The following will use the datasets found in the `nycflights13` R package. Refer to Chapter 19 of [r4ds2e](https://r4ds.hadley.nz/) and the `nycflights13` documentation for details.
    a) Find the 48 hours (over the course of the whole year) that have the worst delays. Cross-reference it with the `weather` data. Can you see any patterns?
    b) Add the latitude and the longitude of the origin and destination airport to `flights`. Is it easier to rename the columns before or after the join?
    c)  Compute the average delay by destination, then join on `airports` and show the spatial distribution of delays. To start, here's an easy way to draw a map of the United States:
    
        ```{r}
          library(ggplot2)
          
          airports |>
            semi_join(flights, join_by(faa == dest)) |>
          ggplot(aes(x = lon, y = lat)) +
            borders("state") +
            geom_point() +
            coord_sf() +
            theme_light() +
            theme(
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank()
            )
        ```

    d) From `flights`, compare `air_time` with the duration between the departure and arrival. Explain your findings. (Hint: consider the location of the airport.)

2.  For each of the following dates, interval and times, show how you'd convert them to a ISO form with a `lubridate` function.
    
    ```{r}
      d1 <- "January 1, 2010"
      d2 <- "2015-Mar-07"
      d3 <- "06-Jun-2017"
      # This is an interval
      d4 <- c("August 19 (2015)", "July 1 (2015)") 
      # Dec 30, 2014
      d5 <- "12/30/14"
      # time 17:05
      t1 <- "1705"
      t2 <- "11:15:10 PM"
    ```
    
3. Write a function that when you input a birthday (as a date), it outputs how old you are in years.

4.  Consider the `islands` vector discussed in class.
    a) Construct a boxplot for these data on a log-scale as well as the original scale.
    b) Construct a dot chart of the island areas. Is a log-transformation needed here?

5. Consider the `pressure` data frame. There are two columns, `temperature` and `pressure`.
    a) Construct a scatterplot with `pressure` on the vertical axis and `temperature` on the horizontal axis. Are the variables related linearly or nonlinearly?
    b)  The graph of the following function passes through the plotted points reasonably well:
        $$
        y=(0.168+0.007x)^{\frac{20}{3}}
        $$
        The differences between the pressure values predicted by the curve and the observed pressure values are called *residuals*. Here is a way to calculate them:
    
        ```{r}
          residuals <- pressure |>
            with(pressure - (0.168 + 0.007*temperature)^(20/3))
        ```
    
        Construct a normal QQ-plot of these residuals and decide whether they are normally distributed or whether they follow a skewed distribution.
    
    c) Now, apply the power transformation $y\to y^{\frac{3}{20}}$ to the pressure data values. Plot these transformed values against temperatures. Is a linear or nonlinear relationship evident now?
    d) Calculated residuals for the difference between the transformed pressure values and those predicted by the straight line $y = 0.168 + 0.007x$. Obtain a normal QQ-plot, and decide whether the residuals follow a normal distribution or not.
