---
title: "`dplyr`: Tibble Operations"
subtitle: "STA3100 Lecture 17"
format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
lang: en
geometry:
  - margin = 1in
  - paperwidth = 11in
  - paperheight = 6.875in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{sectsty}
  - \allsectionsfont{\centering}
  - \pagenumbering{gobble}
execute:
  error: true
---

# Guitars Again
```{r}
#| message: false

mfr <- rep(c("Fender", "Gibson", "Martin"), times = c(2,4,1))
model <- c("Stratocaster", "Telecaster", "Firebird V", "Les Paul Standard", "ES175",
           "J-100 Xtra", "M-21")
serial <- c("EJ15230", "60671", "616619", "00235628", "03531746", "90783019", "1452869")
type <- rep(c("electric", "acoustic"), times = c(5, 2))
body <- rep(c("solid", "hollow", "flat top"), times = c(4, 1, 2))
w_case_kg <- c(8.96, 8.60, 9.88, 8.40, 8.10, 6.78, 5.62)
case_kg <- c(5.70, 4.98, 5.90, 3.46, 4.92, 4.76, 3.72)

library(tidyverse)

guitars <- tibble(mfr, model, serial, type, body, w_case_kg, case_kg)
```

***

* Deleting column of serial numbers to declutter slides.
```{r}
guitars$serial <- NULL
guitars
```

# dplyr One-Table "Verbs" and Base R "Equivalents"

* There are also helper functions to alter/enhance these functions.
* See [\textcolor{blue}{dplyr <-> Base R}](https://cran.r-project.org/web/packages/dplyr/vignettes/base.html) for a more thorough comparison with base R.

|Target     | Purpose                               | dplyr function(s)         |Base R function(s)    |
|:----------|:--------------------------------------|:--------------------------|:---------------------|  
|Columns    |Include or remove a column             |`select()`                 |`[, subset()`         |
|           |Change the names of columns            |`rename()`, `rename_with()`|`names()<-`           |
|           |Change values of / create new columns  |`mutate()`                 |`$<-`, `transform()`  |
|           |Change order of the columns            |`relocate()`               |`[ + union()`         |
|           |Pull out a single variable             |`pull()`                   |`$`, `[[`             |

***

|Target        | Purpose                              | dplyr function(s)   |Base R function(s)                |
|:-------------|:-------------------------------------|:--------------------|:---------------------------------|  
|Rows          |Choose rows based on column values    |`filter()`           |`subset()`                        |
|              |Choose rows based on location         |`slice()`            |`[`                               |
|              |Change order of the rows              |`arrange()`          |`[ + order()`                     |
|              |Remove duplicate rows                 |`distinct()`         |`unique()`, `[ + !duplicated()`   |
|Groups of Rows|Create groupings                      |`group_by()`         |                                  |
|              |Collapse groups into single rows      |`summarise()`        |`aggregate()`, `by()`, `tapply()` |
|Other         |Print a transposed view of the data   |`glimpse()`          |`str()`                           |



# OPERATING ON COLUMNS
# Mutate (Transform) with `dplyr::mutate()`

* Add columns for weights of the guitars in kg and lb.
* Can define `kg` and use it to define `lb` in the same call to `mutate()`.
```{r}
guitars <- mutate(guitars, kg = w_case_kg - case_kg, lb = 2.2 * kg)
guitars
```

# Rename Columns with `dplyr::rename()`

* `new_name = old_name`. Not required to quote new or old column names.
* Can rename more than one column at a time.
```{r}
guitars <- rename(guitars, weight = lb, kilos = kg)
guitars
```

# Select Columns with `dplyr::select()`

* Can use "-" with column numbers **and column names**.
* Not required to quote column names and not require to use `c()`.
```{r}
guitars <- select(guitars, -w_case_kg, -case_kg, -kilos)
## guitars <- select(guitars, -c(w_case_kg, case_kg, kilos))               # Also works.
guitars
```

# Relocate Columns with `dplyr::relocate()`

* Make `type` the first column.
```{r}
guitars <- relocate(guitars, type)
guitars
```

***

What if we now wanted to move the type column back?
```{r}
guitars <- relocate(guitars, type, .after = model)
guitars
```

# OPERATING ON ROWS
# Filter (Subset) Observations with `dplyr::filter()`

* View only guitars under 8 lb

```{r}
filter(guitars, weight < 8)
```

***

* View only electric guitars under 8 lb

```{r}
filter(guitars, (type == "electric") & (weight < 8))
```

***

* Filtering rows and selecting columns are separate operations.
  - Contrast with base R’s `subset()` with its optional argument `select`.

```{r}
guitars |>
    filter((type == "electric") & (weight < 8)) |>
    select(-type, -weight)

## select(filter(guitars, (type == "electric") & (weight < 8)), -type, -weight)
```

# Ordering Observations with `dplyr::arrange()`

* Arrange the guitars by `weight` (lightest to heaviest):

```{r}
arrange(guitars, weight)
```

***

**Decreasing (Descending) Order**

* Arrange the guitars from heaviest to lightest:

```{r}
arrange(guitars, desc(weight))
```

***

**Using Multiple Sort Keys**

* Alphabetically by `mfr` and by `weight` within `mfr`.

```{r}
arrange(guitars, mfr, weight)
```

***

* Alphabetically by `type` and from heaviest to lightest within `type`.

```{r}
arrange(guitars, type, desc(weight))
```

# Subset Rows by Position with `dplyr::slice()`

The main advantage of these functions over the base R analogues is that they interoperate well with other `dplyr` functions.

* Get the 2nd, 4th, and 5th Guitars.

```{r}
slice(guitars, 2, 4, 5)     # Or slice(guitars, c(2,4,5))
```

***

* Get the first 3 guitars.

```{r}
slice_head(guitars, n = 3)     # Default is n = 1.
```

***

* Get the last 3 guitars.

```{r}
slice_tail(guitars, n = 3)
```

***

* Get the 3 heaviest guitars.

```{r}
slice_max(guitars, weight, n = 3)
```

* A base R version could be any of the following:

```r
head(guitars[order(guitars$weight, decreasing = TRUE), ], n = 3)
guitars[order(guitars$weight, decreasing = TRUE), ][1:3,]
guitars[order(guitars$weight, decreasing = TRUE)[1:3], ]
```

***

* Choose a random sample (without replacement) of 3 guitars.

```{r}
slice_sample(guitars, n = 3)
```

* A base R version:

```r
guitars[sample(nrow(guitars), 3), ]
```

# GROUPED CALCULATIONS
# Group and Summarize with `dplyr::group_by()` and `dplyr::summarise()`

**Calculate the Mean Weight by Type**
```{r}
guitars |>
    group_by(type) |>
    summarise(avg_wt = mean(weight)) |>
    arrange(desc(type))
```

* Note that a tibble is returned (ready for more pipes)

***

**An alternative (as of dplyr 1.1.0):**
```{r}
summarise(guitars, avg_wt = mean(weight), .by = type)
```

***

**Include the Number of Observations in Each Group**
```{r}
guitars |>
    group_by(type) |>
    summarise(n = n(), avg_wt = mean(weight))
```

***

**An alternative (as of dplyr 1.1.0):**
```{r}
summarise(guitars, n = n(), avg_wt = mean(weight), .by = type)
```

***

**Group by multiple variables.**
```{r}
guitars |>
    group_by(type, body) |>
    summarise(n = n(), avg_wt = mean(weight))
```

***

* Notice that the result of the previous command was still grouped by `type`:

```{r}
guitars |>
    group_by(type, body) |>
    summarise(n = n(), avg_wt = mean(weight)) |>
    group_vars()
```

***

**We can fix it by `.groups`**
```{r}
guitars |>
     group_by(type, body) |>
     summarise(n = n(), avg_wt = mean(weight), .groups = "keep")
```

***

* Now check the group variables
```{r}
guitars |>
    group_by(type, body) |>
    summarise(n = n(), avg_wt = mean(weight), .groups = "keep") |>
    group_vars()
```

***

**An alternative (as of dplyr 1.1.0):**
```{r}
guitars |>
    summarise(n = n(), avg_wt = mean(weight), .by = c(type, body))
```

* There is a small difference though: the result is not grouped.

```{r}
guitars |>
    summarise(n = n(), avg_wt = mean(weight), .by = c(type, body)) |>
    group_vars()
```

# Revisiting the Titanic

**Converting a Table to a Tibble**
```{r}
is.array(Titanic)
dim(Titanic)
class(Titanic)
dimnames(Titanic)
```

***

```{r}
Titanic_tbl <- as_tibble(Titanic)
Titanic_tbl
```

***

```{r}
Titanic_tbl |> summarize(n = sum(n))
Titanic_tbl |> group_by(Survived) |> summarize(n = sum(n))
```

***

```{r}
Titanic_tbl |> group_by(Survived) |>
    summarize(n = sum(n)) |>
    mutate(prop = n/sum(n))
```

* Why does this work?
  - Because `summarize()` peels off the last and only level of grouping before `mutate()` is called. Thus the `sum(n)` in `mutate()` is ungrouped.
```{r}
Titanic_tbl |> group_by(Survived) |>
    summarize(n = sum(n)) |>
    group_vars()
```

***

```{r}
Titanic_tbl |>
    group_by(Class, Survived) |>
    summarize(n = sum(n))
```

***

```{r}
Titanic_tbl |> group_by(Class, Survived) |>
    summarize(n = sum(n)) |>
    mutate(prop = n/sum(n))
```

* This works because `summarize()` has peeled off the last level of grouping (“Survived”), leaving only grouping by "Class" (see next slide). Thus the `sum(n)` in `mutate` is grouped by “Class”.

***

* Showing the grouping (by “Class”) still remaining after the call to `summarize()`.

```{r}
Titanic_tbl |> group_by(Class, Survived) |>
    summarize(n = sum(n)) |>
    group_vars()
```

* Using `summarize()` with the .by optional argument (new in dplyr 1.1.0) only uses transient grouping. For this reason, the last example does not work in the same way if we use .by instead of `group_by()` (see next slide).

***

* This is not what we want:

```{r}
Titanic_tbl |>
    summarize(n = sum(n), .by = c(Class, Survived)) |>
    mutate(prop = n/sum(n))
```

***

* Here’s how to make the `.by` version give the `group_by()` result.
```{r}
Titanic_tbl |>
    summarize(n = sum(n), .by = c(Class, Survived)) |>
    mutate(prop = n/sum(n), .by = Class)
```
  - Is this easier to understand than the `group_by()` approach?

***

* And how to make the `group_by()` version give the first (unwanted) `.by` result.
  - Just use `ungroup()` to remove all grouping before the `mutate()` step.
```{r}
Titanic_tbl |> group_by(Class, Survived) |>
    summarize(n = sum(n)) |>
    ungroup() |>
    mutate(prop = n/sum(n))
```  

# Pivoting
Here’s a way to make the data frame look like a table.
```{r}
Titanic_tbl |>
    group_by(Class, Survived) |>
    summarize(n = sum(n)) |>
    mutate(prop = n/sum(n)) |>
    pivot_wider(names_from = Survived, values_from = c(n, prop))
```  

***

And a variation, pivoting and then adding a single column of proportions.
```{r}
Titanic_tbl |>
    summarize(n = sum(n), .by = c(Class, Survived)) |>
    pivot_wider(names_from = Survived, names_prefix = "Surv_", values_from = n) |>
    mutate(prop_Yes = Surv_Yes / (Surv_Yes + Surv_No))
``` 


# Example: Run Length Encoding a Sequence of Coin Tosses
  Question: Suppose we toss a fair coin 100 times. What is the probability of seeing at least one run of 6 or more consective heads or tails?
```{r}
x <- sample(c("H","T"), 100, replace = TRUE) 
noquote(x)
xr <- rle(x)
xr
```
There is no `as_tibble()` method for class rle, but this works:
```{r}
library(tidyverse)
xr_df <- as_tibble(unclass(xr))
xr_df |> summarize(max_run = max(lengths)) # Length of longest run overall.

xr_df |> summarize(max_run = max(lengths), .by = values) # Longest runs of H and T.
```
Simulation
```{r}
nsim <- 10000; ntosses <- 100
maxrl <- vector("integer", length = nsim) 
for (i in 1:nsim) {
  x <- sample(c("H","T"), ntosses, replace = TRUE) 
  xr <- rle(x)
  maxrl[i] <- as_tibble(unclass(xr)) |>
    group_by(values) |>
    summarise(max_run = max(lengths)) |>
    filter(values == "H") |>
    pull(max_run)
  #maxrl[i] <- max(xr$lengths)
}
```
Now we plot the maximum run length 
```{r}
df <- tibble(maxrl) |>
  count(maxrl) |>     # similar to summarize(n = n(), .by = maxrl)
  mutate(prop = n / sum(n))
slice_head(df, n = 10)
```

* Using a hint from the help for `geom_segment()` to get a `type = "h"` plot with ggplot2.
* Could use `+ theme(plot.margin = ...)` to change plot margins, but default ggplot2 margins are ok.
```{r}
ggplot(df, aes(x = maxrl, y = prop)) +
  geom_segment(aes(yend = 0, xend = maxrl), linewidth = 2) + 
  labs(title = "Estimated distribution of maximum run length.",
       x = "Maximum Run Length", 
       y = "Relative Frequency")
```
