---
title: "`tidyverse`: Introduction"
subtitle: "Lecture 13"
format:
  html:
    embed-resources: true
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    geometry:
      - margin = 1in
      - paperwidth = 11in
      - paperheight = 6.875in
    header-includes:
      - \usepackage{booktabs}
      - \usepackage{sectsty}
      - \allsectionsfont{\centering}
      - \pagenumbering{gobble}
lang: en
execute:
  error: true
---

# What Is the `tidyverse`?

* The `tidyverse` is a collection of R packages built around a common design philosophy for data analysis, redefining how many common operations are performed in R.
* It emphasizes:
  1. **Tidy data**
  2. **Consistent** function interfaces
  3. **Readable** code
  4. Tools that **work well together**

```{r}
#| message: false

	library(tidyverse)
```

# The Overarching Design Ethos

* **Tibble-in, tibble-out**: Most `tidyverse` functions take a data frame (`tibble`) as the first argument and likewise return a data frame (`tibble`) as output.
* Functions are especially **pipe-friendly**
* Function names are **descriptive** and **consistent** across packages:
  - Ex: `filter()`, `select()`, `mutate()`, `pivot_longer()`, `str_detect()`

* The `tidyverse` prefers:
  - **explicit** code over hidden side effects
  - **rectangular/tabular** data structures
  - every function has **one main job**
  - **consistent** names and argument conventions
* Compared with base-R, `tidyverse` code is often:
  - more **verbose**
  - more **readable**
  - easier to compose with **pipes**

# Tidy Data

* The `tidyverse` is strongly connected to the idea of _tidy data_.
* A common version of the tidy-data principle is:
  1. Each **variable** is a column
  2. Each **observation** is a row
  3. Each **cell** contains one value
* Tidy data are easier to:
  - **filter**
  - **summarize**
  - **reshape**
  - **visualize**

# Why Tibbles?

* A **tibble** is the **main data-frame structure** used throughout the `tidyverse`.
* A tibble is still a data frame, but it has different default behavior.
* Tibbles are a little **stricter** and a little **safer** than base-R data frames.
* Many `tidyverse` functions will:
  - expect a `tibble`
  - return a `tibble`

```{r}
	tibble(x = 1:3, y = c("a", "b", "c"))
```

# Core Packages

* **`tibble`**: modern data-frame structure used throughout the `tidyverse`
* **`dplyr`**: selecting, filtering, transforming, joining, and summarizing rows/columns
* **`tidyr`**: reshaping data between wide and long forms
* **`readr`**: reading and writing flat files such as csv and tsv
* **`ggplot2`**: plotting and statistical graphics

***

* **`stringr`**: character strings and regular expressions
* **`forcats`**: factors and factor levels
* **`lubridate`**: dates and date-times
* **`purrr`**: functional programming tools for iteration and list-processing

# How These Pieces Fit Together

* A common `tidyverse` workflow in this course will look like:
  1. **Import** data with `readr`
  2. **Store** or convert to a `tibble`
  3. **Transform** with `dplyr`
  4. **Reshape** with `tidyr`
  5. Handle strings/factors/dates with `stringr`, `forcats`, and `lubridate`
  6. **Visualize** with `ggplot2`

```{r}

	example_tbl <-
	tibble(
		group = c("A", "A", "B", "B"),
		score = c(10, 12, 8, 15)
		) |>
			summarize(avg_score = mean(score), .by = group)

	example_tbl
```

# A Minimal Example

```{r}

	tibble(
		fruit = c("apple", "banana", "apple", "pear"),
		weight = c(3, 2, 4, 5)
		) |>
			summarize(
		n = n(),
		avg_weight = mean(weight),
		.by = fruit
		)
```

* This is typical `tidyverse` code:
  - data in a **`tibble`**
  - a clear **verb** (`summarize()`)
  - grouping via **`.by`**
  - output returned as another **`tibble`**
