---
title: "Data Frames"
subtitle: "STA3100 Lecture 9"
format:
  html:
    embed-resources: true
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
lang: en
geometry:
  - margin = 1in
  - paperwidth = 11in
  - paperheight = 6.875in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{sectsty}
  - \allsectionsfont{\centering}
  - \pagenumbering{gobble}
execute:
  error: true
---

\newpage

# Data Frames
Data frames are usually created by reading data from a file, such as a spreadsheet, but they can be created manually using `data.frame()`.  
```{r}
mfr <- rep(c("Fender", "Gibson", "Martin"), times = c(2,4,1))

model <- c("Stratocaster", "Telecaster", "Firebird V", "Les Paul Standard", "ES175",
           "J-100 Xtra", "M-21")

serial <- c("EJ15230", "60671", "616619", "00235628", "03531746", "90783019", "1452869")

type <- rep(c("electric", "acoustic"), times = c(5, 2))

body <- rep(c("solid", "hollow", "flat top"), times = c(4, 1, 2))

w_case_kg <- c(8.96, 8.60, 9.88, 8.40, 8.10, 6.78, 5.62)

case_kg <- c(5.70, 4.98, 5.90, 3.46, 4.92, 4.76, 3.72)
```

\newpage

```{r}
  guitars <- data.frame(mfr, model, serial, type, body, w_case_kg, case_kg)
  
  guitars
```

* Notice the column names: data frames always have column names (unlike matrices).
* Data frames also always have row names, usually just the row numbers.

\newpage

```{r}
summary(guitars)
```

\newpage

# A Closer Look
```{r}
  str(guitars) 
```

```{r}
  attributes(guitars)
```

# Indexing Data Frames
**A data frame is NOT a matrix ...**
```{r}
  is.matrix(guitars)
```
... and while there is no explicit `dim` attribute (previous slide), we still get ...
```{r}
  dim(guitars)
```
... and you can ...

\newpage

**Index a Data Frame Like a Matrix**
```{r}
  guitars[2, 6]
  # Or guitars[2, "w_case_kg"]
  guitars[2, ]
  guitars[, 6]
```

```{r}
  guitars[2:3, ]
  guitars[, c(2, 6, 7)]
```

* Note however that `guitars[2, 6]` and `guitars[, 6]` returned vectors
```{r}
  str(guitars[2, 6])
  str(guitars[, 6])
```
while the other examples returned a data frame.  
This is the default behavior for data frames if indexing returns a single column (or part of a single column).

* You can force `guitars[2, 6]` and `guitars[, 6]` to return a (single-column) data frame with `drop = TRUE`.
```{r}
  guitars[2, 6, drop = FALSE]
  # Returns a data frame with 1 row and 1 column.
  guitars[, 6, drop = FALSE]
  # Returns a data frame with 1 column.
```

\newpage

* Soon we will learn about "tibbles" (part of the "tidyverse").  
  - Tibbles are dataframes too.
  - But they are designed to have more consistent and predictable default behaviors, e.g.,
    + `drop` defaults to `FALSE` for tibbles; and
    + row numbers (names) are handled more naturally.

* You can also assign row names, but this is usually discouraged:
```{r}
  guitars2 <- guitars
  rownames(guitars2) <- guitars2$serial
  guitars2$serial <- NULL
  
  rownames(guitars2)
  colnames(guitars2)
```    

```{r}
  dimnames(guitars2)
  guitars2["EJ15230", c("mfr", "model", "type")]
```  

```{r}
  guitars2
  str(guitars2)
```

\newpage

# Surprise: A Data Frame is a List (of It’s Columns)
```{r}
  is.list(guitars)
  names(guitars)
```
So you can also index a data frame like a list (next slide).

***

```{r}
  guitars[5]      # Or guitars["w_case_kg"]
  guitars[[5]]    # Or guitars[["w_case_kg"]]
  guitars$w_case_kg
```

***

Again, note the difference:
```{r}
  str(guitars[5])
  str(guitars[[5]])
```

\newpage

# Two Ways of Getting the Same Thing
```{r}
  guitars[c(1,5)]
  guitars[, c(1,5)]
```
Q. What would happen if you did this with a matrix?

***
```{r}
  guitars[[c(1,5)]]
  # Same as guitars[[1]][[5]]. Returns 5th element of 1st column.
```

Q. What happens if you try this with a matrix? _It’s an error._   
Q. What happens if you try this with an ordinary list (a list that isn’t a dataframe)? _Same as ‘lst[[1]][[5]]._  

\newpage

# `stringsAsFactors`

* In versions of R prior to 4.0, `data.frame()` converted all `character` vectors to `factor` by default (we will cover factors later in the semester).
* This behavior is controlled by the `stringsAsFactors` argument.
* The default changed to `stringsAsFactors = FALSE` in R 4.0 (April 24, 2020).
* Now it is up to you to convert character vectors to factors if you need to, e.g.,
```{r}
  guitars$type <- factor(guitars$type)
  guitars$body <- factor(guitars$body)
  ## guitars$mfr <- factor(guitars$mfr) # Might do this too?
```

* If you want the old behavior (doubtful), set `stringsAsFactors = FALSE`.
* If you want to know more about the history of this default see:
  - [\textcolor{blue}{stringsAsFactors: An unauthorized biography}](https://simplystatistics.org/posts/2015-07-24-stringsasfactors-an-unauthorized-biography/)
  - [\textcolor{blue}{stringsAsFactors = <sigh>}](https://notstatschat.tumblr.com/post/124987394001/stringsasfactors-sigh)
