---
title: "More Strings"
subtitle: "STA3100 Lecture 31"
format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
lang: en
geometry:
  - margin = 1in
  - paperwidth = 11in
  - paperheight = 6.875in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{sectsty}
  - \allsectionsfont{\centering}
  - \pagenumbering{gobble}
execute: 
  error: true
---

# A Meta-stringr Example
**Q: What `stringr` Functions Have We Used So Far? And how often?**  

The file `stringr_lines.txt` was obtained by grepping the R markdown files used to produce the slides used in this class so far. The shell command used (which is run outside R) was:

```{r}
#| eval: false

grep -n "str_" ../lecture_([0-2][0-9]|30)/*.qmd > stringr_lines.txt
```

* Each line of `stringr_lines.txt` contains:
  - the name of an R markdown file,
  - followed by a colon,
  - followed by a line number,
  - followed by another colon,
  - followed by a line from the file containing the string “str_”.

# Reading Lines From the File  

* Using `read_lines()` from the readr package to read in these data.
  - The base R function `readLines()` would also work here.
  
```{r}
#| message: false

library(tidyverse)
strlines <- read_lines("stringr_lines.txt") 
length(strlines)
head(strlines, n = 5)
```  

# Extracting a Vector of the Function Names

Let’s extract all the mentions of functions starting with `str_`.

```{r}
funs <- str_extract_all(strlines, "str_([a-z]+)(_[a-z]+)*\\(") 
funs[1:4]
```  

* `flatten_chr()` from the purrr package converts list to a character vector.
  - Flattens the list by combining vectors in the list.
  - Equivalent here to base R `do.call(c, funs)`
  - As if we ran `c(funs[[1]], funs[[2]], ..., funs[[205]])`

```{r}
funs <- flatten_chr(funs) 
funs[1:5]
```  

* Now strip off the trailing left paren with `str_sub()`.

```{r}
funs <- str_sub(funs, 1, -2) 
funs[1:5]
```  

* So what stringr function have we used so far?

```{r}
unique(funs)
```  

* And how often have we used them?

```{r}
sort(table(funs))
```  

Aside: We haven’t used them all (yet).

```{r}
apropos("^str_", mode = "function")    # Anywhere in our search path, but only functions
ls()                    # Lists everything in global environment (interactive workspace) 
search()                        # Notice that our search path includes `package:stringr` 
ls("package:stringr", pattern = "^str_")     # Within "package:stringr" matching pattern
```  

# Use the Lines to Create a Data Frame

A data frame with columns for the

* file names
* line numbers
* function names

would be more a useful data structure.

* Start by creating a data frame with a single column containing the lines read from the file.

```{r}
df <- tibble(string = strlines) 
slice_head(df, n = 4)
```  

The dplyr function `separate_wider_delim()` is very useful for the next step.

`separate_wider_delim()` is similar to `str_split()` but:

* The first argument is a data frame (no need to call `mutate()`)
* User specifies the column(s) to be separated (`cols` argument)
* User specifies the string delimiter between values (`delim` argument)
* User specifies column names for new columns (`names` argument)
* Always creates same number of pieces from each string
  - Number of pieces is equal to number of non-`NA` elements of `names`
  - `too_many` argument says what to do with any extra pieces
  - `too_few` says what to do if too few pieces.
* See also `separate_wider_position()` and `separate_wider_regex()`.  

```{r}
df <- df|> 
  separate_wider_delim(string, ":", 
                       names = c("file", "lnum", "line"), 
                       too_many = "merge")
slice_head(df, n = 4)
``` 

Note that:

* The `string` column has been removed (by default `cols_remove = TRUE`).
* The `lnum` column is `chr`, not `int` or `dbl`.

Now we can mutate the `line` column with `str_extract_all()`.

* Because the function returns a list, this produces a _list column_!

```{r}
df <- df|> 
  mutate(
    lnum = as.integer(lnum), # Might as well do this conversion now.
    fun = str_extract_all(line, "str_([a-z]+)(_[a-z]+)*\\(") 
  )
slice_head(df, n = 4) |> select(-line)
``` 

* What does this list column look like?

```{r}
df |> pull(fun) |> head(4)
```

* We don’t need the original `line` column anymore.

```{r}
df <- select(df, -line)
```

The tidyr function `unnest_longer()` will

* convert the list column into a vector column
* in a longer version of the data frame (similar to `pivot_longer()`).

```{r}
df_old <- df          # Saving current version for later comparison 
df <- unnest_longer(df, fun)
df |> slice_head(n = 5)
```

* Can you see what has happened?

Compare:

```{r}
df_old |>
  filter(str_detect(file, "lecture_30.qmd"), lnum == 71)
df |> 
  filter(str_detect(file, "lecture_30.qmd"), lnum == 71)
```

* The `fun` column needs a little cleaning, just like before.

```{r}
df <- df|> 
  mutate(fun = str_sub(fun, 1, -2))
head(df)
```

# Using the Data Frame


```{r}
df |> 
  summarize(n = n(), .by = fun) |> 
  arrange(desc(n)) |>
  print(n = Inf)

df |> 
  group_by(file) |> 
  summarize(n = n()) |> 
  arrange(desc(n))

df |> filter(str_detect(file, "lecture_19.qmd"))
```

# Repetition, Back-references, and Looking Around

**Some parenthetic comments**    
Parentheses play two roles in a regex: grouping and creating back-references.

**Question**  
What happens to the back-references if a group is repeated with `?` or `+` or `*` or `{m,n}` or `{m}` or `{m,}`?

**Answer**  
The back-reference refers to the last of the repeated matches.

**Secondary Answer**  
If you want to group for repetition without creating a back-reference, you can use a non-capturing group: (`?: ...` )

For more information see [_\textcolor{blue}{Grouping and Capturing}_](https://www.regular-expressions.info/brackets.html) and [_\textcolor{blue}{Backreferences}_](https://www.regular-expressions.info/backref.html) in this [\textcolor{blue}{tutorial}](https://www.regular-expressions.info/tutorial.html).

***

**An Example**  
Let’s look at an example related to our meta-readr case study.

```{r}
str_match("str_abc_def_ghi(x,y,z)", "str_([a-z]+)(_[a-z]+)*\\(")
```

This shows that:

1. The overall match was: "`str_abc_def_ghi(`"
2. The non-repeating 1st group (`[a-z]+`) matched: "`abc`"
3. The _last match_ of the repeating group (`[a-z]+`)* was: "`_ghi`"

* The first match of the repeating group would have been: "`_def`"
* Change repeating group to non-capturing and enclose in a capturing group

```{r}
str_match("str_abc_def_ghi(x,y,z)", "str_([a-z]+)((?:_[a-z]+)*)\\(")
```

* Or just grab everything after "`str_`" and before "`\\(`"

```{r}
x <- str_replace("str_abc_def_ghi(x,y,z)", "str_((?:[a-z]+)(?:_[a-z]+)*)(\\(.*\\))", "\\1") 
x
str_split(x, "_")        # Possible next step
```

**Looking Ahead**

Our earlier regex could be modified to check for the trailing left-parenthesis without returning it as part of the match.

This can be done using a _positive lookahead_: (`?=...`)

A positive lookahead causes the regex engine to look ahead for a match without
retaining it as part of the overall match.

See [\textcolor{blue}{Lookahead and Lookbehind}](https://www.regular-expressions.info/lookaround.html) for more on lookarounds.

So in the following example:

* in the first string, the `(` after `str_match_all` is necessary for the regex to match, but it is not returned as part of the match
* the second string is not matched by the regex because there is a space and not a ( immediately after `str_match_all`.

```{r}
strngs <- 
  c("str_match_all(x,y,z)",
    "str_match_all and grepl() are functions") 
str_extract(strngs, "str_([a-z]+)(_[a-z]+)*(?=\\()")
```