---
title: "Tidyverse: Tibbles"
subtitle: "STA3100 Lecture 15"
format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
lang: en
geometry:
  - margin = 1in
  - paperwidth = 11in
  - paperheight = 6.875in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{sectsty}
  - \allsectionsfont{\centering}
  - \pagenumbering{gobble}
execute:
  error: true
---

# What are Tidy Data?
* Ideas/principles for how data should be organized for statistical analysis.
* Not necessarily new, but codified by Wickham (2014).
  - Motivated by work on `reshape` package
    + NB: `reshape` (and `reshape2`) have been superceeded by `tidyr`
* A basic principle is that data should be arranged so that  
  – Each variable is a column  
  – Each observation is a row

# So What is the [\textcolor{blue}{Tidyverse}](https://www.tidyverse.org/)?
* A collection of R packages and functions/methods for  
  – importing data into R  
  – tidying, transforming, and otherwise manipulating data in R   
  – plotting and graphing data ('ggplot2')
* In most cases these functions are intended to improve on existing functionality in base R while being  
  – More consistent and predictable  
  – Easier to use
* The tidyverse has sometimes felt like a moving target as it evolved and the developers revised previous design decisions. It is much more settled now, although changes are still made from time to time.

# Tibbles

The `tibble` package is loaded by the `tidyverse` meta-package.
```{r}
#| echo: true
library(tibble)   # Not needed if you have run library("tidyverse") instead

mfr <- rep(c("Fender", "Gibson", "Martin"), times = c(2,4,1))
model <- c("Stratocaster", "Telecaster", "Firebird V", "Les Paul Standard", "ES175",
           "J-100 Xtra", "M-21")
serial <- c("EJ15230", "60671", "616619", "00235628", "03531746", "90783019", "1452869")
type <- rep(c("electric", "acoustic"), times = c(5, 2))
body <- rep(c("solid", "hollow", "flat top"), times = c(4, 1, 2))
w_case_kg <- c(8.96, 8.60, 9.88, 8.40, 8.10, 6.78, 5.62)
case_kg <- c(5.70, 4.98, 5.90, 3.46, 4.92, 4.76, 3.72)

guitars_df <- data.frame(mfr, model, serial, type, body, w_case_kg, case_kg)
guitars_tbl <- tibble(mfr, model, serial, type, body, w_case_kg, case_kg)
```
***

```{r}
#| echo: true
guitars_df
```
***
```{r}
#| echo: true
guitars_tbl
```
***
```{r}
#| echo: true

attributes(guitars_tbl)
```
***
```{r}
#| echo: true

class(guitars_tbl)
```

* `tbl` indicates a generic tibble (for S3 object system purposes)
* `tbl_df` is a subclass of `tbl`
  - `tbl_df` indicates that the tibble is also a data frame
  - there is also a `tbl_dbi` class for database (SQL) connections

We will see another subclass: 
  
  * `spec_tbl_df` is a `tbl_df` with a spec attribute (as described in [\textcolor{blue}{tibble subclass}](https://www.tidyverse.org/blog/2018/12/readr-1-3-1/#tibble-subclass))

You do not normally need to worry about any of this.
    
# So What’s a Tibble Then?   
From the `tbl_df-class` help page:

Description:

        The ‘tbl_df’ class is a subclass of ‘data.frame’, created in order to
        have different default behaviour. The colloquial term "tibble" refers
        to a data frame that has the ‘tbl_df’ class. Tibble is the central data
        structure for the set of packages known as the tidyverse, including
        dplyr, ggplot2, tidyr, and readr.
    
        The general ethos is that tibbles are lazy and surly: they do less and
        complain more than base data.frames. This forces problems to be tackled
        earlier and more explicitly, typically leading to code that is more
        expressive and robust.

***

Properties:

      Objects of class ‘tbl_df’ have:
          • A ‘class’ attribute of ‘c("tbl_df", "tbl", "data.frame")’.
          • A base type of ‘"list"’, where each element of the list has
           the same ‘vctrs::vec_size()’.
          • A ‘names’ attribute that is a character vector the same
           length as the underlying list.
          • A ‘row.names’ attribute, included for compatibility with
           data.frame. This attribute is only consulted to query the
           number of rows, any row names that might be stored there are
           ignored by most tibble methods.
           
***         

Behavior:

      How default behaviour of tibbles differs from that of data.frames,
      during creation and access:
      
         • Column data is not coerced. A character vector is not turned
           into a factor. List-columns are expressly anticipated and do
           not require special tricks. Internal names are never stripped
           from column data. Read more in ‘tibble()’.
         • Recycling only happens for a length 1 input. Read more in
           ‘vctrs::vec_recycle()’.
         • Column names are not munged, although missing names are
           auto-populated. Empty and duplicated column names are
           strongly discouraged, but the user must indicate how to
           resolve. Read more in ‘vctrs::vec_as_names()’.

  ***
  
         • Row names are not added and are strongly discouraged, in
           favor of storing that info as a column. Read about in
           rownames.
         • ‘df[, j]’ returns a tibble; it does not automatically extract
           the column inside. ‘df[, j, drop = FALSE]’ is the default.
           Read more in subsetting.
         • There is no partial matching when ‘$’ is used to index by
           name. ‘df$name’ for a nonexistent name generates a warning.
           Read more in subsetting.
           
      See ‘vignette("invariants")’ for a detailed description of the
      behavior.
      
      Furthermore, printing and inspection are a very high priority. The goal
      is to convey as much information as possible, in a concise way, even
      for large and complex tibbles. Read more in formatting.         

***

**For a tibble, drop = FALSE by default:**
```{r}
#| echo: true
guitars_df[, 5]
guitars_tbl[, 5]
```
***

Even in this case:
```{r}
#| echo: true
guitars_df[1, 5]
guitars_tbl[1, 5]
```
***

**Data frames (and lists) allow partial matching of names when using $:**
```{r}
#| echo: true
names(guitars_df)
guitars_df$mo
```

Although not with `[[`:
```{r}
#| echo: true
guitars_df[["mo"]]
```
***

**Tibbles do not:**
```{r}
#| echo: true
names(guitars_tbl)
guitars_tbl$mo
```

# `vignette("tibble")`
* [\textcolor{blue}{Tibbles}](https://cran.r-project.org/web/packages/tibble/vignettes/tibble.html)

***

**List Columns Are Easy in Tibbles**
```{r}
#| echo: true
tibble(x = 1:3, y = list(1:5, 1:10, 1:20))
```

And not so easy with ordinary data frames:
```{r}
#| echo: true
data.frame(x = 1:3, y = list(1:5, 1:10, 1:20))
```

***
Another example:
```{r}
#| echo: true
tibble(x = 1:2, y = list(num = 1:2, ltr = letters[1:4]))
data.frame(x = 1:2, y = list(num = 1:2, ltr = letters[1:4]))
```
***

So this what `data.frame()` is doing (and was trying to do in the first example):
```{r}
#| echo: true
data.frame(x = 1:2, y.num = 1:2, y.ltr = letters[1:4])
```

Tibbles are lazy and surly (in a good way?):
```{r}
#| echo: true
tibble(x = 1:2, y.num = 1:2, y.ltr = letters[1:4])
```

***

Finally, here are two ways to get a list column in an ordinary data frame.

* Protect the list with `I()`:

```{r}
#| echo: true
data.frame(x = 1:2, y = I(list(num = 1:2, ltr = letters[1:4])))
```

* Add the list column to an existing data frame:

```{r}
#| echo: true
df <- data.frame(x = 1:2)
df$y <- list(num = 1:2, ltr = letters[1:4])
df
```

***

**What’s this about “munging” names?**

From [\textcolor{blue}{advr}](https://adv-r.hadley.nz/):

\setlength{\leftskip}{0.6cm}

R has strict rules about what constitutes a valid name. A syntactic name must consist of letters, digits, . and \_, but can’t begin with \_ or a digit. Additionally, you can’t use any of the reserved words like `TRUE`, `NULL`, `if`, and `function` (see the complete list in ?Reserved). A name that doesn’t follow these rules is a non-syntactic name; if you try to use them, you’ll get an error.
  
\setlength{\leftskip}{0cm}

***      
You can use non-syntactic names (not recommended), but you must quote them with backticks:

```{r}
#| echo: true
_x <- 1:5
```
```{r}
#| echo: true
`_x` <- 1:5
`_x`
```

***

```{r}
#| echo: true
x <- data.frame(`crazy name` = 1)
x

y <- tibble(`crazy name` = 1)
y

y$`crazy name`
```

***

**Notice this difference between tibble() and as_tibble()**
```{r}
#| echo: true
lst <- list(l = letters[1:3], n = 1:3)
tibble(lst)

as_tibble(lst)
```

# References
Wickham, Hadley. 2014. “Tidy Data.” _Journal of Statistical Software_ 59 (1): 1–23. [\textcolor{blue}{https://doi.org/10.18637/jss.v059.i10}](https://doi.org/10.18637/jss.v059.i10).
