---
title: "Random Numbers and Simulation"
subtitle: "STA3100 Lecture 34"
format:
  html:
    embed-resources: true
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    geometry:
      - margin = 1in
      - paperwidth = 11in
      - paperheight = 6.875in
    header-includes:
      - \usepackage{booktabs}
      - \usepackage{sectsty}
      - \allsectionsfont{\centering}
      - \pagenumbering{gobble}
lang: en
execute:
  error: true
---

# Pseudo-Random Number Generation

"Any one who considers arithmetical methods of producing random digits is, of course, in a state of sin. For, as has been pointed out several times, there is no such thing as a random number– there are only methods to produce random numbers, and a strict arithmetic procedure of course is not such a method." - John von Neumann (1951)

[Random Numbers Fall Mainly in the Planes](Random Numbers Fall Mainly in the Planes - George Marsaglia (1968)) - George Marsaglia (1968)

* A sequence of pseudo-random numbers is usually produced by altering the digits in an integer via arithmetic operations and various operations to shift, flip, or shuffling the bits representing the integer.
* Hope to produce a sequence of integers which are practically indistinguishable from a sequence of _independent_ random integers uniformly distributed over the range, e.g., 0 to $2^{32-1} - 1$ or 0 to $2^{64-1} - 1$.
* These pseudo-random integers are then used to produce psuedo-independent pseudo-random numbers from various distributions.

# Random Sampling in R
```r
sample()
```
* Draw (pseudo-)random samples from a finite collection of objects.
* Can sample with or without replacement (without replacement is the default).
* Can sample with equal probabilities or with any specified set of `probability weights' (probabilities specified up to a constant of proportionality).


**Example: Roll Two Fair Dice**
```{r}
	sample(1:6, 2, replace = TRUE)    # Sampling with replacement
```
Equivalent to
```{r}
	sample.int(6, size = 2, replace = TRUE)
```
***

**Roll a Loaded Die 36 Times**
```{r}
	p <- c(1, rep(2, 4), 3) # Probability weights 
	p

	p / sum(p)              # The actual probabilities

	## p<-p/sum(p)
	x <- sample(1:6, 36, replace = TRUE, prob = p) 
	x
```

The results are random, but we generally expect to see more 6’s than 1’s.
```{r}
	sum(x == 1)

	sum(x == 6)

	table(x)
```

Let’s estimate the probability that the number of 6’s exceeds the number of 1’s.
```{r}
	x <- replicate(10000, sample(1:6, 36, replace = TRUE, prob = p))   # Repeat 10,000 times 
	dim(x)      # Each column is a replicate (result of 36 rolls)

	n1 <- apply(x == 1, 2, sum) 
	n6 <- apply(x == 6, 2, sum) 
	mean(n6 > n1)
```

Question: Can you calculate the exact probability?

<!-- sum(dbinom(1:36, 36, 0.25) * pbinom(0:35, 36, 1/12)) -->

Question: How accurate is our _estimate_ of the probability?

***

**Draw a five card poker hand**

`outer()` is a useful function.
```{r}
	outer(1:3, 5:6) # Default is `FUN = "*"`

	outer(1:3, 5:6, "+") # But we can change it.

	cards <- outer(c("C", "D", "H", "S"), c(2:10, c("J","Q","K","A")), paste, sep = "_") 
	cards   # a 4 x 13 character matrix

	sample(cards, size = 5)    # Sampling without replacement
```

**Coin Tossing**

Question:  Suppose we toss a fair coin 5 times. What is the probability of getting all heads?

Answer: There are $2^5=32$ equally likely outcomes and all heads occurs only one way, so the answer is $1/2^5 = 1/32 = 0.03125$.

Question: What is the probablity of all heads in 50 tosses?

Answer: $2^{15}\approx 1.13\times 10^{15}$, so the probability is $1/2^5\approx 8.88\times 10^{-16}$.

Comment: These questions (and harder ones than this) can be answered using the binomial distribution, e.g., `dbinom(5, 5, 1/2)` yields 0.03125.

# What is probability anyway?
An _event_ is a subset of the possible outcomes of an _experiment_.

* E.g., there are many different sequences of 100 coin tosses that contain at
least one run of 5 or more heads in a row.
* The collection of all such sequences is the _event_ that at least one run of 5 or more heads in a row occurs.
* This event is a subset of the _sample space_ of $2^{100}\approx 1.27\times 10^{30}$ possible outcomes of the experiment.

In frequentist terms, the probability of an event is the proportion of times that it occurs in an infinitely long sequence of independent repetitions of the experiment.

<!-- Probability can also be defined as a quantification of one’s subjective belief that the event in question will occur in a single performance of the experiment. -->

**How Are Probabilities Calculated?**

* Sometimes it is possible to calculate the probability of an event by enumerating the individual outcomes contained in the event, assigning a probability to each of these outcomes, and then adding up those probabilities.
  - Counting techniques and various rules for calculating probabilities (the _calculus_ of probability) can be very useful here.
  - However, in some cases this may be difficult, if not impossible to do.
* An alternative is to perform repeated, independent **simulations** of the experiment using pseudo-random numbers, and then to **estimate** the probability of the event by the proportion of replications in which the event occurrs.

***

**Example: An Occupancy Problem**  

Draw a random sample of 9 letters with replacement from the first five letters of the English alphabet:
```{r}
	sample(LETTERS[1:5], 9, replace = TRUE)
```

If you do this repeatedly, you will see that while all five letters _usually_ appear somewhere in the sample, it is not unusual to find at least one letter missing.

This leads to an obvious question:
What is the probability that the sample contains at least one of each of the five letters, i.e., at least one "A", at least one "B", at least one "C", at least one "D", and at least one "E"?


It would probably take you a while to figure out how to calculate the answer to this question, which is a special case of [the occupancy problem](https://probabilityandstats.wordpress.com/2010/03/27/the-occupancy-problem/).

An alternative approach is to _estimate_ the probability by performing many independent replications of the experiment and counting how often the event in question occurs.
    
**Did All the Letters Show Up in My Sample?**

We first need a way to determine when the event has occured in a single replication of the experiment. We use the `%in%` operator and the `all()` function.
           

```{r}
	x <- sample(LETTERS[1:5], 9, replace = TRUE) 
	x

	LETTERS[1:5] %in% x

	all(LETTERS[1:5] %in% x)
```

# The `for` Loop

The `for` loop iterates through the elements of a sequence, repeating the code in
the body of the loop once for each element in the sequence.

```{r}
	for (dog in c("Lassie", "Toto", "Rin Tin Tin")) { 
		writeLines(paste(dog, "is a famous movie dog."))
		}
```

* Note: Because R is vectorized, we usually don’t need loops.

```{r}
	writeLines(paste(c("Lassie", "Toto", "Rin Tin Tin"), "is a famous movie dog."))
```

***

# Our First Simulation

```{r}
	n_successes <- 0L # Initialize counter 
	for (i in 1:100) {
		x <- sample(LETTERS[1:5], 9, replace = TRUE) 
		is_success <- all(LETTERS[1:5] %in% x) 
		n_successes <- n_successes + is_success
		}
	n_successes / 100
```

If we repeat this simulation, we get different answers (try it). So how accurate is our result?

```{r}
	2 * sqrt(0.5*0.5/100) # An upper bound on the "margin of error" of our estimator.
```

***

Generalizing:

* What if we wanted to do more than 100 replications?
* What if we wanted to do this for 6 letters instead of 5?
* What if we wanted to draw a sample of 10 letters instead of 9?

```{r}
	n_reps <- 1000
	n_ltrs <- 6
	n_sample <- 10
	n_successes <- 0L   # Initialize counter 
	for (i in 1:n_reps) {
		x <- sample(LETTERS[1:n_ltrs], n_sample, replace = TRUE) 
		is_success <- all(LETTERS[1:n_ltrs] %in% x)
		n_successes <- n_successes + is_success
		}
	n_successes / n_reps
```

***

Make It a Function

```{r}
	my_simulator <- function(n_reps, n_ltrs, n_sample) { 
		n_successes <- 0L # Initialize counter
		for (i in 1:n_reps) {
		x <- sample(LETTERS[1:n_ltrs], n_sample, replace = TRUE) 
		is_success <- all(LETTERS[1:n_ltrs] %in% x)
		n_successes <- n_successes + is_success
		}
		n_successes / n_reps 
		}

	my_simulator(10000, 5, 9)

	2 * sqrt(0.5*0.5/10000)  # An upper bound on the "margin of error" of our estimator.
```

***

What is the True Probability?

You can find the formula [here](https://probabilityandstats.wordpress.com/2010/04/04/a-formula-for-the-occupancy-problem/). Note, however, that I have reversed the roles of n and k (statisticians use n for sample size).

```{r}
	pr_all_occupied <- function(n, k) {
		sum((-1)^(0:k) * choose(k, 0:k) * (1 - (0:k)/k)^n)
		}
```

For `n_ltrs = 5` and `n_sample = 9` the actual probability is:
```{r}
	pr_all_occupied(9, 5)
```

***

Here is a more challenging question about coin tossing.

**Question: Suppose we toss a fair coin 100 times. What is the probability of seeing at least one run of 6 or more consective heads or tails?**

Answer: We could attempt to derive a formula to calculate the answer exactly, or we could _estimate_ it by simulation.

To approximate the probability of an event by _simulation_, we:

1. Perform a large number of independent replications of the experiment. 
2. Calculate the proportion of replications in which the event occurrs.

Run Length Encoding a Sequence of Coin Tosses
```{r}
	x <- sample(c("H","T"), 100, replace = TRUE) 
	noquote(x)
	xr <- rle(x) 
	class(xr)
	is.list(xr)
	names(xr)
	xr$lengths
	xr$values
```

Note: `inverse.rle()` will reconstruct the original sequence from the _run_ length encoding produced by `rle()`.

***


What is the longest run?
```{r}
	max(xr$lengths)
```

The longest run of heads?
```{r}
	max(xr$lengths[xr$values == "H"])
```

Of tails?
```{r}
	max(xr$lengths[xr$values == "T"])
```

Aside: We can also convert to a data frame and use `group_by()` and `summarize()`. 

<!-- There is no `as_tibble()` method for class rle, but this works: -->
<!-- ```{r} -->
<!-- library(`tidyverse`) -->
<!-- xr_df <- as_tibble(unclass(xr)) -->
<!-- xr_df |> summarize(max_run = max(lengths)) # Length of longest run overall. -->

<!-- xr_df |> summarize(max_run = max(lengths), .by = values) # Longest runs of H and T. -->
<!-- ``` -->

***


Now we simulate it.

In what proportion of the 10,000 experiments did we observe at least one run of length 6 or more?
```{r}
	nsim <- 10000; ntosses <- 100
	maxrl <- vector("integer", length = nsim) # Pre-allocate a vector to hold results. 
	for (i in 1:nsim) {
		x <- sample(c("H","T"), ntosses, replace = TRUE)   # Toss the coins
		xr <- rle(x)     # Get run-length encoding.
		maxrl[i] <- max(xr$lengths)    # Save result for ith replicate.
		}
```



```{r}
	mean(maxrl >= 6)
```

Thus we estimate the probability of observing at least one run of length 6 or more in 100 tosses of a fair coin to be `{r} mean(maxrl >= 6)`, or `{r} mean(maxrl >= 6)*100`%.

Contrast `{r} mean(maxrl >= 6)` with the probability of observing either 6 H’s or 6 T’s in **6 tosses**.

With only 6 tosses:

* There are possible sequences of $2^6=64$ tosses, all equally likely.
* Each sequence has probability 1/64.
* Exactly two of these sequences result in either all heads or all tails: HHHHHH and TTTTTT.

Thus the probability of getting either all heads or all tails in 6 tosses is
$$2\times\frac{1}{64}=\frac{1}{32}=0.03125$$

Of course the difference is that in 100 tosses we have many more opportunities to observe a run of length 6.

***

We could use `replicate()` to replace the for loop.
```{r}
	rmax_run_len <- function(n) {
		x <- sample(c("H","T"), n, replace = TRUE) 
		xr <- rle(x)
		max(xr$lengths)
		}

	nsim <- 10000; ntosses <- 100
	maxrl <- replicate(nsim, rmax_run_len(ntosses)) 
	mean(maxrl >= 6)
```

**The (Estimated) Distribution of the Maximum Run Length**

*  For the distribution of a discrete random variable, I usually prefer a "high
density plot" (`type = "h"`) over a histogram.
* Plot margins in base-R tend to be too generous, which squeezes the plot. Here we change the margins for this plot and restore the previous values afterwards.
```{r}
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "The results of 10,000 simulated replications of 100 tosses of a fair coin."
#| fig-alt: "Plot showing the results of 10,000 simulated replications of 100 tosses of a fair coin."

	opar <- par(mar = c(5.1, 4.1, 2.1, 1.1)) 
	plot(table(maxrl)/nsim, type = "h", lwd = 5,
		xlab = "Maximum Run Length",
		ylab = "Relative Frequency",
		main = "Estimated distribution of maximum run length.")
	par(opar)
```
