---
title: "Pipes"
subtitle: "Lecture 16"
format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
lang: en
geometry:
  - margin = 1in
  - paperwidth = 11in
  - paperheight = 6.875in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{sectsty}
  - \allsectionsfont{\centering}
  - \pagenumbering{gobble}
execute: 
  error: false
---

# The Pipe Operator(s)
* Pipes allow us to string together function calls without  
  – nesting, or  
  – saving intermediate results. (Examples to follow).
* Pipes can improve readability by emphasizing a sequence of actions rather than the object on which the actions are being performed.
* The idea originated (?) with the UNIX operating system, which has had pipes since its development in the 1970s.
  - In UNIX/Linux command shells, the character "`|`" is the pipe operator.
* [\textcolor{blue}{Since the release of R 4.1}](https://developer.r-project.org/blosxom.cgi/R-4-1-branch/NEWS/2021/05/17#n2021-05-17), base R provides an official/native pipe syntax, "`|>`".  
  - You can look up the R help `?pipeOp` to read more about it.

***

* Importantly, whatever is on the LHS of the pipe is used as the **first argument** of the function on the RHS of the pipe. Thus:  
  – `x |> f()` is interpreted as `f(x)`  
  – `x |> f(y)` is interpreted as `f(x, y)`  
  – `x |> f(y, z)` is interpreted as `f(x, y, z)`  
  and so forth.
* You can pass the lhs into an argument other than the first by using the "\_" placeholder with a _named_ argument. Thus,     
  – `x |> f(y, z = _)` is interpreted as `f(y, z = x)`.  
  You can also use an _anonymous function_ to work around the first argument rule for pipes (more on this later).

***

* We will see that many tidyverse functions
  - take a data frame (tibble) as their first argument, and
  - return a data frame (tibble) as their output.  
This makes them especially pipe-friendly.
* The tidyverse also includes the magrittr pipe "`%>%`", **but** the tidyverse developers now encourage users to employ the base R pipe `|>`.
  - There is a [\textcolor{blue}{new edition of r4ds}](https://r4ds.hadley.nz/) on the way and [\textcolor{blue}{it uses |>}](https://r4ds.hadley.nz/workflow-style).
  - We will use `|>` in this class.
  - But you will see a lot of `%>%` in the wild (and in the 1st edition of r4ds).
  - `|>` and `%>%` are _usually_ interchangeable, but they do not always work identically.

# Which do you prefer?

**Saving intermediate results**
```{r}
evens <- seq(from = 2, to = 100, by = 2)
log_evens <- log(evens)
mean_log_evens <- mean(log_evens)
gmean_evens <- exp(mean_log_evens)
gmean_evens
```

* Tends to clutter up the workspace.
```{r}
ls()
```

***

**Nested function calls**
```{r}
gmean_evens <- exp(mean(log(seq(from = 2, to = 100, by = 2))))
gmean_evens
```

* Have to read the function calls from inside out.

***

**Or pipes**

* Follow the sequence of actions.
```{r}
gmean_evens <-
  seq(from = 2, to = 100, by = 2) |>
    log() |>
    mean() |>
    exp()
gmean_evens
```

* [\textcolor{blue}{The tidyverse style}](https://style.tidyverse.org/) guide recommends that a pipe should always have a space before it and should [\textcolor{blue}{usually be followed by a new line}](https://style.tidyverse.org/pipes.html), as demonstrated above.

***

* Of course if we needed to compute geometric means repeatedly, we might want to write a simple function.

```{r}
geom_mean <- function(x) exp(mean(log(x)))
gmean_evens <- geom_mean(seq(from = 2, to = 100, by = 2))
gmean_evens
```

# Which do you prefer?

**Saving intermediate results**
```{r}
#| include: false

mfr <- rep(c("Fender", "Gibson", "Martin"), times = c(2,4,1))
model <- c("Stratocaster", "Telecaster", "Firebird V", "Les Paul Standard", "ES175",
           "J-100 Xtra", "M-21")
serial <- c("EJ15230", "60671", "616619", "00235628", "03531746", "90783019", "1452869")
type <- rep(c("electric", "acoustic"), times = c(5, 2))
body <- rep(c("solid", "hollow", "flat top"), times = c(4, 1, 2))
w_case_kg <- c(8.96, 8.60, 9.88, 8.40, 8.10, 6.78, 5.62)
case_kg <- c(5.70, 4.98, 5.90, 3.46, 4.92, 4.76, 3.72)
guitars <- data.frame(mfr, model, serial, type, body, w_case_kg, case_kg)
```
```{r}
guitars2 <- transform(guitars, kg = w_case_kg - case_kg)
guitars2 <- transform(guitars2, lb = 2.2 * kg)
guitars2 <- subset(guitars2, select = -c(w_case_kg, case_kg, kg))
guitars2
```

***         

**Nested function calls**

```{r}
guitars2 <- subset(transform(transform(guitars, kg = w_case_kg - case_kg),
                             lb = 2.2 * kg),
                   select = -c(w_case_kg, case_kg, kg))
guitars2
```
***

**Or pipes?**
```{r}
#| include: false

library(dplyr)
```

```{r}
guitars2 <-
  guitars |>
    transform(kg = w_case_kg - case_kg) |>
    transform(lb = 2.2 * kg) |>
    subset(select = -c(w_case_kg, case_kg, kg))
guitars2
```


# Send LHS to Not-the-First Argument of RHS

How can we do these two calls to `log()` with pipes?
```{r}
x <- 2^(1:4) 
x
log(x, 4)        # Same as log(x, base = 4)
log(4, x)        # Same as log(4, base = x)
```

***

* The first one is easy because we are using `x` as the first argument.
```{r}
x |> log(4)    # Same as  x |> log(base = 4)
```
* For the second one, we can use the “_” placeholder with a named argument.
```{r}
x |> log(4, base = _)
```
* Or use an anonymous function (which must be enclosed in parentheses and followed by a pair of parentheses).
```{r}
# Same as x |> (\(b) log(4, b))()
x |> (function(b) log(4, b))()    
```
