---
title: "Evaluating Simulations"
subtitle: "STA3100 Lecture 35"
format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
lang: en
geometry:
  - margin = 1in
  - paperwidth = 11in
  - paperheight = 6.875in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{sectsty}
  - \allsectionsfont{\centering}
  - \pagenumbering{gobble}
execute: 
  error: true
---

# Evaluating Simulation Results

**Tossing a Possibly Biased Coin**

Here is a simple function to run  

* `nsim` simulations of
* `ntoss` independent tosses of
* a coin having probablity `probH` of heads (by default `probH = 1/2`)

The function returns an `ntoss x nsim` matrix:

* rows correspond to tosses
* columns correspond to simulation run.

Each toss is coded as

* `1` if the toss results in a head
* `0` if the toss results in a tail.

```{r}
simtosses <- function(nsim, ntoss, probH = 1/2) { 
  matrix(sample(0:1, ntoss * nsim, replace = TRUE,
                prob = c(1 - probH, probH)), 
         ncol = nsim)
}
ntoss <- 100
nsim <- 10000
sims <- simtosses(nsim, ntoss)
```

Recall this function:

```{r}
fmaxrl <- function(x, outcome = NULL) { 
  xr <- rle(x)
  if (!is.null(outcome)) {
    is_outcome <- (xr$values == outcome) 
    xr$lengths <- xr$lengths[is_outcome]
  }
  if (length(xr$lengths) == 0L) 0 else max(xr$lengths) 
}
```

An estimate of the probability of at least one run of length 6 or more in 100 tosses of a fair coin.

```{r}
phat <- mean(apply(sims, 2, fmaxrl) >= 6) 
phat
```

# How Accurate Are Simulation-Based Estimators?

Let $p$ represent the probability of at least one “long” run (length 6 or more) in 100 tosses of a fair coin?

Q: How accurate is our simulation-based estimator of $p$?

* We have run 10,000 independent simulations.
* Our estimator $\hat{p}$ of $p$ is the proportion of simulation runs in which we observe at least that long run.
* $\hat{p}$ is a _random variable_: if we rerun the simulations, we get a different answer.
* The total number of simulations in which we observe at least one long run is a _binomial random variable_ on 10,000 trials with success probability $p$ (unknown).
* So $\hat{p}$ is a binomial proportion.

***

**Variance and Standard Deviation of $\hat{p}$**  

Recall that the variance of a binomial proportion is given by
$$
\text{var}(\hat{p}) = \frac{p(1-p)}{n}
$$
and its standard deviation the square-root of its variance
$$
\text{sd}(\hat{p}) = \frac{\sqrt{p(1-p)}}{\sqrt{n}}
$$
The largest possible value of $p(1-p)$ is $(1/2)^2=1/4$, so
$$
\text{var}(\hat{p}) \leq\frac{1}{4n} \quad\text{and}\quad\text{sd}(\hat{p}) \leq \frac{1}{2\sqrt{n}}
$$
**Standard Errors**

The _standard error_ of an estimator refers either to the standard deviation of its _sampling distribution_, or more often to an estimate of the same.  

Thus for $\hat{p}$
$$
\text{se}(\hat{p}) = \frac{\sqrt{p(1-p)}}{\sqrt{n}} \quad\text{or}\quad\text{se}(\hat{p}) = \frac{\sqrt{\hat{p}(1-\hat{p})}}{\sqrt{n}}
$$
Some bad news from these formulas:

* To half the standard error, we must increase $n$ by a factor of 4.
* To move the decimal point one place to the left, we must increase $n$ by a factor of 100.
* This is typical of most estimators.

***

**So How Accurate is $\hat{p}$?**

In our earlier simulation of 10,000 replications of the coin-tossing experiment

* We could be sure in advance that
$$
\text{se}(\hat{p}) \leq \frac{1}{2\sqrt{10,000}}=0.005
$$
* We observed $\hat{p}=$ `r phat`, so
$$
\text{se}(\hat{p}) = \frac{\sqrt{`r# phat`(1- `r# phat`)}}{\sqrt{10,000}}=0.004
$$
* When the samples size is large, a binomial proportion is approximately normally distributed.
* Our sample size (10,000) is certainly large, so **prior to running the simulation** we knew that
$$
P(|\hat{p}-p|\leq 1.96 * \text{se}(\hat{p})) \approx 0.95
$$
* As a consequence, we say that we are 95% confident that $p$ is within
$$
1.96 * \text{se}(\hat{p}) = 1.96 * 0.004 = 0.0078
$$
of the _observed value_ `r phat` of $\hat{p}$.
* In other words, we are 95% confident that $p$ is between these values:
$$
\begin{aligned}
`r# phat` - 0.0078 &= `r# phat - 0.0078`\\
`r# phat` + 0.0078 &= `r# phat + 0.0078`
\end{aligned}
$$

***

`binom.test()`
```{r}
x <- sum(apply(sims, 2, fmaxrl) >= 6)
binom.test(x, nsim)          # Args are number of successes and number of trials.
```
* Does _exact_ test (no normal approx). Default null hypothesis is $p=1/2$.
* CI does not agree exactly with ours, but very close with $n$ = 10,000.

***

`prop.test()`
```{r}
x <- sum(apply(sims, 2, fmaxrl) >= 6) 
prop.test(x, nsim)
prop.test(x, nsim)$conf.int
```

# Estimating the Mean Maximum Run Length

Let $\mu$ represent the mean (or “expected value”) of the maximum run length.  
The variance and standard deviation of $\hat{\mu}$ (a.k.a. $\bar{X}$) are
$$
\text{var}(\hat{\mu}) =\frac{\sigma^2}{n} \quad\text{and}\quad\text{sd}(\hat{\mu}) = \frac{\sigma}{\sqrt{n}}
$$
where $\sigma^2$ is the “population” variance ($\sigma$ is the population standard deviation).   
So the standard error of $\hat{\mu}$ is its estimated standard deviation
$$
\text{se}(\hat{\mu}) = \frac{\hat{\sigma}}{\sqrt{n}}
$$
```{r}
x <- apply(sims, 2, fmaxrl) 
muhat <- mean(x)
muhat
sigmahat <- sd(x) 
sigmahat
sigmahat / sqrt(nsim)
muhat + qnorm(c(0.025, 0.975)) * sigmahat / sqrt(nsim)
t.test(x)
t.test(x)$conf.int
```

# What Happens With a Biased Coin?

```{r}
nsim <- 10000
ntoss <- 100
pH <- c(0.5, 0.6, 0.7, 0.8, 0.9)
res <- matrix(nrow = nsim, ncol = length(pH)) 
colnames(res) <- pH
for (i in seq_along(pH)) {
  res[, i] <- apply(simtosses(nsim, ntoss, probH = pH[i]), 2, fmaxrl) 
}
muhat <- apply(res, 2, mean) 
sigmahat <- apply(res, 2, sd) 
se.muhat <- sigmahat / sqrt(nsim) 
round(rbind(muhat, se.muhat), 3)
```

***

**What Should We Expect to Observe?**

```{r}
apply(res, 2, quantile, probs = c(0, 0.05, 0.25, 0.5, 0.75, 0.95, 1))
```


```{r}
#| fig-cap: "Maximum run lengths in 10,000 simulations of 100 tosses of a biased coin."

opar <- par(mar = c(5.1, 4.1, 1.1, 1.1)) 
boxplot(res, horizontal = TRUE,
        xlab = "Maximum Run Length", ylab = "Probability of Heads") 
par(opar)
```

***

**100 Heads in a Row?**

With the probability of heads set to 0.9, the probability of getting 100 heads in 100 (independent) tosses of the (biased) coin is pretty small.
```{r}
0.9^100
```

However, if we repeat the experiment 10,000 times, the probability that we get 100 heads at least once is not so small.

```{r}
1 - (1 - 0.9^100)^10000
```
