---
title: "Logical Operators"
subtitle: "STA3100 Lecture 5"
format:
  html:
    embed-resources: true
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    geometry:
      - margin = 1in
      - paperwidth = 11in
      - paperheight = 6.875in
    header-includes:
      - \usepackage{booktabs}
      - \usepackage{sectsty}
      - \allsectionsfont{\centering}
      - \pagenumbering{gobble}
lang: en
execute:
  error: true
---

## Relational operators: <, >, <=, >=, ==, and !=

```{r}
#| label: xy

	set.seed(32139)
	x <- runif(5) ; y <- runif(5)
	x
	y
```

```{r}
#| label: vw

	v <- x < y
	w <- x >= 1/2
	v
	w
```

\newpage

## Logical Operators
### Negation
```{r}
#| label: negation

	v
	!v
```  

\newpage

### And, Or, and All That
```{r}
#| label: aoa

	v
	w
	v | w         # "or"
	v & w         # "and"
	xor(v, w)     # "exclusive or"
	(v | w) & !(v & w)   # Do you see why this is the same as xor(v, w)?
``` 

\newpage

## Summarizing a Logical Vector
* We have seen that, when applied to a logical vector `x`:  
  + `sum(x)` gives the number TRUE and `sum(!x)` gives the number FALSE.
  + `mean(x)` gives proportion TRUE and `mean(!x)` gives proportion FALSE.  
* `table(x)` gives the counts of both TRUE and FALSE entries:  
  + `summary(x)` displays a similar table when `x` is logical.  
```{r}
#| label: table

	x <- (rnorm(10) > 0)
	x
	table(x)
```

\newpage

* Two very important logical functions are `all()` and `any()`:
  + `all(x)` : TRUE if and only if all the entries of `x` are TRUE.
  + `any(x)` : TRUE if any of the entries of `x` are TRUE.
```{r}
#| label: all

	x <- c(TRUE, TRUE)  ;  y <- c(TRUE, FALSE)  ;  z <- c(FALSE, FALSE)
	all(x)
	all(y)
	any(y)
	any(z)
```

### DeMorgan’s Laws in R Syntax  
A little thought should convince you that for any logical vector `x`:  

* `!any(x)` and `all(!x)` are equivalent, and
* `!all(x)` and `any(!x)` are equivalent. 

Similarly, for any logical vectors `x` and `y`:  

* `!(x | y)` and `(!x) & (!y)` are equivalent, and
* `!(x & y)` and `(!x) | (!y)` are equivalent.
