---
title: "Logical Operators"
subtitle: "Lecture 5"
format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
lang: en
geometry:
  - margin = 1in
  - paperwidth = 11in
  - paperheight = 6.875in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{sectsty}
  - \allsectionsfont{\centering}
  - \pagenumbering{gobble}
execute: 
  error: true
---

## Relational operators: <, >, <=, >=, ==, and !=

```{r xy}

  set.seed(32139)
  x <- runif(5) ; y <- runif(5)
  x
  y
```

```{r vw}

  v <- x < y
  w <- x >= 1/2
  v
  w
```

\newpage

## Logical Operators
### Negation
```{r negation}
  v
  !v
```  

\newpage

### And, Or, and All That
```{r aoa}
  v
  w
  v | w         # "or"
  v & w         # "and"
  xor(v, w)     # "exclusive or"
  (v | w) & !(v & w)   # Do you see why this is the same as xor(v, w)?
``` 

\newpage

## Summarizing a Logical Vector
* We have seen that, when applied to a logical vector `x`:  
  + `sum(x)` gives the number TRUE and `sum(!x)` gives the number FALSE.
  + `mean(x)` gives proportion TRUE and `mean(!x)` gives proportion FALSE.  
* `table(x)` gives the counts of both TRUE and FALSE entries:  
  + `summary(x)` displays a similar table when `x` is logical.  
```{r table}
  x <- (rnorm(10) > 0)
  x
  table(x)
```

\newpage

* Two very important logical functions are `all()` and `any()`:
  + `all(x)` : TRUE if and only if all the entries of `x` are TRUE.
  + `any(x)` : TRUE if any of the entries of `x` are TRUE.
```{r all}
  x <- c(TRUE, TRUE)  ;  y <- c(TRUE, FALSE)  ;  z <- c(FALSE, FALSE)
  all(x)
  all(y)
  any(y)
  any(z)
```

### DeMorgan’s Laws in R Syntax  
A little thought should convince you that for any logical vector `x`:  

* `!any(x)` and `all(!x)` are equivalent, and
* `!all(x)` and `any(!x)` are equivalent. 

Similarly, for any logical vectors `x` and `y`:  

* `!(x | y)` and `(!x) & (!y)` are equivalent, and
* `!(x & y)` and `(!x) | (!y)` are equivalent.