---
title: "Probability Distributions"
subtitle: "STA3100 Lecture 4"
format:
  html:
    embed-resources: true
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
lang: en
geometry:
  - margin = 1in
  - paperwidth = 11in
  - paperheight = 6.875in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{sectsty}
  - \allsectionsfont{\centering}
  - \pagenumbering{gobble}
execute:
  error: true
---

# An Overview of Probability Distributions in R

R has a number of built-in functions related to common families of probability distributions.\
If `fam` is the (abbreviated) name of the distribution, then the following functions are provided:

-   `dfam()` : the `d`ensity of `fam` (usually how the distribution is visualized)\
    – a probability mass function if `fam` is a discrete distribution
-   `pfam()` : the cumulative distribution function (CDF) of `fam`\
    – used to calculate `p`robabilities
-   `qfam()` : the `q`uantile function of `fam`
-   `rfam()` : generates `r`andom observations from `fam`

Function arguments specify the parameters of the distribution.

------------------------------------------------------------------------

**From `help(Distributions)`:**

-   `For the beta distribution see 'dbeta'.`
-   `For the binomial (including Bernoulli) distribution see 'dbinom'.`
-   `For the Cauchy distribution see 'dcauchy'.`
-   `For the chi-squared distribution see 'dchisq'.`
-   `For the exponential distribution see 'dexp'.`
-   `For the F distribution see 'df'.`
-   `For the gamma distribution see 'dgamma'.`
-   `For the geometric distribution see 'dgeom'. (This is also a specialcase of the negative binomial.)`
-   `For the hypergeometric distribution see 'dhyper'.`
-   `For the log-normal distribution see 'dlnorm'.`

------------------------------------------------------------------------

-   `For the multinomial distribution see 'dmultinom'.`
-   `For the negative binomial distribution see 'dnbinom'.`
-   `For the normal distribution see 'dnorm'.`
-   `For the Poisson distribution see 'dpois'.`
-   `For the Student's t distribution see 'dt'.`
-   `For the uniform distribution see 'dunif'.`
-   `For the Weibull distribution see 'dweibull'.`

We will use several of these in this course.

# Example: The Normal Family

**From `help(dnorm)`:**

```r
  dnorm(x, mean = 0, sd = 1, log = FALSE)
  pnorm(q, mean = 0, sd = 1, lower.tail = TRUE, log.p = FALSE)
  qnorm(p, mean = 0, sd = 1, lower.tail = TRUE, log.p = FALSE)
  rnorm(n, mean = 0, sd = 1)
```

-   Default is a *standard normal* distribution: `mean = 0, sd = 1`.
-   Can change this by specifying other values for `mean` and `sd` arguments.

# Plotting a Normal Density

```{r}
#| fig-align: center
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "Normal density with mean 100 and standard deviation 15."
#| fig-alt: "Plot of Normal density with mean 100 and standard deviation 15."
  
  # Saving pop mean and std dev for reuse
  mu <- 100 ; sigma <- 15
  
  # Changing margins of plot in graphical params
  opar <- par(mar = c(5.1, 4.1, 0.1, 0.1))
  
  curve(dnorm(x, mean = mu, sd = sigma),
        xlim = qnorm(c(0.001, 0.999), mean = mu, sd = sigma),
        ylab = "Density")
  
  # Resetting to previous graphical parameters
  par(opar)
```

\newpage

# Normal Quantiles

```{r}

  # Given an upper tail probability, find the z-value.
  qnorm(0.025, lower.tail = FALSE) 

  # First and third quartiles
  qnorm(c(0.25, 0.75), mean = mu, sd = sigma)

  # First and third quartiles from z-values
  mu + sigma * qnorm(c(0.25, 0.75))    
```

\newpage

# Normal Probabilities

```{r}

  # Lower tail probability (area)
  pnorm(125, mean = mu, sd = sigma)   
  # Same after converting to z-score
  pnorm((125 - mu)/sigma)             
  # Upper tail probability
  pnorm(125, mean = mu, sd = sigma, lower.tail = FALSE)    
  # Total probability is one, so ...
  1 - pnorm(125, mean = mu, sd = sigma) 
```

\newpage

# Simulated Sampling from a Normal Distribution

```{r}
  
  # Saving sample size for reuse.
  n <- 30
  # Generate a vector of n = 30 observations.
  x <- rnorm(n, mean = mu, sd = sigma)
  # Sample mean
  mean(x)
  # Sample standard deviation
  sd(x)
  # Sample quantiles
  quantile(x, probs = c(0.25, 0.75))
```

\newpage

### Histogram

```{r}
#| fig-align: center
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "Frequency histogram of 30 observations drawn from a normal distribution with mean 100 and standard deviation 15."

  hist(x)                 
```

\newpage

## t-test and Confidence Interval

-   Setting the argument `mu = 100` in `t.test()` causes it to test the null hypothesis $H_0: \mu = 100$ against the two-sided alternative $H_1: \mu \neq 100$.

```{r}
  t.test(x, mu = 100)
```

# Empirical Distribution vs Population Distribution

### Histogram Density Estimator (`freq = FALSE`)

```{r}
#| fig-align: center
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "Histogram of 30 observations drawn from a normal distribution with mean 100 and standard deviation 15 with the true density superimposed."

  opar <- par(mar = c(5.1, 4.1, 1.1, 0.1))
  hist(x, freq = FALSE, main = "")
  curve(dnorm(x, mean = mu, sd = sigma), ylab = "Density",  add = TRUE)
  par(opar)                
```

-   `freq = FALSE` in `hist()` scales heights of bars to make total area equal to 1.
-   `main = ""` makes the main plot title empty. <!-- * Notice that the y-axis on this plot is set up by `hist()` and isn’t high enough to accomodate the graph of the true density which we have added --> to the plot.

\newpage

In this version:

-   we set the y-axis limits to accomodate the heights of both the density curve and the histogram bars.
-   we set the x-axis limits to accomodate all the bars of the histogram and almost all of the density.

```{r}
#| fig-align: center
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "Histogram of 30 observations drawn from a normal distribution with mean 100 and standard deviation 15 with the true density superimposed."

  opar <- par(mar = c(5.1, 4.1, 1.1, 0.1))
  
  dx <- hist(x, plot = FALSE)
  ylim <- c(0, max(c(dnorm(x, mean = mu, sd = sigma), dx$density)))
  xlim <- range(c(qnorm(c(0.001, 0.999), mean = mu, sd = sigma), dx$breaks))
  plot(dx, freq = FALSE, xlim = xlim, ylim = ylim, main = NULL, col = 'grey80')
  curve(dnorm(x, mean = mu, sd = sigma), ylab = "Density",  add = TRUE)
  par(opar)              
```

\newpage

# Kernel Density Estimator

-   The `density()` function computes a smooth estimator of the density.
-   For comparison, we also show the true population density and a fitted normal density, which uses the sample mean and sample standard deviation.

```{r}
#| fig-align: center
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "Kernel density estimator from a sample of 30 normal random variables with mean 100 and standard deviation 15. The true (dashed) and fitted (dotted) normal densities are superimposed for comparison."

  opar <- par(mar = c(5.1, 4.1, 0.1, 0.1))
  dx <- density(x)
  muhat <- mean(x); sigmahat <- sd(x)
  ylim <- c(0, max(c(dnorm(x, mean = mu, sd = sigma),
                     dnorm(x, mean = muhat, sd = sigmahat),
                     dx$y)))
  plot(dx, ylim = ylim, main = "")
  curve(dnorm(x, mean = mu, sd = sigma), lty = 2, add = TRUE)
  curve(dnorm(x, mean = muhat, sd = sigmahat), lty = 3, add = TRUE)
  
  # Shows the data as a rug on the x-axis
  rug(x)     
  
  par(opar)             
```

\newpage

# Q-Q Plots

Plot the quantiles of the empirical distribution (i.e., the order statistics) against the quantiles of the (fitted) theoretical distribution.

```{r}
#| fig-align: center
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "Normal QQ plot from a sample of 30 normal random variables with mean 100 and standard deviation 15."

  opar <- par(mar = c(5.1, 4.1, 0.1, 0.1))
  qqnorm(x, main = NULL)
  qqline(x)
  par(opar)             
```

\newpage

# Empirical CDF

```{r}

  px <- ecdf(x)
  px
```

```{r}
#| fig-align: center
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "Empirical CDF from a sample of 30 normal random variables with mean 100 and standard deviation 15. True CDR superimposed."

  opar <- par(mar = c(5.1, 4.1, 0.1, 0.1))
  plot(px, main = NULL, verticals = TRUE, cex = 0.75,
       col.points = "blue", col.hor = "red", col.vert = "bisque")
  curve(pnorm(x, mean = mu, sd = sigma), lty = 2, add = TRUE)
  par(opar)          
```
