---
title: "`readr`: Importing and Exporting Data"
subtitle: "Lecture 21"
format:
  html:
    embed-resources: true
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    geometry:
      - margin = 1in
      - paperwidth = 11in
      - paperheight = 6.875in
    header-includes:
      - \usepackage{booktabs}
      - \usepackage{sectsty}
      - \allsectionsfont{\centering}
      - \pagenumbering{gobble}
lang: en
execute:
  error: true
---

# Some `readr` Functions and Their Base-R Analogs
| File Type                    | `readr` function(s)  |Base-R function(s)       |
|:-----------------------------|:---------------------|:------------------------|  
|Comma separated (CSV)         |`read_csv()`          |`read.csv()`             |
|Tab separated                 |`read_tsv()`          |`read.delim()`           |
|General delimited             |`read_delim()`        |`read.delim()`           |
|Fixed width                   |`read_fwf()`          |`read.fwf()`             |
|General white-space separated |`read_table()`        |`read.table()`           |
|Web log files                 |`read_log()`          |                         |

* csv = comma-separated values
* tsv = tab-separated values

# Guitars Again?

```{r}
#| message: false

	cat(readLines('guitars.csv'), sep = '\n')
	library(tidyverse)      ## Or library(readr) for the readr package only
```

***


```{r}
	guitars_df <- read.csv("guitars.csv") 
	summary(guitars_df)
```

***

```{r}
	guitars_tbl <- read_csv("guitars.csv")
	summary(guitars_tbl)
```

***

```{r}
	glimpse(guitars_tbl)
```

# Boston Housing Data

Housing data for 506 census tracts of Boston from the 1970 census. The data frame `Boston` contains the original data by Harrison and Rubinfeld (1979).

* `crim`	per capita crime rate by town
* `zn`	proportion of residential land zoned for lots over 25,000 sq.ft
* `indus`	proportion of non-retail business acres per town
* `chas`	Charles River dummy variable (= 1 if tract bounds river; 0 otherwise)
* `nox`	nitric oxides concentration (parts per 10 million)
* `rm`	average number of rooms per dwelling
* `age`	proportion of owner-occupied units built prior to 1940
* `dis`	weighted distances to five Boston employment centres
* `rad`	index of accessibility to radial highways
* `tax`	full-value property-tax rate per USD 10,000
* `ptratio`	pupil-teacher ratio by town
* `b`	1000(B - 0.63)^2 where B is the proportion of blacks by town
* `lstat`	percentage of lower status of the population
* `medv`	median value of owner-occupied homes in USD 1000's
```{r}
	houses_df <- read.table("Boston.txt", header = TRUE)
	head(houses_df)
```

***

```{r}
	houses_tbl <- read_table("Boston.txt", col_names = TRUE)
	names(houses_tbl) <- gsub('"', "", names(houses_tbl))
	houses_tbl
```

***

```{r}
	glimpse(houses_tbl)
```

# UF Academic Programs

```{r}
	library(readxl)
	uf_progs <- read_excel("ufAcademicPrograms.xlsx")
```

***

```{r}
	uf_progs
```

***

```{r}
	glimpse(uf_progs)
```

***

Spreadsheet data are often “untidy”. These data exhibit a common problem.
```{r}
	uf_progs |>
		select(College, Program)    # Or uf_progs[, c("College", "Program")]
```
* Each “college” only appears once.

***

```{r}
	uf_progs |>
		select(College, Program) |>
			filter(!is.na(College))
```

***

**One way to fill in the blanks in this example**

* Note: you could turn this into a function.

```{r}
	has_college <- which(!is.na(uf_progs$College)) 
	length(has_college)
	has_college
```

***

```{r}
	num_prog_by_college <- diff(c(has_college, 1 + nrow(uf_progs))) 
	num_prog_by_college
	sum(num_prog_by_college) # Sanity check
	nrow(uf_progs)
```

***

```{r}
	## uf_progs$College <- rep(uf_progs$College[has_college], times = num_prog_by_college)

	uf_progs <- uf_progs |>
		mutate(College = rep(College[has_college], times = num_prog_by_college))

	uf_progs |>
		select(College, Program) |>
			slice_head(n = 8)
```

***

```{r}
	summarize(uf_progs, n = n(), .by = College)
```

***

```{r}
	uf_progs |>
		select(College, Program) |>
			slice(29:32)
```

# Exporting and Saving Data

| File Type                    | `readr` function(s)   |Base-R function(s)       |
|:-----------------------------|:----------------------|:------------------------|  
|Comma separated (CSV)         |`write_csv()`          |`write.csv()`            |
|Tab separated                 |`write_tsv()`          |`write.table()`          |
|General delimited             |`write_delim()`        |`write.table()`          |
|R serialization               |`write_rds()`          |`writeRDS()`             |

* You can also use `write_xlsx()` from `writexl` package to write an `xlsx` file.

***

**Comments on These Data Formats**

* CSV, TSV, and Other Character Delimited Formats
  - (+) Simple and portable.
  - (–) Things like factors and dates are written as character strings.
  - (–) R class information and other attributes are lost.
* R Data Serialization (RDS)
  - (+) No data or information lost with RDS. Perfect representation of R object. – (+) Load quickly in R using `read_rds()` or `readRDS()`.
  - (–) Not portable. Will not generally be readable by other software.
  
***

* MS Open Office XML Spreadsheet (xlsx)
  - XML (Extensible Markup Language) file, compressed in zip format.
    + (+) More portable than the older, proprietary xls format (avoid).
    + (–) Still not all that portable.
  - (–) Spreadsheets seem to encourage un-tidy data.

***

* By default `write_csv()` and `write.csv()` represent missing values with “NA”.

```{r}
	write_csv(uf_progs, file = "ufAcademicPrograms2.csv")
	cat(readLines("ufAcademicPrograms2.csv", n = 4), sep = "\n")
```

***

* It is more common to use an empty field for missing values.

```{r}
	write_csv(uf_progs, file = "ufAcademicPrograms3.csv", na = "")
	cat(readLines("ufAcademicPrograms3.csv", n = 4), sep = "\n")
```

# FILLING
# A Filling Function

The empty cells problem seen in the UF academic programs spreadsheet comes up often. It would be useful to have a function to fix this problem. We have seen and written

* simple, one-line functions
* with and without optional arguments with default values.

For more complicated functions:

* the “body” of the function is enclosed in curly braces `{...}`.
* the last line or the last function call in the body determines the value returned by the function (otherwise use `return()`).

***

Here is a first try at a function to fill a vector in the way that we filled the `College` column of the UF academic programs dataset.

```{r}

	fill_vector <- function(x) {
		has_val <- which(!is.na(x))
		times_to_repeat <- diff(c(has_val, 1 + length(x))) 
		rep(x[has_val], times = times_to_repeat)
		}

	x <- c(1, NA, NA, 2, NA, NA, NA, 3, NA)
	x

	fill_vector(x)             # Seems to work.
```

***

This fills to the right (or down) the vector. What if we want to to fill left (or up)?

* Giving optional argument as a vector so that all the options can be seen, but only using the first value.
* Will use `rev()` (reverse) function. May be slow for _very_ long vectors.
* Note: changing `x` inside the function does not change any object named `x` outside the function.
```{r}
	fill_vector <- function(x, direction = c("right", "down", "left", "up")) { 
		direction <- direction[1]
		if (direction %in% c("left", "up")) x <- rev(x)
		has_val <- which(!is.na(x))
		times_to_repeat <- diff(c(has_val, 1 + length(x))) 
		x <- rep(x[has_val], times = times_to_repeat)
		if (direction %in% c("left", "up")) x <- rev(x)
		x
		}
```

***

```{r}
	x
	fill_vector(x)                   # Still works as expected.
	y <- c(NA, "a", NA, NA, "b", "c")
	y
	fill_vector(y, direction = "left")
```

***

Let’s apply this to the UF academic programs dataset in a mutate step.
```{r}
	library(tidyverse)
	library(readxl)
	uf_progs <- read_excel("ufAcademicPrograms.xlsx")
	uf_progs <- uf_progs |>
		mutate(College = fill_vector(College))
	uf_progs |>
		select(College, Program) |>
			slice_sample(n = 8) |>
				arrange(College, Program)
```

# `tidyr::fill()`

* While working on this, I decided to check that there wasn't already a function called `fill()` or `fill_something()`.
* To my surprise, I found `tidyr::fill()` which I have never seen mentioned anywhere (it is not in r4ds2e nor any of the `tidyr` vignettes). 
* `tidyr::fill()` works differently than our function (see `?fill`).
  - First argument is a data frame (`tibble`).
  - Followed by columns to fill.
  - `.direction` can be `"down"`, `"up"`, `"downup"`, or `"updown"`. Default is `"down"`.
* Let’s try it on the UF academic programs data.

***

```{r}
	uf_progs <- read_excel("ufAcademicPrograms.xlsx")
	uf_progs <- fill(uf_progs, College)
	uf_progs |>
		select(College, Program) |>
			slice_sample(n = 12) |>
				arrange(College, Program)
```

# More Checking of the College Column

In the example of UF's academic programs, we might want to further verify our repair of the "College" column further by checking that the transitions look correct.

* First lets get back to where the class slides left off. Refer to those slides for explanation.

```{r}
#| message: false

	library(tidyverse)
	library(readxl)
	uf_progs <- read_excel("ufAcademicPrograms.xlsx")
	has_college <- which(!is.na(uf_progs$College))
	num_prog_by_college <- diff(c(has_college, 1 + nrow(uf_progs)))
	uf_progs$College <- rep(uf_progs$College[has_college], times = num_prog_by_college)
```

***

* Lets see the first row and check that the college changes when we go from row `has_college[2] - 1` (= 30) to row `has_college[2]` (= 31).

```{r}
	uf_progs[1, c("College", "Program")]
	uf_progs[has_college[2] + (-1:0), c("College", "Program")]
```

***

* Similarly for the transition from row `has_college[3] - 1` (= 42) to row `has_college[3]` (= 43), this time with `dplyr` verbs instead of base-R indexing.
```{r}
	uf_progs |>
		slice(has_college[3] + (-1:0)) |>
			select(College, Program)
```

***

* Finally the last transition from row `last(has_college) - 1` (= 182) to row `last(has_college)` (= 183) and finally the last row of the data frame.
  - `last()` is a `dplyr` function.
```{r}
	uf_progs[last(has_college) + (-1:0), c("College", "Program")]
	uf_progs[nrow(uf_progs), c("College", "Program")]
```

***

**Can We See All These Transitions at Once?**
```{r}
	has_college
```

* I'm going to create an index vector the rows that will show
  - the first row
  - rows 30 and 31 (change from Ag to Arts)
  - rows 42 and 43 (change from Arts to Biz Admin)
  - ...
  - the last row
* To do this I will create a matrix and then take advantage of column-major order to turn the matrix into the desired vector of indicies.

***

```{r}
	idx <- rbind(has_college - 1, has_college)            # Or idx <- outer(-1:0, has_college, `+`) 
	idx
	idx <- as.vector(idx)
	idx
```

***

```{r}
	idx <- idx[-1]
	idx
	idx <- c(idx, nrow(uf_progs))
	idx
```

That’s it!

***

```{r}
	uf_progs |>
		slice(idx) |>
			select(College, Program) |>
				print(n = Inf)
```
