---
title: "Programming"
subtitle: "Lecture 12"
format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
lang: en
geometry:
  - margin = 1in
  - paperwidth = 11in
  - paperheight = 6.875in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{sectsty}
  - \allsectionsfont{\centering}
  - \pagenumbering{gobble}
execute: 
  error: true
---

# Define a function

```{r}
annuityAmt <- function(n, R, i) {
	R * ((1 + i) ^ n - 1)      # Only return one value
}
annuityAmt(4, 1000, 0.0528)
```
Functions with default values are defined as follows.   
```{r}
#| error: true
annuityAmt2 <- function(n, R = 1000, i = 0.0528) {
	R * ((1 + i) ^ n - 1) 
}
annuityAmt2(4)
annuityAmt(4)
```

You can also change default values.
```{r}
#| error: true
annuityAmt2(4, R = 1000, i = 0.0528)
```
***

* With multiple output values, the outputs are returned as a list.

```{r}
annuityAmt3 <- function(n, R = 1000, i = 0.0528) {
	a <- R * ((1 + i) ^ n - 1) 
	b <- a + R
	return(list(total = b, interest = a))
}
```

```{r}
annuity <- annuityAmt3(4, R = 1000, i = 0.0528)
is.list(annuity)
str(annuity)
```

***

# Scope of variables
```{r}
k <- 200
annuityAmt <- function(n, R, i) {
	R * ((1 + i) ^ n - 1) + k    # if not defined in the function, R will use the variable in the global environment
}
annuityAmt(4, 1000, 0.0528)
```
```{r}
k <- 200
annuityAmt <- function(n, R, i) {
	R * ((1 + i) ^ n - 1) + k
	k <- 300     # Change the value of k locally
}
annuityAmt(4, 1000, 0.0528)
k
```
```{r}
k <- 200
annuityAmt <- function(n, R, i) {
	R * ((1 + i) ^ n - 1) + k
	k <<- 300  # Change the value of k globally
}
annuityAmt(4, 1000, 0.0528)
k
```

***

# The `if` statement
```{r}
x <- 3
if (x > 2) {
	y <- 5 * x
} else if (x < -2)  {
	y <- 3 * x
} else {
  y <- x
}
y
```

***

## The `for` loop
Example: Compute the factorial $100!$.
```{r}
n <- 100
result <- 1
for (i in 1:n) {
  result <- result * i
}
result
```

Example: Fibonacci sequence
```{r}
Fibonacci <- numeric(100)
Fibonacci[1] <- 1
Fibonacci[2] <- 1
for (i in 3:100) Fibonacci[i] <- Fibonacci[i - 2] + Fibonacci[i - 1]
Fibonacci
```


# The `while` loop

Example: Fibonacci sequence
```{r}
# Fib1 <- 1
# Fib2 <- 1
# Fibonacci <- c(Fib1, Fib2)
# while (Fib1 + Fib2 < 300) {
# 	newFib <- Fib1 + Fib2
# 	Fibonacci <- c(Fibonacci, newFib)
# 	Fib1 <- Fib2
# 	Fib2 <- newFib
# }
# Fibonacci

Fibonacci <- c(1, 1)
while (Fibonacci[length(Fibonacci)] < 300) {
  Fibonacci <- c(Fibonacci, sum(Fibonacci[length(Fibonacci)-(0:1)]))
}
Fibonacci <- Fibonacci[-length(Fibonacci)]
```
    
***
Another example: Newton’s method for root finding

Newton's method is a popular numerical method to find a root of an algebraic equation:
$$
f(x)= 0.
$$
If $f(x)$ has derivative $f'(x)$, then the following iteration should converge to a root of the above equation if started close enough to the root.
$$
x_0 = \text{initial guess},
$$
$$
x_n = x_{n-1} - \frac{f(x_{n-1})}{f'(x_{n-1})}.
$$
The idea is based on the Taylor approximation
$$
f(x_n)\approx f(x_{n-1})+(x_n-x_{n-1})f'(x_{n-1}).
$$
Newton's method is equivalent to setting $f(x_n)=0$ and solving for $x_n$ in the approximation above.  We hope that the solution $x_n$ should give a close approximation to the root.  Be careful: there are many examples where $x_n$ will fail to converge to a root unless $x_{n-1}$ is already sufficiently close, and some where it will fail regardless of the starting value.   Now suppose $f(x)=x^3+2x^2-7$, find the root of $f(x)$.
```{r}
x0 <- 1
x <- x0
f <- x ^ 3 + 2 * x ^ 2 - 7
tolerance <- 0.000001
while (abs(f) > tolerance) {
	f.prime <- 3 * x ^ 2 + 4 * x
	x <- x - f / f.prime
	f <- x ^ 3 + 2 * x ^ 2 -7
	print(x)
}
x
```  

***

# The `repeat` loop, and the `break` and `next` statements
We use a `repeat` loop if we do not want a fixed number of repetitions of a loop, and we do not want to put the test at the top of the loop as in `while`.
```{r}
x0 <- 1
x <- x0
tolerance <- 10^-6
repeat {
	f <- x ^ 3 + 2 * x ^ 2 - 7
	if (abs(f) < tolerance){
	  break
	}
	f.prime <- 3 * x ^ 2 + 4 * x
	x <- x - f / f.prime
	print(x)
}
x

x0 <- 1
x <- x0
tolerance <- 10^-19
max_iter <- 10^3
for(i in 1:max_iter) {
	f <- x ^ 3 + 2 * x ^ 2 - 7
	if (abs(f) < tolerance){
	  break
	}
	f.prime <- 3 * x ^ 2 + 4 * x
	x <- x - f / f.prime
	print(x)
}
```

*** 

Example on `break` and `next`
```{r}
x <- 0
a <- rep(0, 10)
repeat {
	x <- x + 1
	if (x == 8) break
	a[x] <- x
}
a
```

```{r}
x <- 0
a <- rep(0, 10)
repeat {
	x <- x + 1
	if (x == 8) break
	if (x == 3) next
	a[x] <- x
}
a
```

# Efficient programming
Loops in R can be slow, so it is fast working with vectors.
```{r}
X <- rnorm(100000)
Y <- rnorm(100000)
Z <- c()

system.time({
	for (i in 1:100000)  Z <- c(Z, X[i] + Y[i])
})

system.time({
	for (i in 1:100000)  Z[i] <- X[i] + Y[i]
})

system.time(Z <- X + Y)
```

