---
title: "Assignment 8"
subtitle: "STA3100"
format:
  html:
    embed-resources: true
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    geometry:
      - margin = 1in
lang: en
execute:
  eval: false
  echo: true
---

## Reading Assignment:

* [r4ds2e](https://r4ds.hadley.nz/): 6, 8, 22

**General Instructions:** For these exercises, we will use the charm city circulator bus ridership dataset, `Charm_City_Circulator_Ridership.csv`. After moving the dataset file to the location of your .qmd, or modifying the working directory to the dataset location, use the following code to read in and transform the dataset to be ready for use in plotting:
```{r}
#| eval: false

  library(tidyverse)
  library(lubridate)
  circ <- read_csv("Charm_City_Circulator_Ridership.csv")
  
  # convert dates
  circ <- circ |>
    mutate(date = mdy(date))
  
  # change colnames for reshaping
  colnames(circ) <- colnames(circ) |>
    str_replace("Board", ".Board") |>
    str_replace("Alight", ".Alight") |>
    str_replace("Average", ".Average")
  
  # make into long format by route/type
  long <- circ |>
    pivot_longer(
      c(
        starts_with("orange"),
        starts_with("purple"),
        starts_with("green"),
        starts_with("banner")
      ),
      names_to = "var", values_to = "number",
    )
  
  # separate route/type into route and type
  long <- long |>
    separate(var, into = c("route", "type"), sep = "[.]")
  
  # make into wide format by type
  type_wide <- long |>
    spread(type, value = number)
  
  ## extract just average ridership per day
  avg <- long |>
    filter(type == "Average", !is.na(number)) |>
    select(-type) |>
    rename(Average = number)
```
In these questions, try to use `ggplot2` if possible. You may use any of the tibbles created in the code above. Take some time to acquaint yourselves with them before proceeding. Each of the tibbles reflects the bus ridership, but they are structured differently.

## Problems:

1. Plot average ridership by date using a scatterplot. Do not compute an average, use the `Average` value for each route.
    a) Color the points by route.
    b) Replot (1a). Add black smoothed curves for each route.
    
2. Plot average ridership by date. Color the points by day of the week.

3.  Plot average ridership by date. Color the points by route. Use the pallete below. Colors should follow the route name (with `banner` corresponding to "blue"):
    ```{r}
    #| eval: false
    
      pal <- c("darkorange", "purple", "darkgreen", "blue")
    ```
4. Plot average ridership by date in separate panels by route.

5. Plot average ridership by date in separate panels by day of the week. Color the points by route.

6. Plot average ridership by date. Color the points by route. Make the x-label "Year". Make the y-label "Number of People". Use the black and white theme `theme_bw()`. Change the text size using the argument `text = element_text(size = 20)` in `theme()`.

7. Plot average ridership on the `orange` route vs. date as a solid line. Add dashed lines based on the `Boardings` and `Alightings` for the route. Line colors should be "darkorange". (HINT: `linetype` is an aesthetic for line geoms - see `scale_linetype()` and `scale_linetype_manual()`. Use `Alightings = "dashed"`, `Boardings = "dashed"`, and `Average = "solid"` within the `values` argument). Do the number of boardings and alightings differ from the average value in your assessment?
