---
title: "`tidyr`: Tidy Data"
subtitle: "Lecture 27"
format:
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
lang: en
geometry:
  - margin = 1in
  - paperwidth = 11in
  - paperheight = 6.875in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{sectsty}
  - \allsectionsfont{\centering}
  - \pagenumbering{gobble}
execute: 
  error: true
---

# Principles of Tidy Data

1. Each variable is a column; each column is a variable. 
2. Each observation is a row; each row is an observation. 
3. Each value is a cell; each cell is a single value.

These principles are commonly violated in a variety of ways.

# Border Crossings Data
The file `Annual Table.tsv` was obtained from the U.S. Department of Transportation's [\textcolor{blue}{Bureau of Transportation Statistics}](https://www.bts.gov/).
```{r}
#| message: false

border_tsv <- "Annual Table.csv"
```

The file was downloaded from [\textcolor{blue}{https://explore.dot.gov/views/BorderCrossingData/Annual}](https://explore.dot.gov/views/BorderCrossingData/Annual) by

* clicking on the download icon at the bottom right of the page, 
* choosing “Crosstab”, and then
* choosing a format.

Strangely, choosing format CSV downloads a tab-separated-values (TSV) file named `Annual Table.csv`.

* I only noticed this after trying to read the file with `read_csv()`.

```{r}
library(tidyverse)
border <- read_csv(border_tsv, locale = locale(encoding = "UTF-16")) 
slice_head(border, n = 6)
```

* Those `\t`’s are tab characters.
* More about the UTF-16 encoding in a moment.

```{r}
border <- read_tsv(border_tsv)
```

* Apparently this error is due to an unexpected text encoding / character set.
* I found the encoding of this file by typing `file -I filename` in a Terminal (on a Macbook), but there are many other ways to get this information (google it).
* My system expects UTF-8 by default.

```{r}
locale()
locale(encoding = "UTF-16")
border <- read_tsv(border_tsv, locale = locale(encoding = "UTF-16")) 
border |> print(n = 8)
```

**Data in Column Names**

Looking at these data, we see that:

* The _values_ of the _variable_ year are recorded in column names.
* The numeric values recorded in the year columns are counts (number of crossings).

We can fix this problem with `tidyr::pivot_longer()`.

* Aside: Interestingly, the xlsx version of the dataset also exhibits the _empty- cells-make-reading-easier_ problem, but we know how to fix that with `tidyr::file()`.

# Pivoting Longer


```{r}
border <- border |>
  pivot_longer(cols = -(1:2), names_to = "Year", values_to = "count")
border |> print(n = 8)
```

* As we saw in the original dataset, there are some missing counts:

```{r}
border |> group_by(is.na(count)) |> summarize(n = n())
```

* The missing counts could possibly be taken to be zeros, but it is not clear at this stage whether they are
  - structural (e.g., a port of entry hadn’t opened in the earlier years), or
  - actual zero counts, or
  - data that simply weren’t available/recorded for some other reason,
  - or some combination of these three possibilities.

# Pivoting Wider

It seems natural to consider the _observational units_ to be the port-by-year combinations. In this case:

* the values (levels) in the `Measure` column are _variables_.

As we show next, the levels of `Measure` won’t make great variable names, but we can change them later.

```{r}
border |> distinct(Measure)
border <- border |> pivot_wider(
  names_from = Measure,
  values_from = count 
  )
border |> print(n = 5)
``` 

* In some cases you might need to specify identifying columns.
* In this example, the columns `Port Name` and `Year` identify the observation
units, so we could have typed:
```{r}
#| eval: false

border <- border |>
  pivot_wider(names_from = Measure, values_from = count,
              id_cols = c(`Port Name`, Year))
```

However, removing the columns `Measure` and `count` used in `names_from` and `values_from` leaves only the two columns that specify the observation units, `Port Name` and `Year`, so there was no need to specify the identifying columns in this example.

* The column names are not great. 

```{r}
glimpse(border)
```

* Could use `janitor::clean_names()` to get consistent, syntactic names:

```{r}
border <- border |>
  rename(Port = `Port Name`) |> 
  janitor::clean_names()
names(border)
```

* If desired(??), we could abbreviate some of these too.

```{r}
border <- border |> 
  rename_with(
    ~ str_replace_all(.,
                      c("passengers" = "psngrs",
                        "containers" = "cntnrs"))
    ) 
names(border)
```