---
title: "Vectors"
subtitle: "STA3100 Lecture 2"
format:
  html:
    embed-resources: true
  pdf:
    pdf-engine: lualatex
    pdf-standard: ua-2
    geometry:
      - margin = 1in
      - paperwidth = 11in
      - paperheight = 6.875in
    header-includes:
      - \usepackage{booktabs}
      - \usepackage{sectsty}
      - \allsectionsfont{\centering}
      - \pagenumbering{gobble}
lang: en
execute:
  error: true
---

\newpage

# `c( )` : The Combine (or Concatenate) Function

The most basic way to create and combine vectors in R.
```{r}
  x <- c(1, exp(1), pi)
  x
  
  is.double(x) 
  
  y <- c(-5, -3, -1) 
  y
  
  is.double(y)
```

```{r}
  x
  y
  
  c(x, y)
  c(y, x)
```

\newpage

# Vector Arithmetic
```{r}
  x
  y
  
  2 * x     # Mulitplying a vector by a scalar
  x + 2     # Adding a scalar to a vector
  x + y     # Adding two vectors
```

\newpage

**Recyling!**
```{r}
  z <- c(y, 10, 20, 30)
  x
  z
  
  length(x)
  length(z)
  
  x + z   
  c(x, x) + z
```

\newpage

**Element-Wise Multiplication**
```{r}
  x
  y
  x * y
```

\newpage

**Dot (or Inner) Product**
```{r}
  x %*% y
  sum(x * y)
```

\newpage

# Generating Sequences
```{r}
  1:10
  
  5:(-2)
  
  2 * (1:5)
  
  seq(from = 2, to = 10, by = 2) 
  
  seq(from = 1.5, to = 3.5, by = 0.25)
  
  seq(from = -pi, to = pi, length = 7)
```

\newpage

# Repeating Things
```{r}
  rep(3, 5)
  
  1:3
  
  rep(1:3, times = 2) 
  
  rep(1:3, each = 4)
  
  rep(1:3, length.out = 10) 
```

\newpage

**A Puzzle**  
Can you see that these should be the same?
```{r}
  rep(rep(1:3, times = 4), each = 2)
  rep(rep(1:3, each = 2), times = 4)
```
Because they are the same, the following is unambiguous:
```{r}
  rep(1:3, each = 2, times = 4)
```

\newpage

# INDEXING VECTORS
# Indexing with Numeric Indicies
* _Indexing_ means selecting specific elements.
* Square brackets (`[ ]`) are used to index a vector.
* Positive integer indices specify which elements to keep.
```{r}
  x <- seq(from = 1, length = 6, by = 2)
  x
  
  x[3]           # A single element of x
  
  x[c(2,5)]      # Multiple elements of x (vector of indicies)
  
  x[2:5]
  
  x[1]
  
  x[length(x)]
```

\newpage

* Negative indices specify which elements to omit

```{r}
  x[-3]          # Leave out a single element of x
  x[-c(2,5)]     # Leave out mulitple elements (vector of negative indicies)
  x[-(2:5)]
```

\newpage

* Missing values and other special values

```{r}
  v <- NULL
  v
  v[seq(2, 20, 2)] <- seq(2, 20, 2)
  v
  x <- c(0, 7, 8)
  x / x
  1 / x
  x[0.5]
```

\newpage

* Logical vectors and relational operators
```{r}
  a <- c(TRUE, FALSE, FALSE, TRUE)
  a
  sum(a)
  !a
  a & a
  a | !a
```

\newpage

# Indexing with a Vector of Logical Indicies
* Vector of logical indicies should have same length as vector being indexed.
* Return elements with index TRUE, omit elements with index FALSE.
```{r}
  (1:5)[c(FALSE, TRUE, TRUE, FALSE, TRUE)]
```

* Indexing with a vector of logical indicies is extremely useful.
* For example, it can be used to return the elements of a vector that satisfy a specified condition.

\newpage

**Example**

* Generate a random sample of 12 IQ scores:
```{r}
  iq <- rnorm(12, mean = 100, sd = 15) # Sample 12 observations from a normal distribution 
  iq <- round(iq)                      # Round to the nearest integer
  iq
```

* This comparison returns a logical vector of length 12:
```{r}
  iq > 105
```

* Use it as an index to return only the iq scores greater than 105:
```{r}
  iq[iq > 105]
```

\newpage

# Named Vectors

* The elements of a vector can be named.

```{r}
  ages <- c("bob" = 39, "carol" = 31, "ted" = 31, "alice" = 32) 
  ages
```

* Note that R allows you to omit the quotes on the names when defining a named vector.

```{r}
  ages <- c(bob = 39, carol = 31, ted = 31, alice = 32) 
  ages
```

\newpage

* The `names()` function returns the character vector of names.
```{r}
  names(ages)
```

* You can also use it to assign names to a vector.
```{r}
  ages <- c(39, 31, 31, 32) 
  ages
  names(ages)
  
  names(ages) <- c("bob", "carol", "ted", "alice") 
  ages
  names(ages)
```

\newpage

# Indexing by Name
```{r}
  
  ages
  ages[ted]  # Bad! Does not work
  ages["ted"] 
  ages[c("ted", "alice")] 
  
  who <- c("ted", "alice") 
  ages[who]
```

\newpage

# More Names and Naming
```{r}
  ages <- c(ages, 35)
  ages
  
  names(ages)
  names(ages)[5]
  
  names(ages)[5] <- "horst"
  ages
```

\newpage

```{r}
  ages <- ages[-5]
  ages
  
  actors <- c("Robert Culp", "Natalie Wood", "Elliott Gould", "Dyan Cannon") 
  actors
  
  names(actors) <- names(ages)
  actors
```

\newpage

```{r}
  actors <- sort(actors)
  actors
  names(ages)
  
  actors[names(ages)]
  
  actors <- actors[names(ages)]
  actors
```

\newpage

# Character vectors
```{r}
  colors <- c("red", "yellow", "blue")
  colors
  more.colors <- c(colors, "green")  # more.colors is a variable name, not a list as in MATLAB
  more.colors
  more.colors[2:3]
  mix.colors <- c(colors, 1)
  mix.colors
  substr(colors, 1, 3)
  substr(colors, 1:3, 2:4)
  paste(colors, "flowers")
  paste("several", colors, "s")
  paste("several ", colors, "s", sep = "")
  paste0("several ", colors, "s")
  paste("I like", colors, collapse = ", ")
```

\newpage

# Vectorized Functions
Most numeric functions in R are vectorized: they accept a vector argument and return a vector result.

```{r}
  angle <- seq(from = 0, to = 180, by = 30) 
  angle
  
  round(sin(angle * pi / 180), 3)
```

Names are (usually) preserved.
```{r}
  ages
  Ages    # R is case-sensitive
  log10(ages)
```

\newpage

# Defining Your Own R Function

* A simple R function can be defined on one line.

```{r}
  deg_to_rad <- function(x) x * (pi / 180) 
  angle
  deg_to_rad(angle)
```

* As we have seen, function calls can be nested.
```{r}
  sin(deg_to_rad(angle))
```

\newpage

* Arguments can be given default values which may be overridden when the function is called.
```{r}
  incr <- function(x, by = 1) x + by
  incr(1:5)
  incr(1:5, by = 3)
```
